
### Load data and libraries
```{r Load_data}
##################
# LOAD LIBRARIES #
##################
library(tidyverse)
library(readxl)
library(org.Hs.eg.db)
library(biomaRt)

select <- dplyr::select
# setwd("/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/src")

#########
# PATHS #
#########
input_dir <- "../data/"

meta_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metadata_svamp.xlsx"
mbol_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/UTF-8SMC_P0800_GCMS_Data.xlsx"

```

To determine the groups you need column from the clinical file

"Fungal culture: C. albicans (y/n)", 
"Symptom score (0-5)", 
"Recurring fungal infections > 2/year (y/n)"

RVVCpos:	Positive fungal cultures, symptom score ≥1, >2 episodes of VVC per year
Control:	Negative fungal culture, ≤2 episodes of VVC per year
AS:     	Positive fungal scores, symptom score = 0, ≤2 episodes of VVC per year
RVVCneg:	Negative fungal culture, >2 episodes of VVC per year

## get group assignment

```{r group-assignment}
meta_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metadata_svamp.xlsx"
meta_data <- read_xlsx(meta_path, sheet = "Metadata", skip = 1)

gr_data <- meta_data %>%
  select(
    DS_ID = "ID",
    svamp_ID,
    Treatment,
    `Fungal culture: C. albicans (y/n)`,
    `Symptom score (0-5)`,
    `Recurring fungal infections > 2/year (y/n)`
  ) %>%
  mutate(
    group = case_when(
      `Fungal culture: C. albicans (y/n)` == "1" &
        `Symptom score (0-5)` >= 1 &
        `Recurring fungal infections > 2/year (y/n)` == "1" ~
        "RVVCpos",

      `Fungal culture: C. albicans (y/n)` == "0" &
        `Recurring fungal infections > 2/year (y/n)` == "1" ~
        "RVVCneg",

      `Fungal culture: C. albicans (y/n)` == "1" &
        `Symptom score (0-5)` == 0 ~
        "AS",

      `Fungal culture: C. albicans (y/n)` == "0" &
        `Recurring fungal infections > 2/year (y/n)` == "0" ~
        "Control",

      `Fungal culture: C. albicans (y/n)` == "1" &
        `Recurring fungal infections > 2/year (y/n)` == "0" ~
        "Candidapos",

      TRUE ~ NA_character_
    )
  ) %>%
  select(ID = "svamp_ID", everything())

write_csv(gr_data, "../Tidy_data/Asigned_groups_info_all_samples.csv")
# gr_data <- read_csv("../Tidy_data/Asigned_groups_info_all_samples.csv")
```

## Format transcriptomics data

There is something off with the naming of sample 12 or S69 it has 12:01 listed as baseline and a 1 month follow up which I have not seen any where else
It turns out 12:01 is a baseline sample that was mistakenly named :01 instead of 02 as it should have ben, and the 1 week follow up should be three month follow up
Additionaly the smaple "S75_BL" is called "S75_Bl" in the metadata sheet so I had to also fix this here

This is recitified bellow

```{r fix-mapping-files} 
##################
# LOAD LIBRARIES #
##################
library(tidyverse)
library(biomaRt)
library(org.Hs.eg.db)
library(AnnotationDbi)

#############
# LOAD DATA #
#############
# NB! I had to add a 0 to the front of some of the patient_ID with numbers less than 10 (02, 03, 07, 09)
# to the svamp_ID_list.xlsx from Margaux
smpl_IDs <- read_xlsx(paste0(input_dir, "svamp_ID_list.xlsx"))
smpl_barcodes <- read_xlsx(
  paste0(input_dir, "Registrering Provrör ID CHLORHEX.xlsx"),
  skip = 1
)
meta_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metadata_svamp.xlsx"
meta_trx <- read_xlsx(
  meta_path,
  sheet = "Transcriptomics long",
  skip = 2,
  na = "NA"
) %>%
  filter(!(grepl("\\*", .$BEA_ID)))

mapping_23P074 <- readRDS(paste0(input_dir, "Transcriptomics/smpl_23P074.rds"))
mapping_24P003 <- read_tsv(
  paste0(input_dir, "Transcriptomics/F.Fagerstrom_24_07_sample_info.txt")
)
mapping_24P003_ <- read_tsv(paste0(input_dir, "Transcriptomics/mapping.txt"))

# NB! Tassos writes:
# BEA24P003_28 is FF06580592 from the ngi
# BEA24P003_28 is FF06580610 in meta
# based on the "Registrering Provrör ID CHLORHEX.xlxs" files it should be FF06580610

# Also BEA23P074_23, BEA23P074_3, BEA23P074_30, BEA23P074_20, BEA23P074_21 were sequenced in both runs
# for batch correction, meaning they have different NGI_ID but same BEA_ID

# Not sequenced due to low RIN values (n=10)
# BEA23P074_1, BEA23P074_2, BEA23P074_5, BEA23P074_6, BEA23P074_13, BEA23P074_36, BEA23P074_40,
# BEA24P003_11, BEA24P003_12, BEA24P003_22

batch <- c(
  "BEA23P074_23",
  "BEA23P074_3",
  "BEA23P074_30",
  "BEA23P074_20",
  "BEA23P074_21"
)

mapping_24P003_ <- mapping_24P003_ %>%
  mutate(
    NGI_ID = gsub("_[^_]*$", "", File),
    .keep = "unused",
    .before = Number_of_input_reads
  ) %>%
  dplyr::select(-...11)

mapping_24P003 <- mapping_24P003 %>%
  separate(`User ID`, into = c("BEA_ID", "Sample_ID"), sep = "_(?=[^_]+$)") %>%
  dplyr::select(NGI_ID = "NGI ID", BEA_ID, Sample_ID, Mreads, `>=Q30`) %>%
  mutate(BEA_ID = str_replace(.$BEA_ID, "i|_i", "")) %>%
  left_join(., mapping_24P003_) %>%
  mutate(
    Sample_ID = ifelse(.$BEA_ID == "BEA24P003_28", "FF06580610", .$Sample_ID)
  ) %>%
  left_join(
    .,
    select(
      smpl_barcodes,
      Sample_ID = "ORANGE (biopsi, mikrobiom, 800µl DNA/RNA shield)",
      DS_ID
    ),
    by = "Sample_ID"
  ) %>%
  left_join(
    .,
    select(smpl_IDs, svamp_ID, DS_ID = "patient_ID"),
    by = "DS_ID"
  ) %>%
  select(-Number_mapped_in_genes) %>%
  mutate(
    svamp_ID = ifelse(
      .$BEA_ID %in% batch,
      paste0(.$svamp_ID, "_run2"),
      .$svamp_ID
    )
  ) %>%
  left_join(
    .,
    select(meta_trx, Sample_ID = sample_ID, RIN),
    by = "Sample_ID"
  ) %>%
  relocate(DS_ID, svamp_ID, RIN, .after = "Sample_ID")


mapping_23P074 <- mapping_23P074 %>%
  as_tibble() %>%
  dplyr::select(
    NGI_ID,
    BEA_ID,
    Sample_ID = "vial",
    any_of(colnames(mapping_24P003[4:14])),
    Reads_in_genes
  ) %>%
  left_join(
    .,
    select(
      smpl_barcodes,
      Sample_ID = "ORANGE (biopsi, mikrobiom, 800µl DNA/RNA shield)",
      DS_ID
    ),
    by = "Sample_ID"
  ) %>%
  left_join(
    .,
    select(smpl_IDs, svamp_ID, DS_ID = "patient_ID"),
    by = "DS_ID"
  ) %>%
  left_join(
    .,
    select(meta_trx, Sample_ID = sample_ID, RIN),
    by = "Sample_ID"
  ) %>%
  relocate(DS_ID, svamp_ID, RIN, .after = "Sample_ID")

write_csv(mapping_23P074, "Tidy_data/Transcriptomics/Mapping_23P074.csv")
write_csv(mapping_24P003, "Tidy_data/Transcriptomics/Mapping_24P003.csv")

# Merge mapping files for run 1 (23P074) and run 2 (24P003)
mapping <- bind_rows(mapping_24P003, mapping_23P074) %>%
  mutate(
    svamp_ID = if_else(
      is.na(svamp_ID),
      paste0("NA_", cumsum(is.na(svamp_ID))),
      svamp_ID
    )
  )
mapping <- mapping %>%
  mutate(svamp_ID = str_replace(svamp_ID, "S75_Bl", "S75_BL")) %>%
  mutate(svamp_ID = str_replace(svamp_ID, "S69_1M", "S69_3M")) %>%
  arrange(svamp_ID) #%>%
#dplyr::summarise(n = dplyr::n(), .by = c("Sample_ID",  "DS_ID", "svamp_ID")) |>
#dplyr::filter(n > 1L)

setdiff(meta_trx$sample_ID, mapping$Sample_ID)

write_csv(mapping, "../Tidy_data/Transcriptomics/Mapping_all_trx_samples.csv")
```

### overviw of transcriptomics data
```{r}
#############
# LOAD DATA #
#############
meta_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metadata_svamp.xlsx"
meta_data <- read_xlsx(meta_path, sheet = "Metadata", skip = 1)

s <- smpl_IDs %>%
  select(svamp_ID, DS_ID = patient_ID) %>%
  mutate(ID_ = str_extract(DS_ID, "K\\d+|^\\d+")) %>%
  mutate(ID = str_extract(svamp_ID, "S\\d\\d")) %>%
  left_join(
    .,
    select(
      mapping,
      svamp_ID,
      BEA_code_transcriptomics_tissue = NGI_ID,
      barcode_transcriptomics_tissue = Sample_ID
    ),
    by = "svamp_ID"
  )

# meta data
meta_data <- meta_data %>%
  select(ID, Treatment, groups) %>%
  left_join(., unique(select(s, ID = ID_, svamp_ID = ID)))

# Transcriptomics samples wide overviw
m <- mapping %>%
  select(Sample_ID, svamp_ID_ = svamp_ID, DS_ID) %>%
  mutate(ID_ = str_extract(DS_ID, "K\\d+|^\\d+")) %>%
  left_join(
    .,
    select(meta_data, svamp_ID, ID_ = "ID", Treatment, groups),
    by = "ID_"
  ) %>%
  mutate(timepoint = str_match(svamp_ID_, "(?:^|_)(BL|3M)(?:_|$)")[, 2]) %>%
  mutate(timepoint = ifelse(is.na(.$timepoint), "BL", .$timepoint)) %>%

  select(-Sample_ID, -DS_ID) %>%
  # dplyr::summarise(n = dplyr::n(), .by = c(ID)) |>
  # dplyr::filter(n > 1L)
  pivot_wider(
    names_from = "timepoint",
    values_from = "svamp_ID_",
    values_fill = NA,
    values_fn = list(svamp_ID_ = ~ paste(unique(.), collapse = ", "))
  )
```

### get symbols for ensembel ids
```{r gene-annotation}
#############
# LOAD DATA #
#############
counts_24P003 <- readRDS(paste0(input_dir, "Transcriptomics/counts.rds"))
counts_23P074 <- readRDS(paste0(input_dir, "Transcriptomics/counts_23P074.rds"))
# counts_comb <- readRDS("/Users/vilkal/Downloads/counts_comb.rds")

####  Get symbol annotation
# 1. Get Ensembl annotation
mart <- useEnsembl("ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl")

anno_ens <- getBM(
  attributes = c("ensembl_gene_id", "entrezgene_id", "hgnc_symbol"),
  mart = mart
) %>%
  as_tibble()

# Collapse symbols and entrez IDs per gene
anno_ens_mod <- anno_ens %>%
  group_by(ensembl_gene_id) %>%
  summarise(
    symbol = paste(na.omit(unique(hgnc_symbol)), collapse = ","),
    entrez = paste(na.omit(unique(entrezgene_id)), collapse = ","),
    .groups = "drop"
  )

# 2. Get org.Hs.eg.db annotation
anno_org <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = anno_ens$ensembl_gene_id,
  columns = c("SYMBOL", "ENTREZID"),
  keytype = "ENSEMBL"
) %>%
  as_tibble()

# Collapse org.Hs.eg.db annotations
anno_org_mod <- anno_org %>%
  group_by(ENSEMBL) %>%
  summarise(
    ensembl = ENSEMBL[1],
    symbol = paste(na.omit(unique(SYMBOL)), collapse = ","),
    entrez = paste(na.omit(unique(ENTREZID)), collapse = ","),
    .groups = "drop"
  )

# 3. Fill missing symbols and entrez IDs using anno_ens_mod
anno_org_mod <- anno_org_mod %>%
  left_join(
    anno_ens_mod,
    by = c("ensembl" = "ensembl_gene_id"),
    suffix = c("", "_ens")
  ) %>%
  mutate(
    symbol = if_else(symbol == "" | is.na(symbol), symbol_ens, symbol),
    entrez = if_else(entrez == "" | is.na(entrez), entrez_ens, entrez)
  ) %>%
  dplyr::select(ensembl, symbol, entrez)

# 4. Annotate counts table
counts_24P003 <- counts_24P003 %>%
  as_tibble() %>%
  left_join(., anno_org_mod, by = "ensembl", ) %>%
  relocate(symbol, entrez, .after = Length) %>%
  dplyr::select(-rn)

counts_23P074 <- counts_23P074 %>%
  as_tibble() %>%
  dplyr::select(-rn)

# save count tables with NGI_ID
write_csv(counts_23P074, "../Tidy_data/Transcriptomics/Counts_23P074.csv")
write_csv(counts_24P003, "../Tidy_data/Transcriptomics/Counts_24P003.csv")
# write_csv(anno_org_mod, "data/gene_annotation.csv")
```

### Combine counts tabels and replace NGI id with svamp ids
```{r combine-and-save-trx-data}
#############
# LOAD DATA #
#############
mapping <- read_csv("../Tidy_data/Transcriptomics/Mapping_all_trx_samples.csv")
counts_23P074 <- read_csv("../Tidy_data/Transcriptomics/Counts_23P074.csv")
counts_24P003 <- read_csv("../Tidy_data/Transcriptomics/Counts_24P003.csv")
anno_org_mod <- read_csv("data/gene_annotation.csv")

ID_list <- set_names(mapping$svamp_ID, mapping$NGI_ID)

counts_23P074 <- counts_23P074 %>%
  rename_with(~ ID_list[.x], .cols = starts_with("P"))

counts_24P003 <- counts_24P003 %>%
  rename_with(~ ID_list[.x], .cols = starts_with("P"))

# combine count tables
# check if anny ensembl id's differ
setdiff(counts_23P074$ensembl, counts_24P003$ensembl)
setdiff(counts_24P003$ensembl, counts_23P074$ensembl)

# 1. Select metadata columns from counts
counts_metadata <- counts_23P074 %>%
  select(ensembl, Chr, Start, End, Strand, Length)

# 2. Select count columns based on sample sheet
count_data <- bind_cols(counts_23P074, counts_24P003) %>%
  select(all_of(mapping$svamp_ID)) %>%
  # colnames(.)
  select(sort(colnames(.)))

# 3. Combine metadata and count values
counts_comb <- bind_cols(counts_metadata, count_data)

# 4. Join with annotations (left join on ensembl)
counts_comb <- counts_comb %>%
  left_join(anno_org_mod, by = "ensembl") %>%
  relocate(symbol, entrez, .after = "Length")

# save count tables with svamp_ID
write_csv(
  counts_comb,
  "../Tidy_data/Transcriptomics/Counts_all_trx_samples.csv"
)
# counts_comb <- read_csv("Tidy_data/Counts_trx.csv")
```

## Format Metagenome data

### Format Metagenome sample info
```{r fix-metadata-metagenomics}
meta_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metadata_svamp.xlsx"
m_path <- "../data/Metagenomics/Identification_samples_RVVC_Gabriella.csv"
m2_path <- "../data/Metagenomics/Sample_identification_CHG_FLZ.csv"

meta_data <- read_xlsx(meta_path, sheet = "Metadata", skip = 1)
smpl_IDs <- read_xlsx(paste0(input_dir, "svamp_ID_list.xlsx")) %>%
  select(SD_ID = "patient_ID", svamp_ID, sample_ID = "barcode_shotgun_swab")
m_data <- read_csv(m_path, col_types = "ccccc")
m2_data <- read_csv(m2_path, col_types = "ccccc")

m <- full_join(m2_data, m_data, by = join_by(sample_ID, DS_ID))

# This skips formatting for `tribble()` and its arguments

m_ <- m %>%
  # mutate(
  #   # adds missing values from the two data frames:
  #   Treatment = case_when(
  #     is.na(.$Treatment) & !(is.na(.$group_assigned)) ~ "cross-sectional",
  #     .default = Treatment
  #   ),
  #   Timepoint = case_when(
  #     is.na(.$Timepoint) & !(is.na(.$group_assigned)) ~ "baseline",
  #     .default = Timepoint
  #   ),
  #   .before = "DS_ID"
  # ) %>%
  left_join(smpl_IDs, ., by = "sample_ID") %>%
  mutate(ID = str_extract(svamp_ID, "S\\d\\d")) %>%
  mutate(
    group_assigned = first(na.omit(group_assigned)),
    "Shotgun (n)" = dplyr::n(),
    .by = c(ID)
  ) %>%
  relocate(ID, DS_ID, svamp_ID, .after = "sample_ID") %>%
  rename(barcode_shotgun_swab = "sample_ID") %>%
  {
    . ->> meta_long
  } %>%
  select(-Group_assigned, -DS_ID) %>% # , -sample_ID
  mutate(value = "1") %>%
  pivot_wider(
    names_from = "Timepoint",
    values_from = "sample_ID",
    values_fill = "0"
  )

write_csv(
  meta_long,
  "../Tidy_data/Metagenomics/Key_all_shotgun_vagswab_samples.csv"
)
# meta_long <- read_csv("../Tidy_data/Metagenomics/Key_all_shotgun_vagswab_samples.csv")
```

```{r meta-data-meatgenomics}

smpl_IDs <- read_xlsx(
  paste0(input_dir, "svamp_ID_list.xlsx"),
  na = "NA"
) %>%
  select(DS_ID = "patient_ID", svamp_ID, barcode_shotgun_swab) %>%
  filter(!(is.na(barcode_shotgun_swab))) %>%
  mutate(
    ID = str_extract(svamp_ID, "S\\d\\d"),
    .after = "barcode_shotgun_swab"
  ) # %>% will add clinical info later
# left_join(
#   .,
#   select(meta_long, barcode_shotgun_swab, Timepoint),
#   by = c("barcode_shotgun_swab")
# ) %>%
# mutate(
#   Timepoint = case_when(
#     str_detect(svamp_ID, "_BL") | !str_detect(svamp_ID, "_") ~ "baseline",
#     str_detect(svamp_ID, "_1W$") ~ "1_week",
#     str_detect(svamp_ID, "_3M$") ~ "3_month",
#     str_detect(svamp_ID, "_6M$") ~ "6_month",
#     TRUE ~ Timepoint # fallback to existing value
#   )
# ) %>%
# left_join(., select(gr_data, ID = "svamp_ID", group), by = c("ID"))

```

### Vagswab data
#### Format count table

```{r Metagenom-vagswab}
#############
# LOAD DATA #
#############
run1_path <- paste0(input_dir, "Metagenomics/all_samples.metaphlan-4.tsv")
run2_path <- paste0(
  input_dir,
  "Metagenomics/all_samples.metaphlan_mikantarell_add19.tsv"
)
mbio_path <- paste0(input_dir, "Metagenomics/250321_mpa4_no_run_repeat.csv")

run1_data <- read_tsv(run1_path, skip = 1)
run2_data <- read_tsv(run2_path, skip = 1)
mbio_data_filt <- read_csv(mbio_path)

get_id.fun <- function(ID) {
  id <- str_extract(ID, "(?<=__).+(?=__)")
  name_df <- tibble(name = id) %>%
    mutate(
      uniqe_name = if (n() == 1) name else str_c(name, "_", row_number()),
      .by = "name"
    )
  names <- name_df$uniqe_name
  return(names)
}

na_id <- c("FF06741471", "FF06741465")
na_n <- c("KX63", "KX64")
ctrls <- select(run1_data, -starts_with(c("FF", "FS"))) %>% names(.)[-1]
ID_list <- set_names(
  c(smpl_IDs$svamp_ID, na_n),
  c(smpl_IDs$barcode_shotgun_swab, na_id)
)

run1_data <- run1_data %>%
  # captures the barcode id
  rename_with(~ get_id.fun(.x), .cols = starts_with("242")) %>%
  # exhange the barcode with the svamp_ID and
  # fix uniqe sample names:
  rename_with(~ ID_list[.x], .cols = starts_with("F"))

run2_data <- run2_data %>%
  rename_with(~ get_id.fun(.x), .cols = starts_with("242")) %>%
  rename_with(~ ID_list[.x], .cols = starts_with("F"))

colnames(run1_data)
colnames(run2_data)

# combine count tables
# check if anny ensembl id's differ
intersect(run1_data$clade_name, run2_data$clade_name)

setdiff(run1_data$clade_name, run2_data$clade_name)
setdiff(run2_data$clade_name, run1_data$clade_name)

# lenght of combined data frame should be:
nrow(run1_data) + length(setdiff(run2_data$clade_name, run1_data$clade_name))

# Safely combine using full_join by clade_name
merged_counts <- full_join(
  run1_data,
  run2_data,
  by = "clade_name",
  suffix = c("", "_run2")
) %>%
  # replace NAs with zeros:
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>%
  select(clade_name, contains("ex"), contains("seq"), sort(colnames(.)))


colnames(merged_counts)
write_csv(
  merged_counts,
  "../Tidy_data/Metagenomics/Metageonome_all_vagswab_metaphlan.csv"
)
# vagswab_counts <- read_csv("../Tidy_data/Metagenomics/Metageonome_all_vagswab_metaphlan.csv")
```

#### Format count info table
```{r count-info-vagswab}
################################
# COMBINE COUNT VAGSWAB TABLES #
################################
read_stat_path <- paste0(
  input_dir,
  "Metagenomics/preprocessing_read_counts.tsv"
)
read_stat2_path <- paste0(
  input_dir,
  "Metagenomics/preprocessing_read_counts_mikantarell_add19.tsv"
)
read_stat1 <- read_tsv(read_stat_path)
read_stat2 <- read_tsv(read_stat2_path)

# adds a number suffix to duplicate controlls and _run2 to samples
get_unique_id.fun <- function(ID) {
  # id <- str_extract(ID, "(?<=__).+(?=__)") used for barcode_shotgun_swab column
  name_df <- tibble(name = ID) %>%
    mutate(
      uniqe_name = if (n() == 1) name else str_c(name, "_", row_number()),
      .by = "name"
    ) %>%
    mutate(
      uniqe_name = case_when(
        str_starts(uniqe_name, "S") & str_ends(uniqe_name, "_1") ~
          str_remove(uniqe_name, "_1$"),
        str_starts(uniqe_name, "S") & str_ends(uniqe_name, "_2") ~
          str_replace(uniqe_name, "_2$", "_run2"),
        TRUE ~ uniqe_name
      )
    )
  names <- name_df$uniqe_name
  return(names)
}

n <- names(select(vagswab_counts, contains(c("ex", "seq"))))
s <- str_remove(n, "_\\d$")
n <- set_names(n, s)

na_id <- c("FF06741471" = "KX63", "FF06741465" = "KX64")

read_stat <- read_stat1 %>%
  bind_rows(., read_stat2) %>%
  mutate(barcode_shotgun_swab = str_extract(Sample, "(?<=__).+(?=__)")) %>%
  left_join(., smpl_IDs, by = "barcode_shotgun_swab") %>%
  relocate(
    barcode_shotgun_swab,
    DS_ID,
    svamp_ID,
    ID,
    #Timepoint,
    #group,
    .after = "Sample"
  ) %>%
  # fill in the control IDs in the svamp_ID column
  mutate(
    svamp_ID = if_else(
      str_detect(barcode_shotgun_swab, "_"),
      barcode_shotgun_swab,
      svamp_ID
    )
  ) %>%
  # fix uniqe control and sample names:
  mutate(
    svamp_ID = get_unique_id.fun(svamp_ID),
    barcode_shotgun_swab = if_else(
      str_detect(barcode_shotgun_swab, "_"),
      NA,
      barcode_shotgun_swab
    )
  ) %>%
  mutate(
    svamp_ID = if_else(
      str_detect(svamp_ID, "_3$"),
      str_replace(svamp_ID, "_3$", ""),
      svamp_ID
    )
  ) %>%
  # adds the two samples with weired ids:
  mutate(
    svamp_ID = ifelse(
      is.na(.$svamp_ID),
      na_id[.$barcode_shotgun_swab],
      .$svamp_ID
    ),
    DS_ID = ifelse(is.na(.$svamp_ID), na_id[.$barcode_shotgun_swab], .$DS_ID)
  )

setdiff(names(ID_list), read_stat$sample_ID)

# add CST types
CST <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metagenomics/260122_CST_output_ALL.csv"

ex <- tibble(
  svamp_ID = c(
    "S87_BL_run2",
    "S90_BL_run2",
    "S89_BL_run2",
    "S83_BL_run2",
    "S85_BL_run2",
    "S15"
  ),
  CST = c("IV-B", "I", "III", "III", "IV-B", "IV-C"),
  subCST = c("IV-B", "I-B", "III-B", "III-B", "IV-B", "IV-C3")
)

CST_data <- read_csv(CST) %>%
  select(svamp_ID = sampleID, CST, subCST) %>%
  bind_rows(., ex)

r <- read_stat %>% left_join(., CST_data, by = "svamp_ID") %>%
  arrange(svamp_ID)

# save file
write_csv(
  r,
  "../Tidy_data/Metagenomics/Read_count_info_vaginalswab.csv"
)
# read_stat <- read_csv("../Tidy_data/Metagenomics/Read_count_info_vaginalswab.csv")
# read_stat <- read_stat %>% select(-CST, -subCST)

```

S10 and S15 was exluded due to poor quality
(S10 due to too few counts and S15 due to total abundance < 60% after taxa filtering)
S86_1W stoped coming after two weeks
S65_3M missed the 3m visit

### Tissue data
#### Format count table

```{r Metagenome-tissue}
smpl_barcodes <- read_xlsx(
  paste0(input_dir, "Registrering Provrör ID CHLORHEX.xlsx"),
  skip = 1
)
mpath1 <- "../data/Metagenomics/biopsi/MGI_20241129_R1100600190018_E250043326 & E250043278.csv"
mpath2 <- "../data/Metagenomics/biopsi/BEA23P074_tissuesamples.csv"
mpath3 <- "../Tidy_data/Transcriptomics/Mapping_all_trx_samples.csv"

#############
# LOAD DATA #
#############

run1_path <- "../data/Metagenomics/biopsi/all_samples.metaphlantissue_mikantarell.csv"
run2_path <- "../data/Metagenomics/biopsi/all_samplesmetaphlantissuemikantarell.csv"
run1_data <- read_delim(run1_path, delim = ";")
run2_data <- read_delim(run2_path, delim = ";")

# run1_data and run2_data both contain 130 sample, however according to the
# sample id sheet from sequencing team "MGI_20241129_R1100600190018_E250043326 & E250043278.csv"
# there are only  49 (run 1, E250043278) and 43 (run 1, E250043326). Which means a total of 92 samples
# pluss 2 controls
# the remaning samples are probably from other projects that was sequenced at the same time

sample_info <- read_delim(mpath1, delim = ";") %>%
  filter(!is.na(adapter_id)) %>%
  mutate(Flowcell_L1_bc = paste0(flowcell_id, "_L01_", adapter_id)) %>% #92 samples
  # Flowcell_L1_bc and BEA_ID in this file has leading zeros, other files dont have this:
  mutate(
    sample_id = gsub("_0", "_", .$sample_id),
    Flowcell_L1_bc = gsub("_0", "_", .$Flowcell_L1_bc)
  )

meta2 <- read_delim(delim = ";", mpath2)
meta3 <- read_csv(mpath3)
smpl_IDs <- read_xlsx(paste0(input_dir, "svamp_ID_list.xlsx"))

#######################
# COMBINE SAMPLE INFO #
#######################

listn <- c(
  "BEA24P003_11" = "25:02",
  "BEA24P003_12" = "K33",
  "BEA24P003_22" = "29:02"
)

smplID <- sample_info %>%
  select(
    Flowcell_L1_bc,
    extraction_plate_id,
    BEA_ID = "sample_id",
    not_sequenced
  ) %>%

  # add barcode information:
  left_join(
    .,
    select(meta2, BEA_ID = "BEAID", barcode_shotgun_tissue = "Vial_number"),
    by = "BEA_ID"
  ) %>%

  # add study ID from the trx sample sheet:
  left_join(., select(meta3, BEA_ID, DS_ID), by = "BEA_ID") %>%

  # add missing barcode information:
  left_join(
    .,
    select(smpl_IDs, barcode = barcode_shotgun_tissue, DS_ID = patient_ID),
    by = c("DS_ID")
  ) %>%
  mutate(
    barcode_shotgun_tissue = ifelse(
      is.na(.$barcode_shotgun_tissue),
      .$barcode,
      .$barcode_shotgun_tissue
    )
  ) %>%
  # add control names
  mutate(
    barcode_shotgun_tissue = ifelse(
      is.na(.$barcode_shotgun_tissue),
      .$BEA_ID,
      .$barcode_shotgun_tissue
    )
  ) %>%
  select(-DS_ID, -barcode) %>%
  left_join(
    .,
    select(smpl_IDs, DS_ID = patient_ID, svamp_ID, barcode_shotgun_tissue),
    by = "barcode_shotgun_tissue"
  ) %>%

  mutate(
    DS_ID = ifelse(is.na(.$DS_ID), .$BEA_ID, .$DS_ID),
    svamp_ID = ifelse(is.na(.$svamp_ID), .$BEA_ID, .$svamp_ID)
  ) %>%
  filter(!(BEA_ID == "0")) %>%
  #dplyr::summarise(n = dplyr::n(), .by = c("svamp_ID")) |>
  #dplyr::filter(n > 1L) %>%
  unique()

write_csv(
  smplID,
  "../Tidy_data/Metagenomics/Key_all_shotgun_tissue_samples.csv"
)
# smplID <- read_csv( "../Tidy_data/Metagenomics/Key_all_shotgun_tissue_samples.csv")

########################################
# RENAME COULMS OF COUNT TISSUE TABLES #
########################################
# the count tables was delivered from the pipline with ids consisting of the flow cell id pluss the adapter id
# this needed to be replaced with the new study ID that Margaux have created
all_samples <- c(colnames(run1_data)[-1], colnames(run1_data)[-1])
our_samples <- all_samples[all_samples %in% smplID$Flowcell_L1_bc]
setdiff(smplID$Flowcell_L1_bc, our_samples)

# Ensure svamp_IDs are unique by appending a suffix to duplicates
svamp_ID_unique <- make.unique(smplID$svamp_ID, sep = "_extr")

# Create named vector: names = barcodes, values = unique svamp_IDs
ID_list <- set_names(svamp_ID_unique, smplID$Flowcell_L1_bc)


run1_data <- run1_data %>% # 48 samples
  # selects just the samples that belong to our project:
  select(clade_name, any_of(names(ID_list))) %>%
  # exhange the barcode with the svamp_ID
  rename_with(~ ID_list[.x], .cols = starts_with("E25"))

run2_data <- run2_data %>% # 48 samples
  # selects just the samples that belong to our project:
  select(clade_name, any_of(names(ID_list))) %>%
  # exhange the barcode with the svamp_ID
  rename_with(~ ID_list[.x], .cols = starts_with("E25"))

# total of 96 samples
#

###############################
# COMBINE COUNT TISSUE TABLES #
###############################
# combine count tables
# check if anny ensembl id's differ
intersect(run1_data$clade_name, run2_data$clade_name)

setdiff(run1_data$clade_name, run2_data$clade_name)
setdiff(run2_data$clade_name, run1_data$clade_name)

# lenght of combined data frame should be:
nrow(run1_data) + length(setdiff(run2_data$clade_name, run1_data$clade_name))

# Safely combine using full_join by clade_name
merged_counts <- full_join(
  run1_data,
  run2_data,
  by = "clade_name",
  suffix = c("", "_run2")
) %>%
  # replace NAs with zeros:
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>%
  select(clade_name, contains("seq"), sort(colnames(.)))

colnames(merged_counts)
write_csv(
  merged_counts,
  "../Tidy_data/Metagenomics/Metageonome_all_tissue_metaphlan.csv"
)

# tissue_counts <- read_csv("../Tidy_data/Metagenomics/Metageonome_all_tissue_metaphlan.csv")

```

#### Format count info table

```{r count-info-tissue}
##################################
# COMBINE READ COUNT TISSUE INFO #
##################################
# add read counts from processing file
# add suffix to distingush non-uniqe sample names

prepro_path1 <- "../data/Metagenomics/biopsi/preprocessing_read_counts_tissue_mikantarell.txt"
prepro_path2 <- "../data/Metagenomics/biopsi/preprocessing_read_counts_tissue_mikantarell2.txt"

prepro1 <- read_delim(prepro_path1, delim = "\t")
prepro2 <- read_delim(prepro_path2, delim = "\t")
# gr_data <- read_csv("../Tidy_data/Asigned_groups_info_all_samples.csv")

prepro <- bind_rows(prepro1, prepro2) %>%
  filter(Sample %in% names(ID_list)) %>%
  rename(Flowcell_L1_bc = Sample) %>%
  left_join(smplID, .) %>%
  # dplyr::summarise(n = dplyr::n(), .by = c("Flowcell_L1_bc")) |>
  # dplyr::filter(n > 1L) %>%
  mutate(ID = str_extract(svamp_ID, "S\\d\\d"), .before = "DS_ID") %>%
  # add "_extr1" to non uniqe svam_IDs:
  mutate(svamp_ID = ID_list[.$Flowcell_L1_bc]) %>%
  select(
    Flowcell_L1_bc,
    BEA_ID,
    barcode_shotgun_tissue,
    DS_ID,
    svamp_ID,
    everything()
  ) %>%
  mutate(
    svamp_ID = get_unique_id.fun(svamp_ID),
    barcode_shotgun_tissue = if_else(
      str_detect(barcode_shotgun_tissue, "_"),
      NA,
      barcode_shotgun_tissue
    )
  )

write_csv(prepro, "../Tidy_data/Metagenomics/Read_count_info_tissue.csv")
```

```{r merge-vagswab-and-tissue}
key_vagswab <- read_csv(
  "../Tidy_data/Metagenomics/Key_all_shotgun_vagswab_samples.csv"
)
key_tissue <- read_csv(
  "../Tidy_data/Metagenomics/Key_all_shotgun_tissue_samples.csv"
)

intersect(key_vagswab$svamp_ID, key_tissue$svamp_ID)

k <- key_vagswab %>%
  full_join(
    select(key_tissue, DS_ID, barcode_shotgun_tissue, BEA_ID, svamp_ID),
    by = c("svamp_ID", "DS_ID")
  ) %>%
  #left_join(select(key_tissue, svamp_ID, BEA_ID), by = "svamp_ID") %>%
  group_by(svamp_ID) %>%
  summarise(
    BEA_ID = paste(na.omit(unique(BEA_ID)), collapse = ", "),
    across(everything(), ~ first(.x)),
    .groups = "drop"
  ) %>%
  select(
    ID,
    DS_ID,
    svamp_ID,
    barcode_shotgun_swab = sample_ID,
    barcode_shotgun_tissue,
    BEA_ID,
    everything()
  )


write_csv(k, "../Tidy_data/Metagenomics/Readcounts.csv")
```

### Functional metagenome data
```{r functional-annotation}
#############
# LOAD DATA #
#############
# Biopsi samples
tissue_run1 <- read_tsv(paste0(
  input_dir,
  "Metagenomics/Functional profiling/",
  "all_samples.humann_pathabundance-5_tissue.txt"
))
tissue_run2 <- read_tsv(paste0(
  input_dir,
  "Metagenomics/Functional profiling/",
  "all_samples.humann_pathabundance-1_tissue.txt"
))

# vagswab run 1
vagswab_run1 <- read_tsv(paste0(
  input_dir,
  "Metagenomics/Functional profiling/",
  "all_samples.humann_pathabundance-4_run1.txt"
))
# vagswab run 2 additional 19 samples
vagswab_run2 <- read_tsv(paste0(
  input_dir,
  "Metagenomics/Functional profiling/",
  "all_samples.humann_pathabundance-4_run2.txt"
))

######################
# MERGE VAGSWAB DATA #
######################
# Safely combine using full_join by Pathway
tissue_data <- full_join(
  tissue_run1,
  tissue_run2,
  by = "# Pathway",
  suffix = c("", "_run2")
) %>%
  # replace NAs with zeros:
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>%
  select(`# Pathway`, contains("Ctrl"), sort(colnames(.)))

colnames(tissue_data)

# Safely combine using full_join by Pathway
vagswab_data <- full_join(
  vagswab_run1,
  vagswab_run2,
  by = "# Pathway",
  suffix = c("", "_run2")
) %>%
  # replace NAs with zeros:
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>%
  select(`# Pathway`, contains("ex"), contains("seq"), sort(colnames(.)))

colnames(vagswab_data)


```

```{r}
################
# GET SVAMP ID #
################
tissue_count_info <- read_csv(
  "../Results/Tissue_QC/Read_count_info_tissue_no_spikein.csv"
) %>%
  # remove samples that have zero bacterial reads after spike-in
  filter(bac_counts_no_spikein > 2000)
vagswab_count_info <- read_csv(
  "../Tidy_data/Metagenomics/Read_count_info_vaginalswab.csv"
)

ID_tissue <- set_names(
  tissue_count_info$svamp_ID,
  tissue_count_info$Flowcell_L1_bc
)
ID_vagswab <- set_names(vagswab_count_info$svamp_ID, vagswab_count_info$Sample)

setdiff(gsub("_Abundance", "", colnames(tissue_data)), names(ID_tissue))
intersect(gsub("_Abundance", "", colnames(tissue_data)), names(ID_tissue))

# rename columns with svamp ID
# tissue
# NB! we have to think through if it is nessecary to do some sort of filtering of this data before use
# I dont know if this is with or withouth the spike-in
# also only 47 samples match, I think we are missing a file with the other samples
tissue_data <- tissue_data %>%
  rename_with(
    ~ str_replace(.x, "_Abundance", ""),
    .cols = starts_with("E25")
  ) %>%
  rename_with(
    ~ ID_tissue[.x],
    .cols = names(.)[names(.) %in% names(ID_tissue)]
  ) %>%
  # remove samples that are not ours or had zero counts after spike-in removal
  select(-starts_with("E25")) %>%
  # order columns
  select(`# Pathway`, contains("Ctrl"), sort(colnames(.)))


vagswab_data <- vagswab_data %>%
  rename_with(
    ~ str_replace(.x, "_Abundance", ""),
    .cols = starts_with("242")
  ) %>%
  rename_with(~ ID_vagswab[.x], .cols = starts_with("242"))

############################
# SAVE PATH ABUNDANCE DATA #
############################
write_csv(
  tissue_data,
  "../Tidy_data/Metagenomics/Functional_annotation_all_tissue.csv"
)
write_csv(
  vagswab_data,
  "../Tidy_data/Metagenomics/Functional_annotation_all_vagswab.csv"
)
```


## Format Metabolomics data
```{r Metabolom}
#############
# LOAD DATA #
#############
mbol_path <- "../data/Metabolomics/UTF-8SMC_P0800_GCMS_Data.xlsx"
gr_id_path <- "../Tidy_data/Asigned_groups_info_all_samples.csv"

mbol_data <- read_xlsx(mbol_path)
smpl_IDs <- read_xlsx(paste0(input_dir, "svamp_ID_list.xlsx")) %>%
  select(svamp_ID, barcode_metabolomics) %>%
  filter(!(is.na(.$barcode_metabolomics)))
gr_id <- read_csv(gr_id_path)

left_join(
  .,
  select(smpl_IDs, svamp_ID, DS_ID = "patient_ID"),
  by = "DS_ID"
)

mbol_data <- mbol_data %>%
  select(
    `Compound Name`,
    #DS_ID = `Subject ID`,
    barcode_metabolomics = "customer sample Identifier",
    `Injection Standard MS`:`N-Acetylneuraminic Acid`
  ) %>%
  left_join(., smpl_IDs, by = "barcode_metabolomics") %>%
  relocate(svamp_ID, .before = "barcode_metabolomics") %>%
  mutate(
    svamp_ID = ifelse(
      grepl("^QC\\d", .$`Compound Name`),
      .$`Compound Name`,
      .$svamp_ID
    )
  )

write_csv(
  mbol_data,
  "../Tidy_data/Metabolomics/Metabolom_all_samples.csv"
)
```

## Cleaning up data
The metabolomic data was provided in a unformatted table, so compiling the data into appropriate format is required. Differences in sample number, metabolite nomenclature, etc.
```{r}
#############
# LOAD DATA #
#############
#---  Load sample info ----
smpl_IDs <- read_xlsx("../data/svamp_ID_list.xlsx") %>%
  select(svamp_ID, barcode_metabolomics)

#---  Load raw file from metabolomics sequencing facility ----
mbol_path <- "../data/Metabolomics/UTF-8SMC_P0800_GCMS_Data.xlsx"
RAW <- read_xlsx(mbol_path)


##############
# CLEAN DATA #
##############
# this code is based on Paulos analysis of the metabolomics data
# View first few column names and first column values
head(colnames(RAW), 30)
head(RAW[[1]], 30)

#--- 1️⃣ Identify structure of the table ----
# Find where annotation columns end (contains "order" keyword)
col_boundary <- which(str_detect(
  colnames(RAW),
  regex("order", ignore_case = TRUE)
))
# Find where sample data begins (first row that starts with "1" in first column)
row_boundary <- which(RAW[[1]] == "1")[1] - 1

#--- 2️⃣ Split into three logical components ----
# Sample metadata (left block, below header)
metadata <- RAW %>%
  slice((row_boundary + 1):n()) %>%
  select(all_of(1:col_boundary))

# Compound annotations (top block, right side)
annotations_raw <- RAW %>%
  slice(1:row_boundary) %>%
  select(1, all_of((col_boundary + 1):ncol(RAW)))
#select(1, all_of((col_boundary + 1):ncol(RAW)))

# Measurement matrix (sample × metabolite intensities)
matrix_data <- RAW %>%
  slice((row_boundary + 1):n()) %>%
  select(
    `Compound Name`,
    "barcode_metabolomics" = `customer sample Identifier`,
    (col_boundary + 1):ncol(RAW)
  ) %>%
  left_join(., smpl_IDs, by = "barcode_metabolomics") %>%
  relocate(svamp_ID, .before = "barcode_metabolomics") %>%
  mutate(
    svamp_ID = ifelse(
      grepl("^QC\\d", .$`Compound Name`),
      .$`Compound Name`,
      .$svamp_ID
    )
  ) %>%
  select(-`Compound Name`, -barcode_metabolomics) %>%
  pivot_longer(cols = -svamp_ID, names_to = "Compund") %>%
  pivot_wider(names_from = svamp_ID) %>%
  select(Compund, starts_with("QC"), sort(colnames(.)))


#--- 3️⃣ Build tidy annotation table ----
# Extract selected annotation rows and pivot them into a tidy format
annotation_table <- annotations_raw %>%
  #rename(row_label=`GCMS-NA@NA`) %>%
  slice(., c(2:6)) %>% # rows with HMDB, KEGG, PubChem, Mz, Rt
  pivot_longer(-1, names_to = "names", values_to = "value") %>%
  pivot_wider(names_from = "Compound Name", values_from = value) %>%
  # Create uniqe compound IDs by using M/z and Rt
  mutate(
    Compound_ID = paste0(
      "GCMS-",
      round(as.numeric(`M/z (unique)`)),
      "@",
      round(as.numeric(`RI (resolved)`))
    )
  )

#--- 4️⃣ Remove quality control compunds ----
annotation_table <- annotation_table %>%
  filter(!grepl("Standard", .$names))

# ✅ Results:
# metadata         → sample metadata (e.g. sample ID, run order)
# matrix_data      → numeric matrix (samples × compounds)
# annotation_table → clean annotation metadata per compound

#############
# SAVE DATA #
#############
write_csv(matrix_data, "../Tidy_data/Metabolomics/Metabolom_matrix.csv")
write_csv(metadata, "../Tidy_data/Metabolomics/metabolomics_metadata.csv")
write_csv(
  annotation_table,
  "../Tidy_data/Metabolomics/metabolomics_annotation.csv"
)
```