
### Load data and libraries
```{r Load_data}
##################
# LOAD LIBRARIES #
##################
library(tidyverse)
library(readxl)
# library(org.Hs.eg.db)
library(biomaRt)

select <- dplyr::select
rename <- dplyr::rename
# setwd("/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/src")

#########
# PATHS #
#########
input_dir <- "../data/"

meta_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metadata_svamp.xlsx"
mbol_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/UTF-8SMC_P0800_GCMS_Data.xlsx"
```

To determine the groups you need column from the clinical file

"Fungal culture: C. albicans (y/n)", 
"Symptom score (0-5)", 
"Recurring fungal infections > 2/year (y/n)"

RVVCpos:	Positive fungal cultures, symptom score ≥1, >2 episodes of VVC per year
Control:	Negative fungal culture, ≤2 episodes of VVC per year
AS:     	Positive fungal scores, symptom score = 0, ≤2 episodes of VVC per year
RVVCneg:	Negative fungal culture, >2 episodes of VVC per year

## get group assignment

```{r group-assignment}
meta_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metadata_svamp.xlsx"
meta_data <- read_xlsx(meta_path, sheet = "Metadata", skip = 1)

gr_data <- meta_data %>%
  select(
    DS_ID = "ID",
    svamp_ID,
    Treatment,
    `Fungal culture: C. albicans (y/n)`,
    `Symptom score (0-5)`,
    `Recurring fungal infections > 2/year (y/n)`
  ) %>%
  mutate(
    group = case_when(
      `Fungal culture: C. albicans (y/n)` == "1" &
        `Symptom score (0-5)` >= 1 &
        `Recurring fungal infections > 2/year (y/n)` == "1" ~
        "RVVCpos",

      `Fungal culture: C. albicans (y/n)` == "0" &
        `Recurring fungal infections > 2/year (y/n)` == "1" ~
        "RVVCneg",

      `Fungal culture: C. albicans (y/n)` == "1" &
        `Symptom score (0-5)` == 0 ~
        "AS",

      `Fungal culture: C. albicans (y/n)` == "0" &
        `Recurring fungal infections > 2/year (y/n)` == "0" ~
        "Control",

      `Fungal culture: C. albicans (y/n)` == "1" &
        `Recurring fungal infections > 2/year (y/n)` == "0" ~
        "Candidapos",

      TRUE ~ NA_character_
    )
  ) %>%
  select(ID = "svamp_ID", everything())

write_csv(gr_data, "../Tidy_data/Asigned_groups_info_all_samples.csv")
# gr_data <- read_csv("../Tidy_data/Asigned_groups_info_all_samples.csv")
```

## Format transcriptomics data

Steps performed bellow:
1. Combine all transcriptomics related information into one mapping file for all three runs
2. Merge datasets
3. Annotate (add symbols) for all datasets using Ensembl 113

The two first sequencing runs were orginaly referenced against version 101 of the Ensembl reference genome.
However, after writing to Tassos, he re-generated the counts files for the first two runs from the new (113) reference.
This makes it a lot easier to merge the three dataset as they now have identical rows.

Originaly the number of genes present in the older sequencing runs was 60671 genes vs 78932 for the new, 
because the counts were generated against differnt versions of the Ensembl reference (101 and 113):
    24P003  → Homo_sapiens.GRCh38.101  
    23P074  → Homo_sapiens.GRCh38.101  
    26P064  → Homo_sapiens.GRCh38.113
If this were the case still we would have needet to only select the genes that are present in both versions.

At the time of writing (23.09.2026) The newest rlease is 116 from ensembl, however they are still using 113 at BEA so we will use this version when getting the symbols.

#### Notes:
I was a bit surprised to learn that all the samples from the first sequencing run were re run? at least there are two sequencing files that tassos calls "run1" and "run2". In a email he writes:

Jo, det var prover som sekvenserades två gånger, så det borde vara samma prover. Så man borde sedan summera run1 och run2 counts när man importerart i R tex. Men som en sanity check, även om prover och genes skall vara desamma mellan båda, så borde de individuella counts variera något, so kolla i själva filerna för de första genera typ at det är olika counts mellan filerna.

I dicided to simply select the counts generated from the second run. I dont know if this is stupid, but it was a lot less complicated. 

There is something off with the naming of sample 12 or S69 it has 12:01 listed as baseline and a 1 month follow up which I have not seen any where else
It turns out 12:01 is a baseline sample that was mistakenly named :01 instead of 02 as it should have ben, and the 1 week follow up should be three month follow up

NB! Tassos writes:
BEA24P003_28 is FF06580592 in the F.Fagerstrom_24_07_sample_info.txt file but
BEA24P003_28 is FF06580610 in meta
based on the "Registrering Provrör ID CHLORHEX.xlxs" files it should be FF06580610
I have now updated F.Fagerstrom_24_07_sample_info.txt with the correct information

```{r load-trx-data}
##################
# LOAD LIBRARIES #
##################
library(tidyverse)
library(biomaRt)
library(org.Hs.eg.db)
library(AnnotationDbi)

#############################
# 1. Load count data #
#############################
input_dir <- "../data/Transcriptomics/"
files <- c(
  counts_23P074 = "BEA23P074run2_subreadCounts_hg38ens_minus_frag.txt",
  counts_24P003 = "BEA24P003_subreadCounts_hg38ens_minus_frag.txt",
  counts_26P064 = "BEA26P064_subreadCounts_hg38ens_minus_frag.txt"
)

counts <- map(
  files,
  ~ read_delim(
    paste0(input_dir, .x),
    delim = "\t",
    comment = "#"
  ) %>%
    rename(ensembl = Geneid)
)

######################
# Load Mapping files #
######################
smpl_IDs <- read_xlsx(
  "../Tidy_data/Transcriptomics/BEA_Extraction_RNA_DNA_QC.xlsx",
  skip = 1,
  na = "NA"
)

mapping_id <- c(
  BEA23P074run2 = "23",
  BEA24P003 = "24_07",
  BEA26P064 = "26_31"
) %>%
  map(., ~ read_tsv(paste0(input_dir, "F.Fagerstrom_", .x, "_sample_info.txt")))

# we choose the second round of the BEA23P074 samples:
mapping <- c("BEA23P074run2", "BEA24P003", "BEA26P064") %>%
  set_names() %>%
  map(., ~ read_tsv(paste0(input_dir, .x, "_mapping.txt"))) %>%
  bind_rows() %>%
  mutate(`NGI ID` = str_extract(File, "^[^_]+_[^_]+"), .after = "File")
```

### Step 1. Combine all transcriptomics related information

```{r get-ID-key}
id <- imap(
  mapping_id,
  # ~ if (.y == "BEA26P064") {
  #   mutate(.x, `BEA ID` = gsub("^[^_]*_[^_]*_", "", `User ID`), ID = `User ID`)
  # } else {
  ~ mutate(
    .x,
    ID = gsub("^[^_]*_[^_]*_(?:i_)?", "", `User ID`),
    `BEA ID` = sub("^(.*)_(\\d+)i?_.*$", "\\1_\\2", `User ID`),
    .keep = "unused"
  )
  #}
) %>%
  map(., ~ select(.x, `NGI ID`, `BEA ID`, ID)) %>%
  bind_rows() %>%
  # add svamp_ID:
  left_join(., select(smpl_IDs, `BEA ID`, svamp_ID, ), by = "BEA ID") %>%
  # Make NA samples unique:
  mutate(
    svamp_ID = ifelse(
      is.na(svamp_ID),
      paste0("NA_", cumsum(is.na(svamp_ID))),
      svamp_ID
    )
  )
```

```{r add-batch-info}
# add batch run:
Batch_2 <- c(
  "P31315_1056",
  "P31315_1053",
  "P31315_1057",
  "P31315_1054",
  "P31315_1055"
)
Batch_3 <- c(
  "P41356_2017",
  "P41356_2018",
  "P41356_2019",
  "P41356_2020",
  "P41356_2021",
  "P41356_2022",
  "P41356_2023",
  "P41356_2024"
)

id <- id %>%
  arrange(svamp_ID) %>%
  mutate(
    svamp_ID = ifelse(
      `NGI ID` %in% Batch_2,
      paste0(.$svamp_ID, "_run2"),
      .$svamp_ID
    )
  ) %>%
  mutate(
    svamp_ID = ifelse(
      `NGI ID` %in% Batch_3,
      paste0(.$svamp_ID, "_run3"),
      .$svamp_ID
    )
  )
```

```{r combine-all-mapping-info}
id <- id %>%
  left_join(., select(smpl_IDs, `BEA ID`, RIN), by = "BEA ID") %>%
  left_join(., select(mapping, everything()), by = "NGI ID")
```

```{r update-svamp_ID_list.xlsx}
svamp_ID_list <- read_xlsx(
  "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/svamp_ID_list.xlsx",
  na = "NA"
) %>%
  select(
    patient_ID,
    svamp_ID,
    barcode_shotgun_tissue,
    BEA_code_transcriptomics_tissue
  )
id_ <- id %>%
  select(ID, File, `BEA ID`) %>%
  group_by(ID) %>%
  summarise(
    `BEA ID` = paste(`BEA ID`, collapse = ", "),
    File = paste(File, collapse = ", "),
    .groups = "drop"
  )
svamp_ID_list_ <- svamp_ID_list %>%
  left_join(
    .,
    select(id_, ID, File, `BEA ID`),
    by = c("barcode_shotgun_tissue" = "ID")
  )
```

### Step 2. Merge counts tabels and replace NGI id with svamp ids
```{r merge-count-tables}
# Define common genes across all studies
# to dubble check that they are the same
common_genes <- reduce(
  map(counts, pull, ensembl),
  intersect
)
length(common_genes)
map(counts, ~ dim(.x))

# Merge datasets on ensembl id
trx_merged <- reduce(
  counts,
  full_join,
  by = c("ensembl", "Chr", "Start", "End", "Strand", "Length")
)

################
# Get Svamp_ID #
################
ID_list <- set_names(id$svamp_ID, id$`NGI ID`)

trx_merged_ <- trx_merged %>%
  # remove adapter id or something:
  rename_with(~ str_extract(.x, "^[^_]+_[^_]+"), .cols = starts_with("P")) %>%
  rename_with(~ ID_list[.x], .cols = starts_with("P"))

colnames(trx_merged_)
```

### Step 3. get symbols for ensembel ids

This workflow is more or less a tidyverse version of what Tassos does:
1. Get Ensembl annotation
2. Get org.Hs.eg.db annotation. 
3. Collapse duplicates.
4. Use org.Hs.eg.db as primary.
5. Fill missing values from Ensembl.

Additionaly Tassos explicitly checks agreement between the two databases before merging, by creating a column that says true false.

This workflow was designed to maximize annotation coverage by combining two databases
I have decided to replace this workflow with a cleaner, easier to reproduce, workflow that is based on Olgas (NBIS) code.  
It avoids potential disagreements between Ensembl and org.Hs.eg.db and is more reproducible. 
```{r gene-annotation-Tassos}
#| eval: false
#| code-fold: true
#############
# LOAD DATA #
#############
counts_24P003 <- readRDS(paste0(input_dir, "Transcriptomics/counts.rds"))
counts_23P074 <- readRDS(paste0(input_dir, "Transcriptomics/counts_23P074.rds"))
# counts_comb <- readRDS("/Users/vilkal/Downloads/counts_comb.rds")

####  Get symbol annotation
# 1. Get Ensembl annotation
mart <- useEnsembl("ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl")

anno_ens <- getBM(
  attributes = c("ensembl_gene_id", "entrezgene_id", "hgnc_symbol"),
  mart = mart
) %>%
  as_tibble()

# Collapse symbols and entrez IDs per gene
anno_ens_mod <- anno_ens %>%
  group_by(ensembl_gene_id) %>%
  summarise(
    symbol = paste(na.omit(unique(hgnc_symbol)), collapse = ","),
    entrez = paste(na.omit(unique(entrezgene_id)), collapse = ","),
    .groups = "drop"
  )

# 2. Get org.Hs.eg.db annotation
anno_org <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = anno_ens$ensembl_gene_id,
  columns = c("SYMBOL", "ENTREZID"),
  keytype = "ENSEMBL"
) %>%
  as_tibble()

# Collapse org.Hs.eg.db annotations
anno_org_mod <- anno_org %>%
  group_by(ENSEMBL) %>%
  summarise(
    ensembl = ENSEMBL[1],
    symbol = paste(na.omit(unique(SYMBOL)), collapse = ","),
    entrez = paste(na.omit(unique(ENTREZID)), collapse = ","),
    .groups = "drop"
  )

# 3. Fill missing symbols and entrez IDs using anno_ens_mod
anno_org_mod <- anno_org_mod %>%
  left_join(
    anno_ens_mod,
    by = c("ensembl" = "ensembl_gene_id"),
    suffix = c("", "_ens")
  ) %>%
  mutate(
    symbol = if_else(symbol == "" | is.na(symbol), symbol_ens, symbol),
    entrez = if_else(entrez == "" | is.na(entrez), entrez_ens, entrez)
  ) %>%
  dplyr::select(ensembl, symbol, entrez)

# 4. Annotate counts table
counts_24P003 <- counts_24P003 %>%
  as_tibble() %>%
  left_join(., anno_org_mod, by = "ensembl", ) %>%
  relocate(symbol, entrez, .after = Length) %>%
  dplyr::select(-rn)

counts_23P074 <- counts_23P074 %>%
  as_tibble() %>%
  dplyr::select(-rn)

# save count tables with NGI_ID
write_csv(counts_23P074, "../Tidy_data/Transcriptomics/Counts_23P074.csv")
write_csv(counts_24P003, "../Tidy_data/Transcriptomics/Counts_24P003.csv")
# write_csv(anno_org_mod, "data/gene_annotation.csv")
```

I wanted to use version 113 of ensemble as its the same as BEA used as reference, however I cannot get the connection with the ensembl server to work except for when I leave the version flagg empty. This means that I will be using the latest version 116.
```{r gene-annotation-new}
############################################################
# 1. Connect to a specific Ensembl release (recommended)   #
############################################################

# Using a fixed release makes the annotation reproducible.
# Replace the release number if you want to match the
# Ensembl version used during read alignment/counting.

# List latest versions:
latest <- biomaRt::listEnsemblArchives()[1:4, ]
latest

annotation_metadata <- list(
  date_accessed = Sys.Date(),
  biomart_version = as.character(packageVersion("biomaRt")),
  host = latest[2, 4]
)

annotation_metadata

# select the newest ensemble version:
mart <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl",
  version = 116
)


############################################################
# 2. Create a table containing the Ensembl IDs to annotate #
############################################################

# Assuming your data frame contains a column called 'ensembl'

df_genes <- tibble(
  ensembl_gene_id = unique(c(
    trx_merged_$ensembl
  ))
)

###########################################################
# 3. Retrieve annotation only for genes in your dataset   #
###########################################################

# hgnc_symbol      = official HGNC gene symbol
# external_gene_name = often identical to symbol but not always
# gene_biotype     = protein_coding, lncRNA, etc.
# entrezgene_id    = optional additional identifier

anno <- getBM(
  attributes = c(
    "ensembl_gene_id",
    "hgnc_symbol",
    "gene_biotype",
    "entrezgene_id"
  ),
  filters = "ensembl_gene_id",
  values = df_genes$ensembl_gene_id,
  mart = mart
)

##########################################
# 4. Check whether duplicate mappings    #
#    are present                         #
##########################################

dup_genes <- anno %>%
  group_by(ensembl_gene_id) %>%
  summarise(
    n_rows = n(),
    hgnc_symbols = paste((unique(na.omit(hgnc_symbol))), collapse = ","),
    gene_biotype = paste((unique(na.omit(gene_biotype))), collapse = ","),
    .groups = "drop"
  ) %>%
  filter(n_rows > 1) %>%
  mutate(
    n_symbol = if_else(
      hgnc_symbols == "",
      0L,
      str_count(hgnc_symbols, ",") + 1L
    )
  ) %>%
  arrange(desc(n_symbol))

cat("Number of Ensembl IDs with >1 annotation:", nrow(dup_genes), "\n")

# Inspect duplicate cases if any exist
dup_genes

###########################################################
# 5. Resolve duplicate rows (usually very few if any)     #
###########################################################

# For most RNA-seq datasets an Ensembl gene ID corresponds
# to one HGNC symbol.
#
# Here we keep one row per Ensembl gene while preserving
# all available symbols if multiple exist.

anno_clean <- anno %>%
  group_by(ensembl_gene_id) %>%
  summarise(
    symbol = paste(
      unique(na.omit(hgnc_symbol)),
      collapse = ","
    ),
    # gene_name = paste(
    #   unique(na.omit(external_gene_name)),
    #   collapse = ","
    # ),
    gene_biotype = dplyr::first(gene_biotype),
    entrezgene_id = paste(
      unique(na.omit(entrezgene_id)),
      collapse = ","
    ),
    .groups = "drop"
  ) %>%
  mutate(
    # gene_name = na_if(gene_name, ""),
    symbol = na_if(symbol, "")
  )

###########################################################
# 7. Join annotation back to the transcriptomics table    #
###########################################################

trx_annotated <- list(trx_merged_) %>%
  map(
    ~ .x %>%
      left_join(
        anno_clean,
        by = c("ensembl" = "ensembl_gene_id")
      ) %>%
      relocate(
        symbol,
        gene_biotype,
        entrez = entrezgene_id,
        .after = ensembl
      )
  )

###########################################################
# 8. Quick quality-control summaries                      #
###########################################################

map_dfr(
  trx_annotated,
  ~ tibble(
    # How many genes?
    n_genes = nrow(.x),
    # How many genes have a symbol?
    n_symbol = sum(!is.na(.x$symbol)),
    # How many genes have no symbol?
    n_no_symbol = sum(is.na(.x$symbol)),
    pct_symbol = mean(!is.na(.x$symbol)) * 100
  ),
  .id = "dataset"
)

# Which biotypes are present?
sort(table(trx_annotated[[1]]$gene_biotype), decreasing = T)
```


```{r save-count-table}
# save count tables with svamp_ID
write_csv(
  trx_merged_,
  "../Tidy_data/Transcriptomics/Counts_all_trx_samples.csv"
)
# counts_comb <- read_csv("Tidy_data/Counts_trx.csv")

write_csv(
  id,
  "../Tidy_data/Transcriptomics/Mapping_all_trx_samples.csv"
)
```


## Format Metagenome data

### Format Metagenome sample info
```{r fix-metadata-metagenomics}
meta_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metadata_svamp.xlsx"
m_path <- "../data/Metagenomics/Identification_samples_RVVC_Gabriella.csv"
m2_path <- "../data/Metagenomics/Sample_identification_CHG_FLZ.csv"

meta_data <- read_xlsx(meta_path, sheet = "Metadata", skip = 1)
smpl_IDs <- read_xlsx(paste0(input_dir, "svamp_ID_list.xlsx")) %>%
  select(SD_ID = "patient_ID", svamp_ID, sample_ID = "barcode_shotgun_swab")
m_data <- read_csv(m_path, col_types = "ccccc")
m2_data <- read_csv(m2_path, col_types = "ccccc")

m <- full_join(m2_data, m_data, by = join_by(sample_ID, DS_ID))

# This skips formatting for `tribble()` and its arguments

m_ <- m %>%
  # mutate(
  #   # adds missing values from the two data frames:
  #   Treatment = case_when(
  #     is.na(.$Treatment) & !(is.na(.$group_assigned)) ~ "cross-sectional",
  #     .default = Treatment
  #   ),
  #   Timepoint = case_when(
  #     is.na(.$Timepoint) & !(is.na(.$group_assigned)) ~ "baseline",
  #     .default = Timepoint
  #   ),
  #   .before = "DS_ID"
  # ) %>%
  left_join(smpl_IDs, ., by = "sample_ID") %>%
  mutate(ID = str_extract(svamp_ID, "S\\d\\d")) %>%
  mutate(
    group_assigned = first(na.omit(group_assigned)),
    "Shotgun (n)" = dplyr::n(),
    .by = c(ID)
  ) %>%
  relocate(ID, DS_ID, svamp_ID, .after = "sample_ID") %>%
  rename(barcode_shotgun_swab = "sample_ID") %>%
  {
    . ->> meta_long
  } %>%
  select(-Group_assigned, -DS_ID) %>% # , -sample_ID
  mutate(value = "1") %>%
  pivot_wider(
    names_from = "Timepoint",
    values_from = "sample_ID",
    values_fill = "0"
  )

write_csv(
  meta_long,
  "../Tidy_data/Metagenomics/Key_all_shotgun_vagswab_samples.csv"
)
# meta_long <- read_csv("../Tidy_data/Metagenomics/Key_all_shotgun_vagswab_samples.csv")
```

```{r meta-data-meatgenomics}

smpl_IDs <- read_xlsx(
  paste0(input_dir, "svamp_ID_list.xlsx"),
  na = "NA"
) %>%
  select(DS_ID = "patient_ID", svamp_ID, barcode_shotgun_swab) %>%
  filter(!(is.na(barcode_shotgun_swab))) %>%
  mutate(
    ID = str_extract(svamp_ID, "S\\d\\d"),
    .after = "barcode_shotgun_swab"
  ) # %>% will add clinical info later
# left_join(
#   .,
#   select(meta_long, barcode_shotgun_swab, Timepoint),
#   by = c("barcode_shotgun_swab")
# ) %>%
# mutate(
#   Timepoint = case_when(
#     str_detect(svamp_ID, "_BL") | !str_detect(svamp_ID, "_") ~ "baseline",
#     str_detect(svamp_ID, "_1W$") ~ "1_week",
#     str_detect(svamp_ID, "_3M$") ~ "3_month",
#     str_detect(svamp_ID, "_6M$") ~ "6_month",
#     TRUE ~ Timepoint # fallback to existing value
#   )
# ) %>%
# left_join(., select(gr_data, ID = "svamp_ID", group), by = c("ID"))
```

### Vagswab data
#### Format count table

```{r Metagenom-vagswab}
#############
# LOAD DATA #
#############
run1_path <- paste0(input_dir, "Metagenomics/all_samples.metaphlan-4.tsv")
run2_path <- paste0(
  input_dir,
  "Metagenomics/all_samples.metaphlan_mikantarell_add19.tsv"
)
mbio_path <- paste0(input_dir, "Metagenomics/250321_mpa4_no_run_repeat.csv")

run1_data <- read_tsv(run1_path, skip = 1)
run2_data <- read_tsv(run2_path, skip = 1)
mbio_data_filt <- read_csv(mbio_path)

get_id.fun <- function(ID) {
  id <- str_extract(ID, "(?<=__).+(?=__)")
  name_df <- tibble(name = id) %>%
    mutate(
      uniqe_name = if (n() == 1) name else str_c(name, "_", row_number()),
      .by = "name"
    )
  names <- name_df$uniqe_name
  return(names)
}

na_id <- c("FF06741471", "FF06741465")
na_n <- c("KX63", "KX64")
ctrls <- select(run1_data, -starts_with(c("FF", "FS"))) %>% names(.)[-1]
ID_list <- set_names(
  c(smpl_IDs$svamp_ID, na_n),
  c(smpl_IDs$barcode_shotgun_swab, na_id)
)

run1_data <- run1_data %>%
  # captures the barcode id
  rename_with(~ get_id.fun(.x), .cols = starts_with("242")) %>%
  # exhange the barcode with the svamp_ID and
  # fix uniqe sample names:
  rename_with(~ ID_list[.x], .cols = starts_with("F"))

run2_data <- run2_data %>%
  rename_with(~ get_id.fun(.x), .cols = starts_with("242")) %>%
  rename_with(~ ID_list[.x], .cols = starts_with("F"))

colnames(run1_data)
colnames(run2_data)

# combine count tables
# check if anny ensembl id's differ
intersect(run1_data$clade_name, run2_data$clade_name)

setdiff(run1_data$clade_name, run2_data$clade_name)
setdiff(run2_data$clade_name, run1_data$clade_name)

# lenght of combined data frame should be:
nrow(run1_data) + length(setdiff(run2_data$clade_name, run1_data$clade_name))

# Safely combine using full_join by clade_name
merged_counts <- full_join(
  run1_data,
  run2_data,
  by = "clade_name",
  suffix = c("", "_run2")
) %>%
  # replace NAs with zeros:
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>%
  select(clade_name, contains("ex"), contains("seq"), sort(colnames(.)))


colnames(merged_counts)
write_csv(
  merged_counts,
  "../Tidy_data/Metagenomics/Metageonome_all_vagswab_metaphlan.csv"
)
# vagswab_counts <- read_csv("../Tidy_data/Metagenomics/Metageonome_all_vagswab_metaphlan.csv")
```

#### Format count info table
```{r count-info-vagswab}
################################
# COMBINE COUNT VAGSWAB TABLES #
################################
read_stat_path <- paste0(
  input_dir,
  "Metagenomics/preprocessing_read_counts.tsv"
)
read_stat2_path <- paste0(
  input_dir,
  "Metagenomics/preprocessing_read_counts_mikantarell_add19.tsv"
)
read_stat1 <- read_tsv(read_stat_path)
read_stat2 <- read_tsv(read_stat2_path)

# adds a number suffix to duplicate controlls and _run2 to samples
get_unique_id.fun <- function(ID) {
  # id <- str_extract(ID, "(?<=__).+(?=__)") used for barcode_shotgun_swab column
  name_df <- tibble(name = ID) %>%
    mutate(
      uniqe_name = if (n() == 1) name else str_c(name, "_", row_number()),
      .by = "name"
    ) %>%
    mutate(
      uniqe_name = case_when(
        str_starts(uniqe_name, "S") & str_ends(uniqe_name, "_1") ~
          str_remove(uniqe_name, "_1$"),
        str_starts(uniqe_name, "S") & str_ends(uniqe_name, "_2") ~
          str_replace(uniqe_name, "_2$", "_run2"),
        TRUE ~ uniqe_name
      )
    )
  names <- name_df$uniqe_name
  return(names)
}

n <- names(select(vagswab_counts, contains(c("ex", "seq"))))
s <- str_remove(n, "_\\d$")
n <- set_names(n, s)

na_id <- c("FF06741471" = "KX63", "FF06741465" = "KX64")

read_stat <- read_stat1 %>%
  bind_rows(., read_stat2) %>%
  mutate(barcode_shotgun_swab = str_extract(Sample, "(?<=__).+(?=__)")) %>%
  left_join(., smpl_IDs, by = "barcode_shotgun_swab") %>%
  relocate(
    barcode_shotgun_swab,
    DS_ID,
    svamp_ID,
    ID,
    #Timepoint,
    #group,
    .after = "Sample"
  ) %>%
  # fill in the control IDs in the svamp_ID column
  mutate(
    svamp_ID = if_else(
      str_detect(barcode_shotgun_swab, "_"),
      barcode_shotgun_swab,
      svamp_ID
    )
  ) %>%
  # fix uniqe control and sample names:
  mutate(
    svamp_ID = get_unique_id.fun(svamp_ID),
    barcode_shotgun_swab = if_else(
      str_detect(barcode_shotgun_swab, "_"),
      NA,
      barcode_shotgun_swab
    )
  ) %>%
  mutate(
    svamp_ID = if_else(
      str_detect(svamp_ID, "_3$"),
      str_replace(svamp_ID, "_3$", ""),
      svamp_ID
    )
  ) %>%
  # adds the two samples with weired ids:
  mutate(
    svamp_ID = ifelse(
      is.na(.$svamp_ID),
      na_id[.$barcode_shotgun_swab],
      .$svamp_ID
    ),
    DS_ID = ifelse(is.na(.$svamp_ID), na_id[.$barcode_shotgun_swab], .$DS_ID)
  )

setdiff(names(ID_list), read_stat$sample_ID)

# add CST types
CST <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metagenomics/260122_CST_output_ALL.csv"

ex <- tibble(
  svamp_ID = c(
    "S87_BL_run2",
    "S90_BL_run2",
    "S89_BL_run2",
    "S83_BL_run2",
    "S85_BL_run2",
    "S15"
  ),
  CST = c("IV-B", "I", "III", "III", "IV-B", "IV-C"),
  subCST = c("IV-B", "I-B", "III-B", "III-B", "IV-B", "IV-C3")
)

CST_data <- read_csv(CST) %>%
  select(svamp_ID = sampleID, CST, subCST) %>%
  bind_rows(., ex)

r <- read_stat %>% left_join(., CST_data, by = "svamp_ID") %>% arrange(svamp_ID)

# save file
write_csv(
  r,
  "../Tidy_data/Metagenomics/Read_count_info_vaginalswab.csv"
)
# read_stat <- read_csv("../Tidy_data/Metagenomics/Read_count_info_vaginalswab.csv")
# read_stat <- read_stat %>% select(-CST, -subCST)
```

S10 and S15 was exluded due to poor quality
(S10 due to too few counts and S15 due to total abundance < 60% after taxa filtering)
S86_1W stoped coming after two weeks
S65_3M missed the 3m visit

### Tissue data
#### Format count table

```{r Metagenome-tissue}
smpl_barcodes <- read_xlsx(
  paste0(input_dir, "Registrering Provrör ID CHLORHEX.xlsx"),
  skip = 1
)
mpath1 <- "../data/Metagenomics/biopsi/MGI_20241129_R1100600190018_E250043326 & E250043278.csv"
mpath2 <- "../data/Metagenomics/biopsi/BEA23P074_tissuesamples.csv"
mpath3 <- "../Tidy_data/Transcriptomics/Mapping_all_trx_samples.csv"

#############
# LOAD DATA #
#############

run1_path <- "../data/Metagenomics/biopsi/all_samples.metaphlantissue_mikantarell.csv"
run2_path <- "../data/Metagenomics/biopsi/all_samplesmetaphlantissuemikantarell.csv"
run1_data <- read_delim(run1_path, delim = ";")
run2_data <- read_delim(run2_path, delim = ";")

# run1_data and run2_data both contain 130 sample, however according to the
# sample id sheet from sequencing team "MGI_20241129_R1100600190018_E250043326 & E250043278.csv"
# there are only  49 (run 1, E250043278) and 43 (run 1, E250043326). Which means a total of 92 samples
# pluss 2 controls
# the remaning samples are probably from other projects that was sequenced at the same time

sample_info <- read_delim(mpath1, delim = ";") %>%
  filter(!is.na(adapter_id)) %>%
  mutate(Flowcell_L1_bc = paste0(flowcell_id, "_L01_", adapter_id)) %>% #92 samples
  # Flowcell_L1_bc and BEA_ID in this file has leading zeros, other files dont have this:
  mutate(
    sample_id = gsub("_0", "_", .$sample_id),
    Flowcell_L1_bc = gsub("_0", "_", .$Flowcell_L1_bc)
  )

meta2 <- read_delim(delim = ";", mpath2)
meta3 <- read_csv(mpath3)
smpl_IDs <- read_xlsx(paste0(input_dir, "svamp_ID_list.xlsx"))

#######################
# COMBINE SAMPLE INFO #
#######################

listn <- c(
  "BEA24P003_11" = "25:02",
  "BEA24P003_12" = "K33",
  "BEA24P003_22" = "29:02"
)

smplID <- sample_info %>%
  select(
    Flowcell_L1_bc,
    extraction_plate_id,
    BEA_ID = "sample_id",
    not_sequenced
  ) %>%

  # add barcode information:
  left_join(
    .,
    select(meta2, BEA_ID = "BEAID", barcode_shotgun_tissue = "Vial_number"),
    by = "BEA_ID"
  ) %>%

  # add study ID from the trx sample sheet:
  left_join(., select(meta3, BEA_ID, DS_ID), by = "BEA_ID") %>%

  # add missing barcode information:
  left_join(
    .,
    select(smpl_IDs, barcode = barcode_shotgun_tissue, DS_ID = patient_ID),
    by = c("DS_ID")
  ) %>%
  mutate(
    barcode_shotgun_tissue = ifelse(
      is.na(.$barcode_shotgun_tissue),
      .$barcode,
      .$barcode_shotgun_tissue
    )
  ) %>%
  # add control names
  mutate(
    barcode_shotgun_tissue = ifelse(
      is.na(.$barcode_shotgun_tissue),
      .$BEA_ID,
      .$barcode_shotgun_tissue
    )
  ) %>%
  select(-DS_ID, -barcode) %>%
  left_join(
    .,
    select(smpl_IDs, DS_ID = patient_ID, svamp_ID, barcode_shotgun_tissue),
    by = "barcode_shotgun_tissue"
  ) %>%

  mutate(
    DS_ID = ifelse(is.na(.$DS_ID), .$BEA_ID, .$DS_ID),
    svamp_ID = ifelse(is.na(.$svamp_ID), .$BEA_ID, .$svamp_ID)
  ) %>%
  filter(!(BEA_ID == "0")) %>%
  #dplyr::summarise(n = dplyr::n(), .by = c("svamp_ID")) |>
  #dplyr::filter(n > 1L) %>%
  unique()

write_csv(
  smplID,
  "../Tidy_data/Metagenomics/Key_all_shotgun_tissue_samples.csv"
)
# smplID <- read_csv( "../Tidy_data/Metagenomics/Key_all_shotgun_tissue_samples.csv")

########################################
# RENAME COULMS OF COUNT TISSUE TABLES #
########################################
# the count tables was delivered from the pipline with ids consisting of the flow cell id pluss the adapter id
# this needed to be replaced with the new study ID that Margaux have created
all_samples <- c(colnames(run1_data)[-1], colnames(run1_data)[-1])
our_samples <- all_samples[all_samples %in% smplID$Flowcell_L1_bc]
setdiff(smplID$Flowcell_L1_bc, our_samples)

# Ensure svamp_IDs are unique by appending a suffix to duplicates
svamp_ID_unique <- make.unique(smplID$svamp_ID, sep = "_extr")

# Create named vector: names = barcodes, values = unique svamp_IDs
ID_list <- set_names(svamp_ID_unique, smplID$Flowcell_L1_bc)


run1_data <- run1_data %>% # 48 samples
  # selects just the samples that belong to our project:
  select(clade_name, any_of(names(ID_list))) %>%
  # exhange the barcode with the svamp_ID
  rename_with(~ ID_list[.x], .cols = starts_with("E25"))

run2_data <- run2_data %>% # 48 samples
  # selects just the samples that belong to our project:
  select(clade_name, any_of(names(ID_list))) %>%
  # exhange the barcode with the svamp_ID
  rename_with(~ ID_list[.x], .cols = starts_with("E25"))

# total of 96 samples
#

###############################
# COMBINE COUNT TISSUE TABLES #
###############################
# combine count tables
# check if anny ensembl id's differ
intersect(run1_data$clade_name, run2_data$clade_name)

setdiff(run1_data$clade_name, run2_data$clade_name)
setdiff(run2_data$clade_name, run1_data$clade_name)

# lenght of combined data frame should be:
nrow(run1_data) + length(setdiff(run2_data$clade_name, run1_data$clade_name))

# Safely combine using full_join by clade_name
merged_counts <- full_join(
  run1_data,
  run2_data,
  by = "clade_name",
  suffix = c("", "_run2")
) %>%
  # replace NAs with zeros:
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>%
  select(clade_name, contains("seq"), sort(colnames(.)))

colnames(merged_counts)
write_csv(
  merged_counts,
  "../Tidy_data/Metagenomics/Metageonome_all_tissue_metaphlan.csv"
)

# tissue_counts <- read_csv("../Tidy_data/Metagenomics/Metageonome_all_tissue_metaphlan.csv")
```

#### Format count info table

```{r count-info-tissue}
##################################
# COMBINE READ COUNT TISSUE INFO #
##################################
# add read counts from processing file
# add suffix to distingush non-uniqe sample names

prepro_path1 <- "../data/Metagenomics/biopsi/preprocessing_read_counts_tissue_mikantarell.txt"
prepro_path2 <- "../data/Metagenomics/biopsi/preprocessing_read_counts_tissue_mikantarell2.txt"

prepro1 <- read_delim(prepro_path1, delim = "\t")
prepro2 <- read_delim(prepro_path2, delim = "\t")
# gr_data <- read_csv("../Tidy_data/Asigned_groups_info_all_samples.csv")

prepro <- bind_rows(prepro1, prepro2) %>%
  filter(Sample %in% names(ID_list)) %>%
  rename(Flowcell_L1_bc = Sample) %>%
  left_join(smplID, .) %>%
  # dplyr::summarise(n = dplyr::n(), .by = c("Flowcell_L1_bc")) |>
  # dplyr::filter(n > 1L) %>%
  mutate(ID = str_extract(svamp_ID, "S\\d\\d"), .before = "DS_ID") %>%
  # add "_extr1" to non uniqe svam_IDs:
  mutate(svamp_ID = ID_list[.$Flowcell_L1_bc]) %>%
  select(
    Flowcell_L1_bc,
    BEA_ID,
    barcode_shotgun_tissue,
    DS_ID,
    svamp_ID,
    everything()
  ) %>%
  mutate(
    svamp_ID = get_unique_id.fun(svamp_ID),
    barcode_shotgun_tissue = if_else(
      str_detect(barcode_shotgun_tissue, "_"),
      NA,
      barcode_shotgun_tissue
    )
  )

write_csv(prepro, "../Tidy_data/Metagenomics/Read_count_info_tissue.csv")
```

```{r merge-vagswab-and-tissue}
key_vagswab <- read_csv(
  "../Tidy_data/Metagenomics/Key_all_shotgun_vagswab_samples.csv"
)
key_tissue <- read_csv(
  "../Tidy_data/Metagenomics/Key_all_shotgun_tissue_samples.csv"
)

intersect(key_vagswab$svamp_ID, key_tissue$svamp_ID)

k <- key_vagswab %>%
  full_join(
    select(key_tissue, DS_ID, barcode_shotgun_tissue, BEA_ID, svamp_ID),
    by = c("svamp_ID", "DS_ID")
  ) %>%
  #left_join(select(key_tissue, svamp_ID, BEA_ID), by = "svamp_ID") %>%
  group_by(svamp_ID) %>%
  summarise(
    BEA_ID = paste(na.omit(unique(BEA_ID)), collapse = ", "),
    across(everything(), ~ first(.x)),
    .groups = "drop"
  ) %>%
  select(
    ID,
    DS_ID,
    svamp_ID,
    barcode_shotgun_swab = sample_ID,
    barcode_shotgun_tissue,
    BEA_ID,
    everything()
  )


write_csv(k, "../Tidy_data/Metagenomics/Readcounts.csv")
```

### Functional metagenome data
```{r functional-annotation}
#############
# LOAD DATA #
#############
# Biopsi samples
tissue_run1 <- read_tsv(paste0(
  input_dir,
  "Metagenomics/Functional profiling/",
  "all_samples.humann_pathabundance-5_tissue.txt"
))
tissue_run2 <- read_tsv(paste0(
  input_dir,
  "Metagenomics/Functional profiling/",
  "all_samples.humann_pathabundance-1_tissue.txt"
))

# vagswab run 1
vagswab_run1 <- read_tsv(paste0(
  input_dir,
  "Metagenomics/Functional profiling/",
  "all_samples.humann_pathabundance-4_run1.txt"
))
# vagswab run 2 additional 19 samples
vagswab_run2 <- read_tsv(paste0(
  input_dir,
  "Metagenomics/Functional profiling/",
  "all_samples.humann_pathabundance-4_run2.txt"
))

######################
# MERGE VAGSWAB DATA #
######################
# Safely combine using full_join by Pathway
tissue_data <- full_join(
  tissue_run1,
  tissue_run2,
  by = "# Pathway",
  suffix = c("", "_run2")
) %>%
  # replace NAs with zeros:
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>%
  select(`# Pathway`, contains("Ctrl"), sort(colnames(.)))

colnames(tissue_data)

# Safely combine using full_join by Pathway
vagswab_data <- full_join(
  vagswab_run1,
  vagswab_run2,
  by = "# Pathway",
  suffix = c("", "_run2")
) %>%
  # replace NAs with zeros:
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) %>%
  select(`# Pathway`, contains("ex"), contains("seq"), sort(colnames(.)))

colnames(vagswab_data)
```

```{r}
################
# GET SVAMP ID #
################
tissue_count_info <- read_csv(
  "../Results/Tissue_QC/Read_count_info_tissue_no_spikein.csv"
) %>%
  # remove samples that have zero bacterial reads after spike-in
  filter(bac_counts_no_spikein > 2000)
vagswab_count_info <- read_csv(
  "../Tidy_data/Metagenomics/Read_count_info_vaginalswab.csv"
)

ID_tissue <- set_names(
  tissue_count_info$svamp_ID,
  tissue_count_info$Flowcell_L1_bc
)
ID_vagswab <- set_names(vagswab_count_info$svamp_ID, vagswab_count_info$Sample)

setdiff(gsub("_Abundance", "", colnames(tissue_data)), names(ID_tissue))
intersect(gsub("_Abundance", "", colnames(tissue_data)), names(ID_tissue))

# rename columns with svamp ID
# tissue
# NB! we have to think through if it is nessecary to do some sort of filtering of this data before use
# I dont know if this is with or withouth the spike-in
# also only 47 samples match, I think we are missing a file with the other samples
tissue_data <- tissue_data %>%
  rename_with(
    ~ str_replace(.x, "_Abundance", ""),
    .cols = starts_with("E25")
  ) %>%
  rename_with(
    ~ ID_tissue[.x],
    .cols = names(.)[names(.) %in% names(ID_tissue)]
  ) %>%
  # remove samples that are not ours or had zero counts after spike-in removal
  select(-starts_with("E25")) %>%
  # order columns
  select(`# Pathway`, contains("Ctrl"), sort(colnames(.)))


vagswab_data <- vagswab_data %>%
  rename_with(
    ~ str_replace(.x, "_Abundance", ""),
    .cols = starts_with("242")
  ) %>%
  rename_with(~ ID_vagswab[.x], .cols = starts_with("242"))

############################
# SAVE PATH ABUNDANCE DATA #
############################
write_csv(
  tissue_data,
  "../Tidy_data/Metagenomics/Functional_annotation_all_tissue.csv"
)
write_csv(
  vagswab_data,
  "../Tidy_data/Metagenomics/Functional_annotation_all_vagswab.csv"
)
```


## Format Metabolomics data
```{r Metabolom}
#############
# LOAD DATA #
#############
mbol_path <- "../data/Metabolomics/UTF-8SMC_P0800_GCMS_Data.xlsx"
gr_id_path <- "../Tidy_data/Asigned_groups_info_all_samples.csv"

mbol_data <- read_xlsx(mbol_path)
smpl_IDs <- read_xlsx(paste0(input_dir, "svamp_ID_list.xlsx")) %>%
  select(svamp_ID, barcode_metabolomics) %>%
  filter(!(is.na(.$barcode_metabolomics)))
gr_id <- read_csv(gr_id_path)

left_join(
  .,
  select(smpl_IDs, svamp_ID, DS_ID = "patient_ID"),
  by = "DS_ID"
)

mbol_data <- mbol_data %>%
  select(
    `Compound Name`,
    #DS_ID = `Subject ID`,
    barcode_metabolomics = "customer sample Identifier",
    `Injection Standard MS`:`N-Acetylneuraminic Acid`
  ) %>%
  left_join(., smpl_IDs, by = "barcode_metabolomics") %>%
  relocate(svamp_ID, .before = "barcode_metabolomics") %>%
  mutate(
    svamp_ID = ifelse(
      grepl("^QC\\d", .$`Compound Name`),
      .$`Compound Name`,
      .$svamp_ID
    )
  )

write_csv(
  mbol_data,
  "../Tidy_data/Metabolomics/Metabolom_all_samples.csv"
)
```

## Cleaning up data
The metabolomic data was provided in a unformatted table, so compiling the data into appropriate format is required. Differences in sample number, metabolite nomenclature, etc.
```{r}
#############
# LOAD DATA #
#############
#---  Load sample info ----
smpl_IDs <- read_xlsx("../data/svamp_ID_list.xlsx") %>%
  select(svamp_ID, barcode_metabolomics)

#---  Load raw file from metabolomics sequencing facility ----
mbol_path <- "../data/Metabolomics/UTF-8SMC_P0800_GCMS_Data.xlsx"
RAW <- read_xlsx(mbol_path)


##############
# CLEAN DATA #
##############
# this code is based on Paulos analysis of the metabolomics data
# View first few column names and first column values
head(colnames(RAW), 30)
head(RAW[[1]], 30)

#--- 1️⃣ Identify structure of the table ----
# Find where annotation columns end (contains "order" keyword)
col_boundary <- which(str_detect(
  colnames(RAW),
  regex("order", ignore_case = TRUE)
))
# Find where sample data begins (first row that starts with "1" in first column)
row_boundary <- which(RAW[[1]] == "1")[1] - 1

#--- 2️⃣ Split into three logical components ----
# Sample metadata (left block, below header)
metadata <- RAW %>%
  slice((row_boundary + 1):n()) %>%
  select(all_of(1:col_boundary))

# Compound annotations (top block, right side)
annotations_raw <- RAW %>%
  slice(1:row_boundary) %>%
  select(1, all_of((col_boundary + 1):ncol(RAW)))
#select(1, all_of((col_boundary + 1):ncol(RAW)))

# Measurement matrix (sample × metabolite intensities)
matrix_data <- RAW %>%
  slice((row_boundary + 1):n()) %>%
  select(
    `Compound Name`,
    "barcode_metabolomics" = `customer sample Identifier`,
    (col_boundary + 1):ncol(RAW)
  ) %>%
  left_join(., smpl_IDs, by = "barcode_metabolomics") %>%
  relocate(svamp_ID, .before = "barcode_metabolomics") %>%
  mutate(
    svamp_ID = ifelse(
      grepl("^QC\\d", .$`Compound Name`),
      .$`Compound Name`,
      .$svamp_ID
    )
  ) %>%
  select(-`Compound Name`, -barcode_metabolomics) %>%
  pivot_longer(cols = -svamp_ID, names_to = "Compund") %>%
  pivot_wider(names_from = svamp_ID) %>%
  select(Compund, starts_with("QC"), sort(colnames(.)))


#--- 3️⃣ Build tidy annotation table ----
# Extract selected annotation rows and pivot them into a tidy format
annotation_table <- annotations_raw %>%
  #rename(row_label=`GCMS-NA@NA`) %>%
  slice(., c(2:6)) %>% # rows with HMDB, KEGG, PubChem, Mz, Rt
  pivot_longer(-1, names_to = "names", values_to = "value") %>%
  pivot_wider(names_from = "Compound Name", values_from = value) %>%
  # Create uniqe compound IDs by using M/z and Rt
  mutate(
    Compound_ID = paste0(
      "GCMS-",
      round(as.numeric(`M/z (unique)`)),
      "@",
      round(as.numeric(`RI (resolved)`))
    )
  )

#--- 4️⃣ Remove quality control compunds ----
annotation_table <- annotation_table %>%
  filter(!grepl("Standard", .$names))

# ✅ Results:
# metadata         → sample metadata (e.g. sample ID, run order)
# matrix_data      → numeric matrix (samples × compounds)
# annotation_table → clean annotation metadata per compound

#############
# SAVE DATA #
#############
write_csv(matrix_data, "../Tidy_data/Metabolomics/Metabolom_matrix.csv")
write_csv(metadata, "../Tidy_data/Metabolomics/metabolomics_metadata.csv")
write_csv(
  annotation_table,
  "../Tidy_data/Metabolomics/metabolomics_annotation.csv"
)
```