---
title: "Figure 3"
date: today
date-format: "D MMM YYYY"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    highlight-style: github
    code-tools:
      source: true
      toggle: true
---
## 1. Libraries and global settings

```{r libraries}
#| code-fold: true
#| output: false
#| warning: false
#| message: false
library(MOFA2)
library(tidyverse)
library(jsonlite)
library(cowplot)
library(patchwork)
library(ggExtra)
# library(openxlsx)
# library(ggrepel)
library(RColorBrewer)
library(ggh4x)

library(ComplexHeatmap)
library(circlize)
library(paletteer)

# setwd("/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/manuscript")

select <- dplyr::select
map <- purrr::map
rename <- dplyr::rename
```


```{r load-data}
#| output: false
#| warning: false
#| message: false
# meta data
m <- read_csv("../Results/Sample_omic_table.csv")
comb_meta_path <- "../Results/Combined_gr_meta_data.csv"
meta <- read_csv(comb_meta_path) %>%
  #rename(sample = "svamp_ID") %>%
  mutate(across(2:27, ~ factor(.x)))

# omic data
MG_path <- "../Results/MixOmic/Metageonome_vagswab_CLR_transformed_1e-6.RDS"
FM_path <- "../Tidy_data/Metagenomics/Functional/Functional_annotation_relative_abundance_filtered.csv"
MB_path <- "../Results/MixOmic/Metabolom_Normalized.csv"
MB_annot_path <- "../Tidy_data/Metabolomics/metabolomics_annotation.csv"

MG_matrix <- readRDS(MG_path)
FM_data <- read_csv(FM_path)
MB_matrix <- read_csv(MB_path)

annotation_MB <- read_csv(MB_annot_path)

# omic colour pallet
MG_hirarchy <- read_csv("../Results/Metagenomics/Metagenome_col_pal.csv")
FM_colours <- read_csv("../Results/Metagenomics/Functional_col_pal.csv")
MB_colours <- read_csv("../Results/Metagenomics/Metabolome_col_pal.csv")
```

```{r pallets}
#| code-fold: true
# colour pallet
gene_col <- c(
  "#B71E42FF",
  "#DE478E",
  "#BC72F0FF",
  "#795FAFFF",
  "#D53DD0",
  "#EA6312",
  "#FFAE34FF",
  "#EF6F6AFF",
  "#8CC2CAFF",
  "#6388B4FF"
)
pal <- list(
  fct1_clus = c("#33A02C", "#B2DF8A", "#FF7F00", "#FDBF6F"),
  RVVC_pos = c("#A6CEE3", "#1F78B4", "#e490c4ff", "#c12d88ff"),
  "Clinical score (0-5)" = brewer.pal(6, name = "Reds"),
  CST = c(
    "I" = "#148F77",
    "II" = "#6d9f06",
    "III" = "#A3E4D7",
    "IV-B" = "#E794C1",
    "IV-C" = "#b5cb8c",
    "V" = "#43BA8F"
  )
)

# default ggplot2 pallet:
gg_color_hue <- function(n) {
  hues <- seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
gg_color_hue(4)

tax_pallet <- c(brewer.pal(12, "Set3"), brewer.pal(8, "Dark2"))
other <- c(
  "#E3E6AD",
  "#F8D0A4",
  "#c4ce96",
  "#9aacce",
  "#e1caff",
  "#abc5bf",
  "#ffffd4",
  "#c0a2c1",
  "#c8ffd5",
  "#c8ffd5",
  "#afb7ee",
  "#ffc8d9",
  "#ffc8d9",
  "#e7b993",
  "#c8ffd5",
  "#c4cea9",
  "#a1b37d",
  "#a6cca7",
  "#d1b9ee",
  "#88c29c",
  "#fdcc8a",
  "#91c6f7",
  "#f5f8bd",
  "#8db1c5",
  "#fab0aa",
  "#7cb6b6",
  "#96f3eb",
  "#6ececc"
)
```

## 2. Load trained MOFA model
We load the model with the highest elbo score

```{r load-MOFA-model}
#| eval: true
#| include: true
#| warning: false
#| message: false
#| tbl-colwidths: [60,40]
elbo_table <- read_csv("../Results/MOFA/elbo_table_noS84_1e-6_57MG.csv")
knitr::kable(elbo_table)
best_seed <- elbo_table$seed[which.max(abs(elbo_table$elbo))]

MOFAmodel <- load_model(
  paste0("../Results/MOFA/MOFA_allBL_noS84_1e-6_57MG_seed_", best_seed, ".hdf5")
)

# add meta data to model
samples_metadata(MOFAmodel) <- samples_metadata(MOFAmodel)[, c(
  "sample",
  "group"
)] %>%
  left_join(meta, by = c("sample"))
```

## 3. Plot Factor 1 cluster
```{r factor-with-distribution-function}
#| code-fold: true
# ----------------------
# Plot Factor Density Function
# ----------------------
plot_mofa_factor_vs_variable <- function(
  df,
  factor = 1,
  y_var,
  palette = pal[[y_var]],
  jitter_width = 0.1,
  add_marginal = TRUE,
  marginal_type = "density"
) {
  # jitter object
  jitterer <- position_jitter(width = jitter_width, seed = 123)

  df_ <- df %>% mutate(!!sym(y_var) := factor(!!sym(y_var)))

  # Main ggplot
  p <- ggplot(
    df,
    aes(
      x = .data[[paste0("Factor", factor)]],
      y = .data[[y_var]],
      colour = .data[[y_var]]
    )
  ) +
    geom_point(
      size = 1,
      alpha = 0.8,
      show.legend = FALSE,
      position = jitterer
    ) +
    geom_point(
      aes(colour = NULL),
      pch = 21,
      size = 1,
      alpha = 0.4,
      show.legend = FALSE,
      colour = "gray",
      position = jitterer
    ) +
    geom_boxplot(
      aes(colour = .data[[y_var]]),
      fill = "transparent",
      box.linewidth = 0.2,
      whisker.linewidth = 0.2,
      show.legend = FALSE,
      outlier.shape = NA,
      median.color = "black"
    ) +
    coord_cartesian(expand = TRUE) +
    theme_minimal() +
    theme(
      panel.grid = element_line(linewidth = 0.2),
      text = element_text(size = 7),
      plot.margin = unit(c(0, 0, 0, -0), "cm"),
      axis.text.y.left = element_blank(),
      axis.title.x = element_text(vjust = 1),
      axis.title.y.left = element_blank()
    )

  # Add palette if provided
  if (!is.null(palette)) {
    p <- p + scale_colour_manual(values = palette)
  }

  # Add marginal plot if requested
  if (add_marginal) {
    p <- ggMarginal(
      p,
      type = marginal_type,
      groupColour = TRUE,
      groupFill = TRUE,
      trim = FALSE,
      linewidth = 0.2,
      margins = "x",
      size = 1
    )
  }
  #p <- p + theme( plot.margin = unit(c(0, 0, -1, -0), "cm"))
  return(p)
}
```

```{r factor-with-distribution-plots}
#| fig-width: 3
#| fig-height: 2
# ----------------------
# Prepp df
# ----------------------
# Extract factors, reshape wide, join metadata
df <- get_factors(MOFAmodel, factors = c(1:4), as.data.frame = TRUE) %>%
  as_tibble() %>%
  select(-group) %>%
  #mutate(fct1_clus = as.character(clusters$cluster[.$sample])) %>%
  pivot_wider(names_from = "factor") %>%
  left_join(samples_metadata(MOFAmodel), by = "sample") %>%
  mutate(across(fct1_clus:fct4_clus, ~ factor(.x)))

# ----------------------
# Plot
# ----------------------
# quartz(width = 3, height = 2, dpi = 150)
A1 <- plot_mofa_factor_vs_variable(
  df,
  y_var = "fct1_clus"
)
A1
```

## 4. Plot top factor 1 features by view
```{r modified-MOFA-heatmap-function}
#| code-fold: true
# Modified function: plot_data_heatmap2() with sample_order argument
# It is identical to the original except for 13 added lines.
plot_data_heatmap2 <- function(
  object,
  factor,
  view = 1,
  groups = "all",
  features = 50,
  legend = TRUE,
  max_val = NULL,
  annotation_features = NULL,
  annotation_samples = NULL,
  labels_row = NULL,
  transpose = FALSE,
  imputed = FALSE,
  denoise = FALSE,
  max.value = NULL,
  min.value = NULL,
  sample_order = NULL, # ✅ NEW ARGUMENT
  ...
) {
  if (!is(object, "MOFA")) {
    stop("'object' has to be an instance of MOFA")
  }

  stopifnot(length(factor) == 1)
  stopifnot(length(view) == 1)

  # ✅ INTERNAL HELPERS MUST USE triple-colon
  groups <- MOFA2:::.check_and_get_groups(object, groups)
  factor <- MOFA2:::.check_and_get_factors(object, factor)
  view <- MOFA2:::.check_and_get_views(object, view)

  # Extract model matrices
  W <- do.call(
    rbind,
    get_weights(object, views = view, factors = factor, as.data.frame = FALSE)
  )
  Z_list <- lapply(get_factors(object)[groups], function(z) {
    as.matrix(z[, factor])
  })
  Z <- do.call(rbind, Z_list)[, 1]
  Z <- Z[!is.na(Z)]

  # Choose raw, imputed, or denoised data
  if (isTRUE(denoise)) {
    data <- predict(object, views = view, groups = groups)[[1]]
  } else if (isTRUE(imputed)) {
    data <- get_imputed_data(object, view, groups)[[1]]
  } else {
    data <- get_data(object, views = view, groups = groups)[[1]]
  }
  if (is(data, "list")) {
    data <- do.call(cbind, data)
  }

  # ----------------------
  # FEATURE SELECTION
  # ----------------------
  if (is.numeric(features)) {
    if (length(features) == 1) {
      features <- rownames(W)[tail(order(abs(W)), n = features)]
    } else {
      features <- rownames(W)[order(-abs(W))[features]]
    }
    features <- names(W[features, ])[order(W[features, ])]
  } else if (is.character(features)) {
    stopifnot(all(features %in% features_names(object)[[view]]))
  } else {
    stop("features must be numeric or character vector")
  }

  data <- data[features, ]
  data <- data[, names(Z), drop = FALSE]
  data <- data[, apply(data, 2, function(x) !all(is.na(x))), drop = FALSE]

  # ----------------------
  # ✅ CUSTOM SAMPLE ORDER
  # ----------------------
  if (is.null(sample_order)) {
    # default MOFA2 ordering (by factor value)
    order_samples <- names(sort(Z, decreasing = TRUE))
    order_samples <- order_samples[order_samples %in% colnames(data)]
  } else {
    # user‑specified ordering
    sample_order <- sample_order[sample_order %in% colnames(data)]
    if (!all(sample_order %in% colnames(data))) {
      stop("sample_order contains sample names not found in data")
    }
    order_samples <- sample_order
  }

  data <- data[, order_samples, drop = FALSE]

  # ----------------------
  # ANNOTATION HANDLING
  # ----------------------
  if (!is.null(annotation_samples)) {
    if (is.data.frame(annotation_samples)) {
      if (any(!order_samples %in% rownames(annotation_samples))) {
        stop("annotation_samples rownames must match sample names")
      }

      annotation_samples <- annotation_samples[order_samples, , drop = FALSE]
    } else if (is.character(annotation_samples)) {
      tmp <- object@samples_metadata
      rownames(tmp) <- tmp$sample
      tmp$sample <- NULL
      tmp <- tmp[order_samples, , drop = FALSE]
      annotation_samples <- tmp[, annotation_samples, drop = FALSE]
      rownames(annotation_samples) <- order_samples
    } else {
      stop("annotation_samples must be a data.frame or character vector")
    }

    # logical/character must become factor
    to_factor <- sapply(annotation_samples, function(x) {
      is.logical(x) || is.character(x)
    })
    if (any(to_factor)) {
      annotation_samples[, which(to_factor)] <-
        lapply(annotation_samples[, which(to_factor), drop = FALSE], as.factor)
    }
  }

  if (transpose) {
    data <- t(data)
    tmp_ann <- annotation_samples
    annotation_samples <- annotation_features
    annotation_features <- tmp_ann
  }

  # cap values
  if (!is.null(max.value)) {
    data[data >= max.value] <- max.value
  }
  if (!is.null(min.value)) {
    data[data <= min.value] <- min.value
  }
  data_ <<- data
  if (!is.null(labels_row)) {
    row_names <- str_replace(
      rownames(data),
      "^(superpathway of [^,]+),.*",
      "\\1"
    )
  } else {
    row_names <- rownames(data)
  }

  # ----------------------
  # ✅ FINAL HEATMAP CALL
  # ----------------------
  #RdBu <- brewer.pal(11, "RdBu")
  # RdYlBu <- rev(brewer.pal(7, "RdYlBu")) # NB! 7 vs 11 makes a huge dfference, 7 gives smoother transition of colour
  col2 <- rev(brewer.pal(7, "RdYlBu"))
  pheatmap::pheatmap(
    data,
    treeheight_row = 0,
    cellwidth = 1,
    fontsize = 8,
    show_colnames = F,
    border_color = NA,
    labels_row = row_names,
    legend = legend,
    annotation_legend = legend,
    color = colorRampPalette(col2)(100),
    breaks = if (!is.null(max_val)) {
      seq(-max_val, max_val, length.out = 101)
    } else {
      NULL
    },
    annotation_row = annotation_features,
    annotation_col = annotation_samples,
    ...
  )
}

# old alternativer version:
# set sample order (columns)
# NB! can not be used with denoise = TRUE!! (because it changes the Z values used to compute the denoised values used in the heatmap )
set_sample_order <- function(MOFAmodel, factor, order) {
  Z <- MOFAmodel@expectations$Z$single_group[, factor]

  # assign artificial values based on your desired order
  Z[order] <- seq_along(order)

  MOFAmodel@expectations$Z$single_group[, factor] <- Z
  MOFAmodel
}

# MOFAmodel2 <- set_sample_order(MOFAmodel, factor = 1, order = sample_order)
```

```{r transcriptomics_heatmap}
#| fig-width: 10
#| output: false
#| code-fold: true
meta_headers <- colnames(samples_metadata(MOFAmodel))
gr_df <- samples_metadata(MOFAmodel)[, c(
  "pos",
  "fct1_clus",
  "Clinical score (0-5)",
  "RVVC_pos",
  "CST",
  "BV",
  "Wet smear: clue cells (y/n)",
  "metab_louvain"
)] %>% #"sample",
  mutate_at(., 1:ncol(.), factor)
rownames(gr_df) <- samples_metadata(MOFAmodel)$sample

# determine column order when cluster_cols=F
sample_order <- meta %>%
  filter(sample %in% rownames(gr_df)) %>%
  # mutate(fct1_clus = factor(.$fct1_clus, levels = c("3", "4", "1", "2"))) %>%
  arrange(RVVC_pos) %>%
  #arrange(RVVC_pos, `Clinical_score_(0-5)`) %>%
  arrange(fct1_clus, desc(`Clinical_score_(0-5)`)) %>%
  pull(sample)

gr_df <- gr_df[sample_order, ]

ann_colors <- map(colnames(gr_df), function(val) {
  # Levels for this column
  levs <- levels(gr_df[[val]])

  # CASE 1: column exists in pal → use provided palette
  if (val %in% names(pal)) {
    colors <- pal[[val]]
    return(set_names(colors, levs))
  }
  # CASE 2: column not in pal → generate paired palette
  colors <- gg_color_hue(length(levs))
  set_names(colors, levs)
  # colors <- brewer.pal(length(levs), "Paired")
  # if (length(levs) == length(colors)) {
  #   set_names(colors, levs)
  # } else {
  #   set_names(colors[1:length(levs)], levs)
  # }
})
# Name the list by column names
ann_colors <- set_names(ann_colors, colnames(gr_df))

# quartz(width = 6, height = 4, dpi = 150)
TR <- plot_data_heatmap2(
  MOFAmodel,
  view = "Transcriptomics",
  factor = 1,
  features = 20,
  legend = FALSE,
  max_val = 3.5,
  #sample_order = sample_order,
  cluster_cols = F,
  annotation_samples = gr_df[, 2:5], #gr_df[,-1], "Clin_gr"
  denoise = FALSE,
  scale = "row",
  annotation_colors = ann_colors
)

# quartz(width = 7, height = 4, dpi = 50)
MB <- plot_data_heatmap2(
  MOFAmodel,
  view = "Metabolomics",
  factor = 1,
  features = 20,
  legend = FALSE,
  max_val = 3.5,
  #sample_order = NULL,
  cluster_cols = F,
  annotation_samples = gr_df[, c(2:5)], #gr_df[,-1], "Clin_gr"
  denoise = F,
  scale = "row",
  annotation_colors = ann_colors
)

FM <- plot_data_heatmap2(
  MOFAmodel,
  view = "Functional",
  factor = 1,
  features = 20,
  labels_row = "shorten",
  max_val = 3.5,
  #sample_order = NULL,
  cluster_cols = F,
  annotation_samples = gr_df[, c(2:5)], #gr_df[,-1], "Clin_gr"
  denoise = F,
  scale = "row",
  annotation_colors = ann_colors
)
```

```{r combine-heatmaps-v}
#| fig-width: 11.5
#| fig.height: 3
# quartz(width = 11.5, height = 3, dpi = 100)
B <- plot_grid(
  TR$gtable,
  MB$gtable,
  FM$gtable,
  nrow = 1,
  rel_widths = c(0.75, 0.9, 2)
)
B
```

```{r combine-heatmaps-h}
#| fig-width: 6.5
#| fig.height: 10
# quartz(width = 6.5, height = 10, dpi = 90)
plot_grid(
  TR$gtable, #NULL,
  MB$gtable, #NULL,
  FM$gtable,
  #axis = "l",
  align = c("v"),
  ncol = 1
  #rel_widths = c(1, 1.2, 2)
)
```

plot_data_heatmap has an interesting argument to “beautify” the heatmap: `denoise = TRUE`. Instead of plotting the (noisy) input data, we can plot the data reconstructed by the model, where noise has been removed:

## bar plots

```{r facet-bar-function}
#| code-fold: true
# ============================================================
# FUNCTION: gr.fun()
# ============================================================
# Description:
# Assigns dominant microbial groups to samples based on
# highest relative abundance taxa. Used for downstream
# grouping and visualization of metagenomic profiles.
#
# Workflow:
# 1. Group data by sample
# 2. Identify taxa with highest abundance
# 3. Extract dominant genus/group labels
# 4. Recode labels into simplified categories
#
# Arguments:
# df: Data frame containing sample-level abundance data
#
# Returns:
# Data frame with added dominant group annotation column
#
# Dependencies:
# tidyverse, stringr
# ============================================================
gr.fun <- function(df) {
  groups <- df %>%
    # df_long %>% filter(sample == "S29") %>%
    # df_long %>%
    group_by(sample) %>%
    arrange(desc(abundance)) %>%
    mutate(gr = species[which.max(abundance)]) %>%
    mutate(gr = str_extract(first(gr), "^\\w+")) %>%
    mutate(
      gr = case_when(
        str_detect(gr, "Lactobacillus|Gardnerella") ~ gr,
        TRUE ~ "other"
      )
    ) %>%
    mutate(
      gr = ifelse(
        str_detect(first(species), "iners"),
        "iners",
        gr
      )
    ) %>%
    mutate(
      gr = ifelse(
        gr == "iners" &
          str_detect(nth(species, n = 2), "crispatus"),
        "Lactobacillus",
        gr
      )
    ) %>%

    ungroup() %>%
    select(sample, gr) %>%
    unique() %>%
    deframe()
  return(groups)
}
# ============================================================
# FUNCTION: factor.fun()
# ============================================================
# Description:
# Categorizes microbial taxa into predefined community
# groups and prepares formatted abundance summaries for
# downstream plotting and interpretation.
#
# Workflow:
# 1. Define ordered microbial group levels
# 2. Match taxa to group categories
# 3. Aggregate and summarize abundance patterns
# 4. Prepare output for plotting functions
#
# Arguments:
# df: Data frame containing abundance and taxonomy data
#
# Returns:
# Formatted data frame with group annotations
#
# Dependencies:
# tidyverse, forcats, stringr
# ============================================================
# order samples by percentage of specified taxa based on assigned microbiome group
# NB! I found that I got a cleaner looking graph by using gardnerella also for L4 (typ = identity)
factor.fun <- function(df) {
  # title1 ----
  l <- c(
    "Lactobacillus" = "crispatus",
    "iners" = "iners",
    "Gardnerella" = "Gardnerella",
    "other" = "Gardnerella"
  )
  imap(
    l,
    ~ filter(df, gr == .y & str_detect(species, .x)) %>%
      arrange(., desc(abundance)) %>%
      select(sample, gr) %>%
      unique() %>%
      pull(., "sample")
  ) %>%
    unlist()
}
#| code-fold: true
# ============================================================
# FUNCTION: legend_height_cm()
# ============================================================
# Description:
# Calculates the total height of a ggplot legend object
# in centimeters. Used for dynamically scaling multi-panel
# figure layouts.
#
# Workflow:
# 1. Extract legend height units
# 2. Sum total legend height
# 3. Convert height to centimeters
#
# Arguments:
# leg :
#   ggplot legend grob object
#
# Returns:
# Numeric legend height in centimeters
#
# Dependencies:
# grid
# ============================================================
legend_height_cm <- function(leg) {
  grid::convertUnit(
    sum(leg$heights),
    unitTo = "cm",
    valueOnly = TRUE
  )
}
# ============================================================
# FUNCTION: get_legend.fun()
# ============================================================
# Description:
# Generates customized legends for abundance-based plots.
# Used to create compact and publication-ready annotation
# legends with controlled spacing and formatting.
#
# Workflow:
# 1. Create temporary ggplot object
# 2. Apply custom fill scales and formatting
# 3. Extract legend object
# 4. Return formatted legend grob
#
# Arguments:
# df :
#   Data frame containing plotting annotations
#
# title :
#   Legend title
#
# id :
#   Identifier column used for plotting
#
# colours :
#   Named vector of colors
#
# txt_size :
#   Legend text size
#   Default = 10
#
# Returns:
# ggplot legend grob object
#
# Dependencies:
# ggplot2, cowplot/grid utilities
# ============================================================
# https://stackoverflow.com/questions/73711918/how-to-remove-majority-of-blank-space-from-get-legend
get_legend.fun <- function(df, title, feature, id, colours, txt_size = 10) {
  p <- ggplot() +
    geom_col(
      data = df,
      aes(fill = .data[[feature]], y = abundance, x = .data[[id]])
    ) +
    scale_fill_manual(title, values = colours) +
    guides(
      fill = guide_legend(
        override.aes = list(size = 2),
        keyheight = .5,
        keywidth = .5
      )
    ) +
    theme(
      text = element_text(size = txt_size),
      legend.background = element_blank(),
      legend.title = element_text(margin = margin(b = -0.1))
    )

  cowplot::get_legend(p)
}
#grid::grid.draw(c$legend[[1]])
# ============================================================
# FUNCTION: Facet_bar.fun()
# ============================================================
# Description:
# Creates a faceted stacked bar plot using ggplot2, where
# samples are displayed on the x-axis and abundance values
# are stacked according to a specified grouping variable.
# NB! position = "fill", scales each bar to have the same height (100%)
#
# The plot is further divided into facets based on a user-
# defined variable, allowing comparison of compositional
# patterns across experimental groups or conditions.
#
# Workflow:
# 1. Map x-axis, y-axis, and fill aesthetics from input data
# 2. Generate stacked bar plot of abundance values
# 3. Apply minimal theme for clean visualization
# 4. Remove plot expansion to avoid extra spacing
# 5. Apply manual color scale for fill categories
# 6. Customize theme elements (axes, margins, legend)
# 7. Rotate and format x-axis labels for readability
# 8. Remove axis titles for cleaner faceted output
# 9. Facet plot by a user-defined variable with free x-scale
#    and free x-space for independent panel scaling
#
# Arguments:
# df :
#   Data frame containing plotting data.
#
#   Required columns:
#     - abundance : numeric values representing feature abundance
#     - grouping variable specified in `fill`
#     - grouping variable specified in `x_axis`
#
# facet_var :
#   Variable used to define faceting structure.
#   Supports tidy evaluation ({{ }}), typically categorical.
#
# fill :
#   Character string specifying the column used for fill
#   aesthetics (default: "annot").
#
# colours :
#   Named vector of colors used in scale_fill_manual().
#   Names must match levels of the fill variable.
#
# x_axis :
#   Character string specifying the x-axis variable
#   (default: "sample").
#
# Returns:
#   A ggplot object representing a faceted stacked bar plot.
#
# Plot characteristics:
#   - Stacked bar plot (position = "stack")
#   - Faceted by `facet_var` with free x-scales
#   - Minimal theme with customized axis and strip styling
#   - Legend removed for cleaner multi-panel visualization
#
# Side effects:
#   None
#
# Dependencies:
#   ggplot2
#   rlang (for {{ }} tidy evaluation)
#   grid (for unit())
# ============================================================

Facet_bar.fun <- function(
  df,
  facet_var,
  fill = "pathway",
  colours,
  x_axis = "sample",
  annot_col = NULL, #"Clinical score (0-5)",
  strip_colours = NULL,
  legend = "none"
) {
  # -------------------------------
  # Apply strip colours if provided
  # -------------------------------
  # Only colour strips in x-direction
  if (!is.null(strip_colours)) {
    strip <- facet_grid2(
      cols = vars(.data[[facet_var]]),
      scales = "free_x",
      space = "free_x",
      strip = strip_themed(background_x = elem_list_rect(fill = strip_colours))
    )
  } else {
    strip <- facet_grid(
      cols = vars(.data[[facet_var]]),
      scales = "free_x",
      space = "free_x"
    )
  }

  # -------------------------------
  #  Annotation above barplot
  # -------------------------------
  library(ggnewscale)
  if (!is.null(annot_col)) {
    annot <- list(
      new_scale_fill(),
      geom_tile(
        aes(x = sample, y = 1.001, fill = .data[[annot_col]]),
        inherit.aes = FALSE,
        show.legend = F,
        height = 0.03, #width = 0.9
      ),
      scale_fill_manual(values = pal[[annot_col]], name = NULL),
      guides(
        fill = guide_legend(
          override.aes = list(size = 2),
          ncol = 1,
          byrow = T,
          keyheight = .4,
          keywidth = .6
        )
      )
    )
  } else {
    annot <- NULL
  }

  p <- ggplot(
    df,
    aes(fill = .data[[fill]], y = abundance, x = .data[[x_axis]])
  ) +
    geom_bar(position = "fill", stat = "identity") +
    theme_minimal(base_size = 8) +
    coord_cartesian(expand = FALSE) +
    scale_fill_manual(values = colours, name = NULL) +
    scale_y_continuous(labels = scales::label_percent(scale = 100)) +
    guides(
      fill = guide_legend(
        override.aes = list(size = 2),
        #ncol = 4, byrow = FALSE,
        keyheight = .4,
        keywidth = .6
      )
    ) +
    theme(
      strip.background = element_rect(colour = "gray", fill = NA),
      panel.border = element_rect(colour = "gray", fill = NA),
      plot.margin = unit(c(0.1, 0, 0.2, 0), "cm"),
      axis.text.x = element_blank(),
      #strip.text = element_text(lineheight = 0.5),
      strip.text.x = element_text(margin = margin(0.1, 0, 0.1, 0, "cm")),
      # axis.text.x = element_text(
      #   angle = 90,
      #   vjust = 0.5,
      #   hjust = 1,
      #   size = 8,
      #   margin = margin(t = -5, b = 3)
      # ),
      axis.title.y = element_blank(),
      axis.title.x = element_blank(),
      legend.spacing.y = unit(.1, 'cm'),
      legend.box.spacing = unit(.1, 'cm'),
      legend.key.spacing.y = unit(.1, 'cm'),
      legend.text = element_text(
        size = 8,
        margin = margin(r = 0.3, l = 0.3, unit = "pt"),
      ),
      legend.position = legend
    ) +
    strip +
    annot
}
```

```{r weights}
weights <- get_weights(MOFAmodel, factors = 1, as.data.frame = T) %>%
  as_tibble() %>%
  mutate(feature = as.vector(.$feature)) %>%
  group_by(view) %>%
  arrange(desc(abs(value))) %>%
  split(~view)

weights$Metagenomics <- weights$Metagenomics %>%
  arrange(desc(abs(value))) %>%
  mutate(feature = str_replace(feature, "_", " "))
```

## 5. Prepp indivdual omics for barplot

### prepp MB
```{r prepp-MB-data}
# select samples
MB_data <- MB_matrix %>%
  select(names = "svamp_ID", any_of(m$sample[m$at_least_two & m$all_BL]))

# add hirarchy
chem_hierarchy <- read_csv("../Results/Metagenomics/MB_chem_hierarchy.csv")
MB_data <- MB_data %>%
  left_join(select(annotation_MB, names, HMDB), .) %>%
  left_join(chem_hierarchy, .) %>%
  rename(
    MB_Class1 = "ClassyFire_super_class",
    MB_Class2 = "ClassyFire_class",
    MB_Class3 = "ClassyFire_sub_class"
  )
# col pal
MB_colours <- set_names(MB_colours$col, MB_colours$Pathway)
```

### prepp MG
```{r prepp-MG-data}
# select samples
MG_data <- MG_matrix$data$strain %>%
  as_tibble(., rownames = "sample") %>%
  filter(sample %in% m$sample[m$at_least_two & m$all_BL])

# clean up names
# remove long strain id, replace with int
new_names <- sub("\\s+\\S+$", "", colnames(MG_data)) # remove strain id
new_names <- make.unique(new_names, sep = " ") # Make names unique
new_names <- sub("_", " ", new_names) # remove strain id
colnames(MG_data) <- new_names # Assign new names

# get colpal for top weights
# it has already colour for strain lvl, but you can also choose for species/genus level
MG_hirarchy <- MG_hirarchy %>%
  arrange(desc(sum)) %>%
  mutate(species = factor(.$species, levels = unique(.$species))) %>%
  mutate(genus = factor(.$genus, levels = unique(.$genus))) %>%
  arrange(genus, species, strain)

MG_colours <- set_names(MG_hirarchy$col, MG_hirarchy$strain)

top_feat <- pull(weights$Metagenomics, "feature")[1:20]
new <- setdiff(top_feat, names(MG_colours))
MG_colours <- set_names(
  c(MG_colours, c(tax_pallet[1:length(new)], "gray")),
  c(names(MG_colours), c(new, "other"))
)

# you can choose if you want to include species by sum abundance or by alfabetical
# dpending on if you use the levels() or just the order of n
n_lvl <- MG_hirarchy$species %>% unique() %>% as.vector()
n <- c(levels(MG_hirarchy$species)[1:20], new)

# I have to get MG_colours to contain all of the taxa at species level in order to set factor levels or i have to apply other before
```

### prepp FM
```{r prepp-FM-data}
# select samples:
FM_data <- FM_data %>%
  select(level, Pathway, any_of(m$sample[m$at_least_two & m$all_BL]))

# clean up columns a bit
FM_data <- FM_data %>%
  separate(col = "Pathway", sep = "\\|", into = c("Pathway", "taxa")) %>%
  separate(col = "Pathway", sep = ": ", into = c("pathway_id", "pathway")) %>%
  mutate(taxa = ifelse(is.na(taxa), "total", taxa)) %>%
  mutate(taxa = gsub("^.+s__", "", .$taxa))

# select community level:
FM_data <- FM_data %>%
  filter(level == "community_lvl")

# add MetaCys hirarchy
mc_env <- new.env()
load("/Users/vilkal/Downloads/MetaCyc_pathway_map.RData", envir = mc_env)
pathway_map <- mc_env[[ls(mc_env)]] %>% as_tibble(rownames = "pathway_id")

FM_data <- FM_data %>%
  left_join(
    pathway_map,
    by = c("pathway_id", "pathway")
  ) %>%
  relocate(
    any_of(c("Superclass1", "Superclass2", "pathway")),
    .after = "pathway"
  )

# col pal
FM_colours <- set_names(FM_colours$col, FM_colours$Pathway)
```

```{r MetaCyc-pathway-hierarchy}
# Number of pathways per superclass
superclass_counts <- c("Superclass1", "Superclass2") %>%
  set_names() %>%
  map(
    .,
    ~ FM_data %>%
      # filter(!pathway_id %in% c("UNMAPPED", "UNINTEGRATED")) %>%
      summarise(
        n_pathways = n_distinct(pathway_id),
        total_abundance = sum(
          rowSums(pick(where(is.numeric)), na.rm = TRUE),
          na.rm = TRUE
        ),
        .by = .x
      ) %>%
      mutate(perc_total = total_abundance / sum(total_abundance) * 100) %>%
      arrange(desc(perc_total))
  )
knitr::kable(superclass_counts[[1]], caption = "Superclass 1", digits = 2)
knitr::kable(superclass_counts[[2]], caption = "Superclass 2", digits = 2)
```

## 6. Combine all omics into one tibble
```{r omics-tibble}
#| code-fold: true
# ============================================================
# 1. STORE BOTH OMICS IN A LIST (LONG FORMAT)
# ============================================================
omic_data <- list()

# -------- Functional (FM) --------
omic_data$Functional <- FM_data %>%
  rename(feature = "pathway") %>%
  filter(!pathway_id %in% c("UNMAPPED", "UNINTEGRATED")) %>%
  pivot_longer(
    cols = where(is.double),
    names_to = "sample",
    values_to = "abundance"
  ) %>%
  mutate(
    abundance = abundance / sum(abundance, na.rm = TRUE),
    .by = "sample"
  )

# -------- Metabolomics (MB) --------
omic_data$Metabolomics <- MB_data %>%
  rename(feature = "names") %>%
  pivot_longer(
    cols = where(is.double),
    names_to = "sample",
    values_to = "abundance"
  )

# -------- Metagenomics (MG) --------
omic_data$Metagenomics <- MG_data %>%
  pivot_longer(
    cols = where(is.double),
    names_to = "strain",
    values_to = "abundance"
  ) %>%
  # create species column
  mutate(species = str_replace(strain, " \\d", "")) %>%
  mutate(genus = str_replace(strain, " \\w+$", "")) %>%
  # set low abundant species to "other"
  mutate(feature = ifelse(.$species %in% n, .$species, "other")) %>%
  # fix order of taxa
  mutate(species = factor(species, levels = c(n_lvl, "other"))) %>%
  # fix order of samples by getting gr before factor.fun()
  mutate(gr = gr.fun(.)[as.character(.$sample)])

sample_order <- factor.fun(omic_data$Metagenomics) %>%
  as.vector() # test

# ============================================================
# 2. DEFINE FEATURE COLUMN + COLOURS + TOP FEATURES
# ============================================================
omics_tbl <- tibble(
  omic = c("Functional", "Metabolomics", "Metagenomics"),
  df = list(
    omic_data$Functional,
    omic_data$Metabolomics,
    omic_data$Metagenomics
  ),
  # feature = c("pathway", "metabolite"),
  feature = c("feature", "feature", "feature"),
  top = c("feature", "feature", "species"),
  colours = list(
    FM_colours,
    MB_colours,
    MG_colours
  ),
  top_features = list(
    pull(weights$Functional, "feature")[1:20],
    pull(weights$Metabolomics, "feature")[1:20],
    pull(weights$Metagenomics, "feature")[1:20]
  )
)

# ============================================================
# 3. STANDARDIZE ALL OMICS DATASETS
# ============================================================
# 1. add meta data
# 2. make feature and fct1_clus factor
# 3. arrange by feature
meta_cols <- c("RVVC_pos", "Clinical score (0-5)", "fct1_clus")

omics_tbl <- omics_tbl %>%
  mutate(df = df, omic, df = set_names(df, omic)) %>%
  mutate(
    df = pmap(
      list(df, colours),
      function(df, colours) {
        # add meta information:
        df %>%
          left_join(
            select(meta, sample, all_of(meta_cols)),
            by = "sample"
          ) %>%
          # make factor:
          mutate(
            # fct1_clus = factor(fct1_clus),
            sample = factor(sample, levels = sample_order),
            across(
              all_of("feature"),
              ~ factor(.x, levels = names(colours))
            )
          ) %>%
          arrange(match(feature, names(colours)))
      }
    )
  )

# ============================================================
# 4. ADD TOP FEATURE DATASETS
# ============================================================
omics_tbl <- omics_tbl %>%
  mutate(
    df_top = pmap(
      list(df, top, top_features),
      function(df, col, vals) {
        filter(df, df[[col]] %in% vals)
      }
    )
  ) %>%
  mutate(df_top = set_names(df_top, omic))
```

## Plot barplot per omic

```{r plots}
# ============================================================
# 5. GENERATE BARPLOTS USING pmap()
# ============================================================
# ---- All features ----
plots_all <- omics_tbl %>%
  mutate(
    plot = pmap(
      list(df, feature, colours),
      function(df, feature, colours) {
        Facet_bar.fun(
          df = df,
          facet_var = "fct1_clus",
          fill = feature,
          colours = colours,
          strip_colours = c("#a4d19b", "#d5edbe", "#ffc99e", "#ffe3c1"),
          x_axis = "fct1_clus"
        )
      }
    )
  )
plots_all$plot
```

### Create sub-grouped legend

```{r get-subgroup-legend}
# ---- Legend ----
# ------------------------------------------------------------
# Function: create legends for one omic
# ------------------------------------------------------------
# df <- plots_all$df_top$Functional
make_legends <- function(df, meta_group, colours) {
  df %>%
    mutate(meta_group_var = .[[meta_group]]) %>%
    # nest subgroup by meta group:
    nest(.by = meta_group_var) %>%

    mutate(
      legend = pmap(
        list(data, meta_group_var),
        function(df_sub, group_name) {
          # map original colours to labels
          legend_colours <- df_sub %>%
            distinct(feature, feature_label) %>%
            {
              setNames(
                colours[.$feature],
                .$feature_label
              )
            }
          get_legend.fun(
            df = df_sub,
            title = group_name,
            feature = "feature_label",
            id = "sample",
            colours = legend_colours,
            txt_size = 8
          )
        }
      )
    ) %>%
    # get height of indvcidual meta groups:
    mutate(
      height = map_dbl(
        legend,
        ~ as.numeric(legend_height_cm(.x))
      ),
      rel_height = height / sum(height)
    )
}

# ------------------------------------------------------------
# Function: combine legends into one grob
# ------------------------------------------------------------
combine_legends <- function(legend_df) {
  plot_grid(
    plotlist = legend_df$legend,
    ncol = 1,
    rel_heights = legend_df$rel_height,
    align = "v"
  )
}

# ------------------------------------------------------------
# Define which metadata column belongs to each omic
# ------------------------------------------------------------
meta_map <- c(
  Functional = "Superclass2",
  Metabolomics = "MB_Class2",
  Metagenomics = "genus"
)

# ------------------------------------------------------------
# Generate legend plots
# ------------------------------------------------------------
# shorten name
# df <- omics_tbl$df_top$Functional %>% filter(feature == "superpathway of N-acetylglucosamine, N-acetylmannosamine and N-acetylneuraminate degradation")
preprocess_legend_df <- function(df, omic) {
  df <- df %>%
    mutate(feature_label = feature)
  if (omic == "Functional") {
    df <- df %>%
      mutate(
        feature_label = str_replace(
          feature,
          "^(superpathway of [^,]+),.*",
          "\\1"
        )
      )
  }
  df
}

plots_all <- plots_all %>%
  mutate(
    meta_gr = meta_map[omic],
    legend_data = pmap(
      list(df_top, meta_gr, colours, omic),
      function(df, meta_group, colours, omic) {
        df <- preprocess_legend_df(df, omic)
        make_legends(
          df = df,
          meta_group = meta_group,
          colours = colours
        )
      }
    ),

    legend_plot = map(
      legend_data,
      combine_legends
    )
  )

plots_all$legend_plot$Functional
#plots_all <- plots_all %>% select(omic, meta_gr, df, plot, legend_plot)
plots_all
```

### Combine barplot + legend

```{r all-FM-plot}
#| fig.width: 6
#| fig.height: 4
# quartz(width = 6, height = 4, dpi = 120)
plot_grid(
  plots_all$plot$Functional,
  plots_all$legend_plot$Functional,
  rel_widths = c(0.4, 1.1)
) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
```

```{r all-MB-plot}
#| fig.width: 3.5
#| fig.height: 4
# quartz(width = 3.5, height = 4, dpi = 120)
plot_grid(
  plots_all$plot$Metabolomics,
  plots_all$legend_plot$Metabolomics,
  rel_widths = c(1, 1.2)
) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
```

```{r all-MG-plot}
#| fig.width: 2.9
#| fig.height: 4
# quartz(width = 2.9, height = 4, dpi = 120)
plot_grid(
  plots_all$plot$Metagenomics,
  plots_all$legend_plot$Metagenomics,
  rel_widths = c(1, 0.8)
) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
```

```{r old-legend-chunk}
#| eval: false
#| code-fold: true
# old working version of legend, remove when above has been confirmed
meta_group <- c("Superclass2", "MB_Class2", "genus")
# df <- omics_tbl$df_top$Metagenomics
plots_top <- plots_top %>%
  mutate(meta_gr = meta_group) %>%
  mutate(
    legend_data = pmap(
      list(df_top, meta_gr, colours),
      function(df, meta_gr, colours) {
        df %>%
          # ensure grouping column exists
          mutate(meta_group_var = .[[meta_gr]]) %>%
          nest(.by = meta_group_var) %>%

          # get height of indvcidual meta groups
          mutate(n_s = map_int(data, ~ n_distinct(.x[["feature"]]))) %>%

          # build legend per group
          mutate(
            legend = pmap(
              list(data, meta_group_var),
              function(df_sub, group_name) {
                get_legend.fun(
                  df = df_sub,
                  title = group_name,
                  feature = "feature",
                  id = "sample",
                  colours = colours,
                  txt_size = 8
                )
              }
            )
          ) %>%

          # compute heights for stacking legends
          mutate(
            height = map_dbl(legend, ~ as.numeric(legend_height_cm(.x))),
            rel_height = height / sum(height)
          )
      }
    )
  )
plots_top$legend_data

legend_plots <- plots_top %>%
  mutate(
    legend_plot = map(legend_data, function(x) {
      plot_grid(
        plotlist = x$legend,
        ncol = 1,
        rel_heights = x$rel_height,
        align = "v"
      )
    })
  )

# quartz(width = 6, height = 4, dpi = 120)
plot_grid(FM_fct1_bar, l, rel_widths = c(0.4, 1.1)) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
plot_grid(FM_fct1_bar_top, l, rel_widths = c(0.4, 1.1)) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
```

```{r top-features}
# quartz(width = 10.5, height = 6 )
# ---- Top features ----
plots_top <- plots_all %>%
  mutate(
    plot = pmap(
      list(df_top, top, colours),
      function(df, feature, colours) {
        Facet_bar.fun(
          df = df,
          facet_var = "fct1_clus",
          fill = feature,
          colours = colours,
          strip_colours = c("#a4d19b", "#d5edbe", "#ffc99e", "#ffe3c1"),
          x_axis = "fct1_clus"
        )
      }
    )
  )
plots_top
```

```{r top-FM-plot}
#| fig.width: 6
#| fig.height: 4
# quartz(width = 4, height = 4, dpi = 120)
C2 <- plot_grid(
  plots_top$plot$Functional,
  plots_all$legend_plot$Functional,
  rel_widths = c(0.5, 1)
) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
C2
```

```{r top-MB-plot}
#| fig.width: 3.5
#| fig.height: 4
# quartz(width = 3.2, height = 4, dpi = 120)
C3 <- plot_grid(
  plots_top$plot$Metabolomics,
  plots_all$legend_plot$Metabolomics,
  rel_widths = c(0.7, 1)
) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
C3
```

```{r top-MG-plot}
#| fig.width: 2.9
#| fig.height: 4
# quartz(width = 2.6, height = 4, dpi = 120)
C4 <- plot_grid(
  plots_top$plot$Metagenomics,
  plots_all$legend_plot$Metagenomics,
  rel_widths = c(0.43, 0.42)
) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
C4
```

```{r individual-samples}
#| fig.width: 9.4
#| fig.height: 3.5
# ---- All samples ----
plots_all <- plots_all %>%
  mutate(
    plot_s = pmap(
      list(df, colours),
      function(df, colours) {
        Facet_bar.fun(
          df = df, #mutate(df, sample = forcats::fct_relevel(sample, sample_order)),
          facet_var = "fct1_clus",
          fill = "feature",
          colours = colours,
          strip_colours = c("#a4d19b", "#d5edbe", "#ffc99e", "#ffe3c1"),
          x_axis = "sample",
          annot_col = "RVVC_pos",
          legend = "bottom"
        )
      }
    )
  )
# quartz(width = 9.4, height = 3.5, dpi = 120)
plots_all$plot_s$Metagenomics
# df <- omics_tbl$df_top$Metagenomics

# ---- All samples top ----
plots_top <- plots_top %>%
  mutate(
    plot_s = pmap(
      list(df_top, colours),
      function(df, colours) {
        Facet_bar.fun(
          df = df, #mutate(df, sample = forcats::fct_relevel(sample, sample_order)),
          facet_var = "fct1_clus",
          fill = "feature",
          colours = colours,
          strip_colours = c("#a4d19b", "#d5edbe", "#ffc99e", "#ffe3c1"),
          x_axis = "sample",
          #annot_col = "RVVC_pos"
        )
      }
    )
  )
plots_top$plot_s$Metagenomics

#plots_top <- plots_top %>% select(omic, meta_gr, plot, plot_s, legend_plot)
```

## 7. MB line plots

```{r extra-plots}
#| code-fold: true
df_summary <- omics_tbl$df_top$Metabolomics %>%
  group_by(fct1_clus, feature, MB_Class3, MB_Class2, MB_Class1) %>%
  summarise(
    mean_abund = mean(abundance, na.rm = TRUE),
    se = sd(abundance, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  left_join(., select(weights$Metabolomics, feature, value)) %>%
  mutate(facet = ifelse(value > 0, "up", "down"))

ggplot(
  df_summary,
  aes(fct1_clus, mean_abund, group = feature, color = feature)
) +
  geom_line(alpha = 0.5) +
  geom_point(size = 1.5) +
  geom_errorbar(
    aes(ymin = mean_abund - se, ymax = mean_abund + se),
    width = 0.1
  ) +
  scale_color_manual(values = MB_colours) +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(~facet, ncol = 1)


ggplot(
  df_summary,
  aes(fct1_clus, mean_abund, group = feature, color = feature)
) +
  geom_line() + # group = 1
  geom_point() +
  #facet_wrap(~ metabolite, scales = "free_y") +
  facet_wrap(~MB_Class2, scales = "free_y", ncol = 2) +
  theme_minimal() +
  scale_color_manual(values = MB_colours)
```

```{r iners-only-samples}
#| eval: false
#| fig.width: 6
#| fig.height: 4
MG_iners <- MG_data %>%
  select(sample, contains("iners")) %>%
  arrange(desc(`Lactobacillus_iners SGB7025`)) %>%
  filter(`Lactobacillus_iners SGB7025` >= 50)

# quartz(width = 6, height = 4, dpi = 120)
FM_fct1_bar_iners <- FM_df_top %>%
  filter(sample %in% MG_iners$sample) %>%
  mutate(sample = factor(.$sample, levels = MG_iners$sample)) %>%
  Facet_bar.fun(
    .,
    "fct1_clus",
    fill = "pathway",
    colours = FM_colours,
    strip_colours = c("#a4d19b", "#d5edbe", "#ffc99e", "#ffe3c1"),
    x_axis = "sample",
    annot_col = "Clinical score (0-5)"
  )
FM_fct1_bar_iners
```

## 8. Combine plots

```{r combine-plot-A}
#| fig-width: 13
#| fig-height: 5
# quartz(width = 13, height = 2.5, dpi = 110)

# quartz(width = 11.5, height = 4, dpi = 100)

A <- plot_grid(
  A1,
  plots_all$plot_s$Metagenomics,
  rel_widths = c(0.25, 0.75),
  nrow = 1
)
# C <- plot_grid(NULL, C3, C2, C4, rel_widths = c(0.15, 0.29, 0.36, 0.24), nrow = 1) plots_top$plot$Metagenomics
C <- plot_grid(
  NULL,
  C3,
  C2 + theme(plot.margin = unit(c(0, 0.3, 0, -0.2), "cm")),
  plots_top$plot$Metagenomics,
  NULL,
  rel_widths = c(0.18, 0.28, 0.37, 0.13, 0.07),
  nrow = 1
)
```

```{r combine-plots}
#| fig.width: 11.5
#| fig.height: 9
#############################
# COMBINE ALL FIGURE PANELS #
#############################
# quartz(width = 5.5, height = 4, dpi = 150)

# quartz(width = 11.5, height = 9, dpi = 100)

Figure3 <- plot_grid(
  A,
  B,
  C,
  rel_heights = c(0.27, 0.33, 0.39),
  labels = c('A', 'B', 'C'),
  ncol = 1
)

ggsave(
  "./Figures/Figure3.png",
  Figure3,
  width = 11.5,
  height = 9,
  bg = "white",
  dpi = 600
)

Figure3
```



