---
title: "Figure 2"
date: today
date-format: "D MMM YYYY"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    highlight-style: github
    code-tools:
      source: true
      toggle: true
---
## 1. Libraries and global settings

```{r libraries}
#| code-fold: true
#| output: false
#| warning: false
#| message: false
library(MOFA2)
library(tidyverse)

library(RColorBrewer)
library(ggExtra)
library(patchwork)
library(cowplot)
library(gridGraphics)

# setwd("/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/manuscript")

select <- dplyr::select
map <- purrr::map
rename <- dplyr::rename
```


```{r load-data}
# meta data
comb_meta_path <- "../Results/Combined_gr_meta_data.csv"
meta <- read_csv(comb_meta_path) %>%
  #rename(sample = "svamp_ID") %>%
  mutate(across(2:27, ~ factor(.x)))


# colour pallet
pal <- list(
  "Clinical_score_(0-5)" = brewer.pal(6, name = "Reds"),
  RVVC_pos = c(
    "#A6CEE3",
    "#1F78B4",
    "#e490c4ff",
    "#c12d88ff"
  )
)
RdBu <- rev(c(
  "#67001F",
  "#B2182B",
  "#D6604D",
  "#F4A582",
  "#FDDBC7",
  "#FFFFFF",
  "#D1E5F0",
  "#92C5DE",
  "#4393C3",
  "#2166AC",
  "#053061"
))
```

## 2. Load trained MOFA model
We load the model with the highest elbo score

```{r load-MOFA-model}
#| eval: true
#| include: true
#| warning: false
#| message: false
#| tbl-colwidths: [60,40]
elbo_table <- read_csv("../Results/MOFA/elbo_table_noS84_1e-6_57MG.csv")
knitr::kable(elbo_table)
best_seed <- elbo_table$seed[which.max(abs(elbo_table$elbo))]

MOFAmodel <- load_model(
  paste0("../Results/MOFA/MOFA_allBL_noS84_1e-6_57MG_seed_", best_seed, ".hdf5")
)

# add meta data to model
samples_metadata(MOFAmodel) <- samples_metadata(MOFAmodel)[, c(
  "sample",
  "group"
)] %>%
  left_join(meta, by = c("sample"))
```

```{r get-variance-df}
ve_df <- get_variance_explained(MOFAmodel)$r2_per_factor$single_group %>%
  as.data.frame() %>%
  rownames_to_column("Factor") %>%
  pivot_longer(
    -Factor,
    names_to = "view",
    values_to = "variance_explained"
  ) %>%
  mutate(Factor = factor(.$Factor, levels = paste0("Factor", 1:15)))
```

```{r variance-stats}
# summarized by view for Factor 1-10:
ve_df %>%
  filter(grepl("^Factor[1-9]$|^10$", .$Factor)) %>%
  summarize(sum = sum(variance_explained), .by = c("view")) %>%
  add_case(., view = "Total", sum = sum(.$sum))

# Factors with <2% in all views
ve_df %>%
  summarise(all_below_2 = all(variance_explained < 2), .by = c("Factor")) %>%
  filter(all_below_2)

# summarized by view for Factor 1:
ve_df %>%
  filter(grepl("^Factor1$", .$Factor)) %>%
  #filter(grepl("^Factor[1-9]$", .$Factor)) %>%
  summarize(sum = sum(variance_explained), .by = c("Factor", "view"))

# Sum variance per factor:
ve_df %>%
  #filter(grepl("^Factor[1-9]$", .$Factor)) %>%
  summarize(sum = sum(variance_explained), .by = c("Factor"))
```


```{r factor-with-distribution-function}
# ----------------------
# Plot Factor Density Function
# ----------------------
plot_mofa_factor_vs_variable <- function(
  df,
  factor = 1,
  y_var,
  palette = pal[[y_var]],
  jitter_width = 0.1,
  add_marginal = TRUE,
  marginal_type = "density"
) {
  # jitter object
  jitterer <- position_jitter(width = jitter_width, seed = 123)

  df_ <- df %>% mutate(!!sym(y_var) := factor(!!sym(y_var)))

  # Main ggplot
  p <- ggplot(
    df,
    aes(
      x = .data[[paste0("Factor", factor)]],
      y = .data[[y_var]],
      colour = .data[[y_var]]
    )
  ) +
    geom_point(
      size = 1,
      alpha = 0.8,
      show.legend = FALSE,
      position = jitterer
    ) +
    geom_point(
      aes(colour = NULL),
      pch = 21,
      size = 1,
      alpha = 0.4,
      show.legend = FALSE,
      colour = "gray",
      position = jitterer
    ) +
    geom_boxplot(
      aes(colour = .data[[y_var]]),
      fill = "transparent",
      box.linewidth = 0.2,
      whisker.linewidth = 0.2,
      show.legend = FALSE,
      outlier.shape = NA,
      median.color = "black"
    ) +
    coord_cartesian(expand = TRUE) +
    theme_minimal() +
    theme(
      panel.grid = element_line(linewidth = 0.2),
      text = element_text(size = 7),
      plot.margin = unit(c(0, 0, 0, -0), "cm"),
      axis.text.y.left = element_blank(),
      axis.title.x = element_text(vjust = 1),
      axis.title.y.left = element_blank()
    )

  # Add palette if provided
  if (!is.null(palette)) {
    p <- p + scale_colour_manual(values = palette)
  }

  # Add marginal plot if requested
  if (add_marginal) {
    p <- ggMarginal(
      p,
      type = marginal_type,
      groupColour = TRUE,
      groupFill = TRUE,
      trim = FALSE,
      linewidth = 0.2,
      margins = "x",
      size = 1
    )
  }
  #p <- p + theme( plot.margin = unit(c(0, 0, -1, -0), "cm"))
  return(p)
}
```

```{r factor-with-distribution-plots}
#| fig-width: 3
#| fig-height: 2
# ----------------------
# Prepp df
# ----------------------
# Extract factors, reshape wide, join metadata
df <- get_factors(MOFAmodel, factors = c(1:4), as.data.frame = TRUE) %>%
  as_tibble() %>%
  select(-group) %>%
  #mutate(fct1_clus = as.character(clusters$cluster[.$sample])) %>%
  pivot_wider(names_from = "factor") %>%
  left_join(samples_metadata(MOFAmodel), by = "sample")

# ----------------------
# Plot
# ----------------------
# quartz(width = 3, height = 2, dpi = 150)
A1 <- plot_mofa_factor_vs_variable(
  df,
  y_var = "Clinical_score_(0-5)"
)
A1
plot.new()

A2 <- plot_mofa_factor_vs_variable(
  df,
  y_var = "RVVC_pos"
)
A2
```

```{r UMAP-function}
#| code-fold: true
# ----------------------
# Plot UMAP Function
# ----------------------
UMAP_fun <- function(
  umap_df,
  group_var,
  txt_var = NULL,
  shape_gr = NULL,
  shape = NULL,
  title = ""
) {
  if (is.null(shape_gr)) {
    point <- geom_point(size = 1, alpha = 1, shape = 21) # , show.legend = F
  } else {
    point <- geom_point(
      size = 2,
      alpha = 1,
      aes(shape = if (!is.null(shape_gr)) .data[[shape_gr]] else NULL)
    )
  }

  p <- ggplot(
    umap_df,
    aes(
      x = `UMAP 1`,
      y = `UMAP 2`,
      #shape = RVVC_AS,
      fill = .data[[group_var]],
      color = .data[[group_var]]
    )
  ) +
    point +
    #geom_jitter( shape=21, size=3, color="white", width=.5, height=.5) +  # Tassos used jitter

    scale_fill_manual(
      values = pal[[group_var]],
      na.value = "transparent",
      name = ""
      #aesthetics = c("colour", "fill")
    ) +
    guides(
      fill = guide_legend(
        override.aes = list(size = 2),
        keyheight = .5,
        keywidth = .5
      )
    ) +
    scale_colour_manual(
      values = pal[[group_var]],
      na.value = "gray",
      name = ""
    ) +
    {
      if (!is.null(shape_gr) && !is.null(shape)) {
        scale_shape_manual(values = shape, name = "")
      }
    } +
    # scale_shape_manual(values = c(21, 5, 4), name = "" ) + #specify individual shapes+
    {
      if (!is.null(txt_var)) {
        geom_text(
          data = umap_df,
          aes(x = `UMAP 1`, y = `UMAP 2`, label = .data[[txt_var]]),
          size = 1,
          hjust = 0,
          nudge_x = 0.07,
          color = "gray51"
        )
      }
    } +
    ggtitle(title) +
    theme_bw(base_size = 7) +
    theme(
      plot.title = element_text(vjust = -1, size = 7),
      plot.margin = unit(c(0, 0, 0, -0), "cm"),
      legend.background = element_blank(),
      legend.title = element_blank(),
      legend.margin = margin(0, -0, -0, -0, unit = "cm"), # moves the legend box
      legend.box.margin = margin(-0, -0, -0, -0, unit = "cm"), #t,r,b,l
      legend.box.spacing = unit(0, "cm"), # space between plot and legend
      # legend.box.background = element_rect(colour = "black"), See the outline of the box
      # point = element_point(shape = 21),
      line = element_blank(),
      axis.ticks = element_line(color = "black"),
      # aspect.ratio = 1,
      axis.title = element_blank()
    )
  return(p)
}

# ----------------------
# Plot UMAP Panel Function
# ----------------------
#### Combine and save plots ####
make_umap_panel <- function(var, umap_dfs, outdir = "../Results/UMAP2/") {
  # create plots for each dataset
  p_list <<- imap(
    umap_dfs,
    ~ UMAP_fun(.x, var, "txt_clin", title = .y)
  )
  l <- get_legend(p_list[[1]])
  # combine with shared legend
  comb <- map(p_list, ~ .x + theme(legend.position = "none")) %>%
    plot_grid(plotlist = ., nrow = 1)
  comb <- plot_grid(comb, l, rel_widths = c(1, 0.07))

  # safe filename
  fname <- gsub("[^A-Za-z0-9_]", "_", var)

  # save
  ggsave(
    filename = paste0(outdir, "UMAP_", fname, ".png"),
    plot = comb,
    width = 14,
    height = 5,
    dpi = 300
  )

  return(comb)
}
```

```{r plot-UMAP}
#| fig-width: 10
#| fig-height: 2.5
#| code-fold: true
#| warning: false
# ----------------------
# Prepp df
# ----------------------
#### add labels ####
annot_col <- meta %>%
  select(
    1:6,
    `Clinical_score_(0-5)`
  ) %>%
  mutate(across(2:ncol(.), ~ factor(.x))) %>%
  mutate(
    txt_clin = ifelse(
      grepl(
        "0",
        .$`Clinical_score_(0-5)`
      ),
      NA,
      as.character(
        .$`Clinical_score_(0-5)`
      )
    )
  ) %>%
  mutate(txt_id = ifelse(RVVC == "RVVC", .$sample, NA))

#### UMAP analysis ####
set.seed(1)

omic_ord <- c("Transcriptomics", "Metabolomics", "Metagenomics", "Functional")
data <- get_data(MOFAmodel)

umap_dfs <- data[omic_ord] %>%
  map(., ~ t(.x$single_group)) %>%
  map(
    .,
    ~ uwot::umap(
      na.omit(.x),
      #seed = 123,
      seed = 463,
      n_neighbors = 15,
      #metric = "cosine",
      init = "spectral",
      scale = T
    )
  ) %>%
  map(
    .,
    ~ tibble(
      "UMAP 1" = .x[, 1],
      "UMAP 2" = .x[, 2],
      "ID" = rownames(.x)
    )
  ) %>%
  map(., ~ left_join(., annot_col, by = c("ID" = "sample")))


color_vars <- c(
  "Clinical_score_(0-5)",
  "RVVC_pos"
) %>%
  set_names(.)

plots <- map(color_vars, ~ make_umap_panel(.x, umap_dfs))
# quartz(width = 10, height = 2.5, dpi = 150)
plots
```

```{r combine-plot-A}
#| fig-width: 13
#| fig-height: 5
# quartz(width = 13, height = 2.5, dpi = 150)

A1 <- plot_grid(
  A1,
  plots$`Clinical_score_(0-5)`,
  rel_widths = c(0.25, 0.75),
  nrow = 1
)
A2 <- plot_grid(A2, plots$RVVC_pos, rel_widths = c(0.25, 0.75), nrow = 1)

A <- plot_grid(A1, A2, nrow = 2, labels = c('A'))
# quartz(width = 13, height = 5, dpi = 150)
A
```

```{r variance-by-factor}
#| fig-width: 6
#| fig-height: 4
B <- ve_df %>%
  filter(grepl("^Factor[1-9]$", .$Factor)) %>%
  ggplot(., aes(x = Factor, y = variance_explained, fill = view)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c("#CC79A7", "#D55E00", "#CF1C90", "#E69F00"),
    na.value = "gray",
    name = ""
  ) +
  labs(
    x = "Factor",
    y = "Variance explained (%)",
    fill = "view"
  ) +
  guides(fill = guide_legend(keywidth = .5, keyheight = .5)) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  theme_minimal(base_size = 8) +
  theme(
    axis.title = element_blank(),
    legend.position = "inside",
    legend.position.inside = c(0.75, 0.85), # Places legend at top-right
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# quartz(width = 2, height = 4)
B
```


### Why we can use `mean()` to calculate proportions

In R, logical values (`TRUE` and `FALSE`) are automatically treated as numbers:
- `TRUE` becomes 1  
- `FALSE` becomes 0  

This allows us to compute proportions directly using `mean()`.

**Key idea**
If we evaluate a condition like:
    value == 1

we get a logical vector indicating whether each observation satisfies the condition.

Taking the mean of this vector calculates:
    (number of TRUE values) / (total number of observations)

which is exactly the definition of a proportion.


```{r scores-figure}
d <- samples_metadata(MOFAmodel)[, c(1, 7, 28:39)] %>%
  as_tibble() %>%
  mutate(across(4:ncol(.), as.integer)) %>%

  pivot_longer(
    cols = matches("^(Clinical|Symtom) score:"),
    names_to = c("type", "variable"),
    names_pattern = "(Clinical|Symtom) score: (.*)",
    values_to = "value"
  ) %>%
  mutate(variable = sub(" \\(y/n\\)$", "", variable)) %>%
  mutate(
    type = recode(
      type,
      "Clinical" = "Clinician-assessed",
      "Symtom" = "Self-reported"
    ), # fix spelling
    score = ifelse(
      type == "Clinician-assessed",
      as.character(`Clinical score (0-5)`),
      as.character(`Symptom score (0-5)`)
    ),
  ) %>%
  mutate(score = factor(.$score)) %>%
  select(sample, RVVC_pos, type, score, variable, value) %>%
  distinct()


combined_prop <- d %>%
  group_by(type, score, variable) %>%
  summarise(
    sum = sum(value == 1),
    prop = mean(value == 1),
    .groups = "drop"
  )


# plot
brighter_pastels <- c(
  "#FF69B4", # hot pink
  "#FFB6C1", # light pink
  "#F4A460", # sandy peach
  "#BA55D3", # orchid
  "#87CEFA", # light blue
  #"#98FB98", # mint green
  "#EE82EE", # violet
  # "#FDB462",
  "#B3DE69",
  #"#FCCDE5",
  #"#D9D9D9"
  "#BC80BD"
) %>%
  set_names(., unique(arrange(d, type, variable) %>% .$variable))
```

```{r prop-figure}
#| fig.width: 5.5
#| fig.height: 4
# proportions
# quartz(width = 4, height = 4)
C2 <- combined_prop %>%
  ggplot(aes(x = score, y = prop, color = variable, group = variable)) +
  geom_line(size = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~type, ncol = 1) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(values = brighter_pastels, name = "feature") +
  labs(
    x = "Score (0–5)",
    y = "% of participants with feature present",
    #title = "Feature prevalence trends across clinical score per group"
  ) +
  guides(col = guide_legend(keywidth = .5)) +
  theme_minimal(base_size = 7) +
  theme(
    legend.position = "bottom",
    legend.key.height = unit(0.3, "cm"),
    legend.box.spacing = unit(0, "pt"),
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 0, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

```{r counts-figure}
#| fig.width: 5.5
#| fig.height: 4
# count
# quartz(width = 6, height = 4)
combined_prop %>%
  ggplot(aes(x = score, y = sum, color = variable, group = variable)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~type, ncol = 2) +
  #scale_y_continuous(labels = scales::percent_format()) +
  scale_color_manual(values = brighter_pastels, name = "") +
  labs(
    x = "Score (0–5)",
    y = "number of participants with feature present",
    title = "Feature prevalence trends across clinical/symptom score"
  ) +
  guides(col = guide_legend(byrow = TRUE)) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    #legend.spacing.y = unit(0.5, 'cm'),
    legend.key.height = unit(0.3, "cm"),
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

```{r covariates-factor-correlation}
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 6
#samples_metadata(MOFAmodel) <- samples_metadata(MOFAobject)

# improve names
m <- meta %>%
  select(sample, 27:38) %>%
  rename_with(., ~ str_replace(.x, "Clinical score:", "Clin:")) %>%
  rename_with(., ~ str_replace(.x, "Symtom score:", "Symp:"))

samples_metadata(MOFAmodel) <- samples_metadata(MOFAmodel)[, c(
  "sample",
  "group"
)] %>%
  left_join(m, by = c("sample"))


covariates <- samples_metadata(MOFAmodel)[, -c(1:2)] %>% colnames()

# remove columns
# covariates <- covariates[!(grepl("Follow_up", covariates))]
Clin_info <- list(
  Clin_score = covariates # Clinical score + Symptom score
)

# reproductive info
# quartz(width = 6, height = 6, dpi = 100)
C1 <- function() {
  par(mar = c(-0, 0, 0, 0), bg = NA)
  correlate_factors_with_covariates(
    MOFAmodel,
    col = colorRampPalette(RdBu)(100),
    covariates = covariates,
    factors = 1:9,
    tl.cex = 0.5,
    tl.offset = 0.2,
    cl.cex = 0.5,
    win.asp = 1, # max(6, max(nchar(.x)) * 0.18) / max(3, length(.x) * 0.35) ,
    plot = "r", # use "log_pval" to plot log p-values
    main = "",
    transpose = T
  )
}
ggdraw(C1)
# ggdraw(C1) + theme(plot.background = element_rect(fill = "cornsilk"))
```

```{r}
# quartz(width = 4, height = 3.5)
C <- plot_grid(ggdraw(C1), C2, ncol = 2, rel_widths = c(1, 1))

# quartz(width = 8, height = 3.5)
B_C <- plot_grid(B, C, ncol = 2, rel_widths = c(0.3, 1), labels = c('B', 'C'))
```
```{r combine-plots}
#| fig.width: 8
#| fig.height: 6
#############################
# COMBINE ALL FIGURE PANELS #
#############################
# quartz(width = 8, height = 6, dpi = 150)
Figure2 <- plot_grid(A, B_C, ncol = 1, rel_heights = c(1, 1))
ggsave(
  "./Figures/Figure2.pdf",
  Figure2,
  width = 8,
  height = 6,
  bg = "white",
  dpi = 1000
)

Figure2
```
