---
title: "Multi-omics integration with MOFA2"
date: today
date-format: "D MMM YYYY"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    highlight-style: github
    code-tools:
      source: true
      toggle: true
params:
  sparse: false
---

# Overview

This report presents a **structured MOFA2 analysis** integrating transcriptomics, metagenomics, and metabolomics data. The goal is to identify **latent factors capturing shared and view-specific sources of variation**, and to generate a **set of publication-ready figures** that can be combined into a multi‑panel figure.

**Main outcomes**:
- Latent factor structure across multi‑omics layers
- Variance explained per factor and per view
- Associations between factors and clinical / technical covariates
- Feature loadings driving key factors

---

## Analysis workflow (step‑by‑step)

1. Load required libraries and define plotting palettes
2. Load omics data and sample metadata
3. Format MOFA input
4. Select samples for inclusion
5. Unsupervised UMAPs
6. Configure and run the MOFA model
7. Load trained MOFA model
8. Perform model diagnostics and quality checks
9. Covariates- Factor correlation
10. Annotate latent factors using group information
11. Feature-level figures
12. Unified UMAP (all views)

---

## 1. Libraries and global settings

```{r libraries}
#| code-fold: true
#| output: false
#| warning: false
#| message: false
library(MOFA2)
library(tidyverse)
library(ggpubr)
library(GGally)
library(psych)
library(pheatmap)
library(cowplot)
library(RColorBrewer)
library(readxl)
library(ComplexHeatmap)
library(circlize)
library(patchwork)

# setwd("/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/src")
```

Define consistent colour palettes to ensure visual coherence across all figures.
```{r colour-pallet}
#| code-fold: true
select <- dplyr::select
map <- purrr::map
rename <- dplyr::rename

pal <- list(
  CST = c(
    "I" = "#148F77",
    "II" = "#6d9f06",
    "III" = "#A3E4D7",
    "IV-B" = "#E794C1",
    "IV-C" = "#b5cb8c",
    "V" = "#43BA8F"
  ),
  subCST = c(
    "#a5c97b",
    "#cc93cf",
    "#de9548",
    "#63d3b4",
    "#9e8dec",
    "#902267",
    '#FB5273',
    "#91c5de"
  ),
  RVVC = c("#EAEAEA", "#fd814e"),
  BV = c("#EAEAEA", "#fd814e"),
  NAC = c("#EAEAEA", "#fd814e"),
  ST = c("#EAEAEA", "#fd814e"),
  missing_page = c("#EAEAEA", "#fd814e"),
  Clin = c("#a5c97b", "#cc93cf", "#de9548", "#63d3b4", "#9e8dec", "#902267"),
  Clin_gr = c("#ffa998", '#FB5273', "#902267"),
  RVVC_AS = c("#ffa998", '#FB5273', "#902267"),
  RVVC_pos = c(
    "#A6CEE3",
    "#1F78B4",
    #"#EAEAEA",
    #"gray",
    "#e490c4ff",
    "#c12d88ff"
  ), # "#FB9A99", "#E31A1C", "#B2DF8A", "#609c5bff", "#FDBF6F", "#FF7F00",, "#CAB2D6", "#6a3d9a" "#e490c4ff", "#c12d88ff"
  trx_louvain_7 = c(
    "#de9548",
    "#cc93cf",
    "#a5c97b",
    "#9e8dec",
    "#902267",
    "#63d3b4",
    '#FB5273'
  ),
  metab_louvain = c(
    "#cc93cf",
    "#a5c97b",
    "#de9548",
    "#902267",
    "#9e8dec",
    "#63d3b4"
    #'#FB5273'
  ),
  microb_louvain = c(
    "#cc93cf",
    "#de9548",
    "#a5c97b",
    "#9e8dec",
    "#902267",
    "#63d3b4",
    '#FB5273'
  ),
  fun_louvain = c(
    "#a5c97b",
    "#cc93cf",
    "#de9548",
    "#63d3b4",
    "#9e8dec",
    "#902267",
    '#FB5273'
    #"#984EA3"
    #"#B3CDE3"
  ),
  integ_louvain = c(
    "#a5c97b",
    "#cc93cf",
    "#de9548",
    "#63d3b4",
    "#9e8dec",
    "#902267",
    '#FB5273',
    "#984EA3"
    #"#B3CDE3"
  ),
  "Wet_smear:_clue_cells_(y/n)" = c("#2266ac", "#91c5de"),
  hyphae = c("#f26386", "#fd814e", "#a4d984", "#fbbc52"),
  pos = c("#91c5de", "#2266ac"),
  fct1_clus = c("#33A02C", "#B2DF8A", "#FF7F00", "#FDBF6F"), #"#FDBF6F" "#FF7F00", "#CAB2D6", "#6A3D9A"
  "Clinical_score_(0-5)" = brewer.pal(6, name = "Reds")
)

RdBu <- rev(c(
  "#67001F",
  "#B2182B",
  "#D6604D",
  "#F4A582",
  "#FDDBC7",
  "#FFFFFF",
  "#D1E5F0",
  "#92C5DE",
  "#4393C3",
  "#2166AC",
  "#053061"
))

# default ggplot2 pallet:
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
gg_color_hue(4)
```

## 2. Data loading
Multi‑omics matrices and sample metadata are loaded from pre‑processed RDS/CSV files. These data have already undergone quality control and normalisation upstream.
```{r load_data}
#| warning: false
#| message: false
Clin_path <- "/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/data/Metadata_svamp_FINAL.xlsx"
comb_meta_path <- "../Results/Combined_gr_meta_data.csv"

# meta data
Clin_data <- read_xlsx(Clin_path, sheet = "Metadata", na = "na")
meta <- read_csv(comb_meta_path) %>%
  #rename(sample = "svamp_ID") %>%
  mutate(across(2:27, ~ factor(.x)))
top_hvg <- read_csv("../Results/Tissue_QC/HVG.csv")


TRX_path <- "../Results/MixOmic/Transcriptomics_batch_filt.csv"
MG_path <- "../Results/MixOmic/Metageonome_vagswab_CLR_transformed.RDS"
MB_path <- "../Results/MixOmic/Metabolom_Normalized.csv"
FM_path <- "../Tidy_data/Metagenomics/Functional_annotation_CLR_filtered.RDS"

# expression data
TRX_batch <- read.csv(TRX_path, row.names = 1)
MG_matrix <- readRDS(MG_path)
MB_matrix <- read.csv(MB_path, row.names = 1)
FM_matrix <- readRDS(FM_path)
```

## 3. Prepare MOFA input

### 4.1 Assemble omics views
MOFA describes each omic dataset as a "view". One of the possible inputs to mofa is a list of matrixes for each view.
Each omics layer is formatted as a **features × samples** matrix. Sample ordering needs to be the same.
The top 5000 most higly variable genes identified in the `QC_Transcriptomics.qmd` script is used to filter the transcriptomics data.
```{r assemble_views}
# remove long strain id, replace with int
MG_matrix <- MG_matrix[["clr_data"]][["strain"]]
new_names <- sub("\\s+\\S+$", "", colnames(MG_matrix)) # remove strain id
new_names <- make.unique(new_names, sep = " ") # Make names unique
colnames(MG_matrix) <- new_names # Assign new names

# select 5000 most higly variable genes
TRX_HVG <- TRX_batch[top_hvg$gene[1:5000], ]

views <- list(
  Transcriptomics = as.matrix(TRX_HVG),
  Metabolomics = as.matrix(MB_matrix),
  Metagenomics = unclass(t(MG_matrix)),
  Functional = unclass(FM_matrix$path_lvl)
)
```

### 4.2 Convert matrices to long format
The matrix list input is only valid if you have matching samples for all "views". If you wish to include samples that are not "complete" in this sense the long format version is another input alternative to create a MOFA model
```{r omics_to_long}
omics_to_long <- function(mat, view_name) {
  mat %>%
    as.data.frame() %>%
    rownames_to_column("feature") %>%
    pivot_longer(
      cols = -feature,
      names_to = "sample",
      values_to = "value"
    ) %>%
    mutate(view = view_name)
}

mofa_df <- imap_dfr(views, ~ omics_to_long(.x, .y))
```

### 4.3 Sample criteria df
The tibble called `m` shows what samples excist across views for each study participant
```{r identify-samples-overlap}
# select samples to include
m <- mofa_df %>%
  distinct(sample, view) %>%
  mutate(present = TRUE) %>%
  pivot_wider(
    id_cols = sample,
    names_from = view,
    values_from = present,
    values_fill = FALSE
  )

m <- m %>%
  mutate(
    n_views = rowSums(across(-c("sample", "Functional"))),
    at_least_two = n_views >= 2,
    all_BL = !(grepl("_3M|_1W|_6M", .$sample)),
    all_BL = ifelse(grepl("S84_BL", sample), FALSE, all_BL),
    matched = n_views == 3
  ) %>%
  filter(!(grepl("KX63|KX64", .$sample)))

write_csv(m, "../Results/Sample_omic_table.csv")
# m <- read_csv("../Results/Sample_omic_table.csv")

knitr::kable(head(m))
```

Based on the various criteria we filter out the samples we do not wish to include from our long format table
```{r select-samples}
mofa_df <- mofa_df %>%
  filter(sample %in% m$sample[m$at_least_two & m$all_BL])

meta_ <- meta %>% filter(sample %in% m$sample[m$at_least_two & m$all_BL])
```

## 4. Sample heatmap overview
With our samples selected we can now plot a overviw heatmap showing all our included samples, which omic views they are represented in and how they are distributed by some of our groups of intrest
NB! S10 and S15 was excluded from metagenome analysis due to low total counts before and after taxa filtering respectivelyl
Thats why I have given them a lighter pink colour
```{r heatmap-function}
#| code-fold: true
heatmap_prepp <- function(keep_s, vars, arrange_by = "Clinical_score_(0-5)") {
  ### ------------------------------------------------------------
  ### 1. Select symptom columns and prepare binary clinical data
  ### ------------------------------------------------------------

  # Extract only samples of interest and symptom score columns
  # Convert TRUE/FALSE columns to numeric 0/1 and clean column names
  df_bin <- m %>%
    #filter(sample %in% m$sample[m$at_least_two & m$all_BL]) %>%
    filter(sample %in% m$sample[keep_s]) %>%

    left_join(
      .,
      select(meta, sample, RVVC_AS, RVVC_pos, "Clinical_score_(0-5)"),
      by = "sample"
    ) %>%
    mutate(
      RVVC_AS = factor(RVVC_AS, levels = c("RVVC", "Control", "RVVC_AS"))
    ) %>%
    mutate(
      RVVC_pos = factor(
        RVVC_pos,
        levels = c(
          "RVVC_pos",
          "RVVC_neg",
          "Control_pos",
          "Control_neg"
          #"NA_pos",
          #"NA_neg"
        )
      )
    ) %>%
    arrange(desc(n_views)) %>%
    arrange(RVVC_pos) %>%
    arrange(`Clinical_score_(0-5)`) %>%
    arrange(
      desc(.data[[arrange_by]])
    ) %>%
    select(1:5) %>%
    mutate(across(where(is.logical), as.double)) %>%
    mutate(
      Metagenomics = ifelse(grepl("S10|S15", sample), 0.5, Metagenomics),
      Functional = ifelse(grepl("S10|S15", sample), 0.5, Functional)
    ) %>%
    column_to_rownames("sample") # convert sample ID to rownames

  # Convert to numeric matrix for ComplexHeatmap
  mat <- as.matrix(df_bin)

  # Determine row annotation height scaling factor
  my_h <<- 0.1 * nrow(mat)

  ### ------------------------------------------------------------
  ### 2. Prepare annotation dataframe (metadata)
  ### ------------------------------------------------------------

  annot_col <<- meta %>%
    select(sample, any_of(vars)) %>% # select annotation variables
    mutate(across(all_of(vars), ~ factor(.x))) %>%
    # {. ->> all_annot} %>%
    filter(sample %in% rownames(df_bin)) %>% # keep same samples as heatmap matrix
    arrange(match(sample, rownames(df_bin))) %>%

    #mutate(across(1:length(vars) + 1, ~ factor(.x))) %>% # ensure annotations are factors
    column_to_rownames("sample") # convert sample ID to rownames

  # Quick check: ensure annotation rows match heatmap rows
  dim(mat)
  dim(annot_col)

  setdiff(rownames(mat), rownames(annot_col))

  ### ------------------------------------------------------------
  ### 4. Define row annotations for heatmap (ComplexHeatmap)
  ### ------------------------------------------------------------

  annot <- HeatmapAnnotation(
    df = annot_col[, c(vars)],
    which = c("row"),
    #which = c("column"),
    # Clin_gr = annot_col$Clin_gr, # binned clinical grade
    # "Clinical_score_(0-5)" = annot_col$`Clinical_score_(0-5)`, # raw clinical score
    # RVVC_AS = annot_col$RVVC_AS, # sample group
    # BV = annot_col$BV, # sample group
    # RVVC = annot_col$RVVC, # sample group
    # hyphae = annot_col$hyphae, # sample group

    show_annotation_name = TRUE, # show labels
    annotation_name_gp = gpar(fontsize = 8), # label font size

    # Legend formatting
    annotation_legend_param = list(
      grid_height = unit(.1, "mm"),
      grid_width = unit(2, "mm"),
      title = "",
      labels_gp = gpar(fontsize = 7),
      title_gp = gpar(fontsize = 8)
    ),

    simple_anno_size = unit(.3, "cm"), # annotation row height

    # Colors for each annotation variable
    na_col = "white",
    col = pal[vars] %>% imap(., ~ set_names(.x, levels(annot_col[[.y]])))
  )
  ### ------------------------------------------------------------
  ### Create heatmap
  ### ------------------------------------------------------------
  H <- Heatmap(
    mat,
    width = unit(ncol(mat) * 5, "mm"),
    height = unit(nrow(mat) * 2.2, "mm"),
    name = "Symptom Present",
    col = col_fun,

    # vertical orientation:
    right_annotation = annot,
    cluster_rows = FALSE,
    row_order = rownames(mat),

    row_names_side = "right",

    column_names_rot = 90,
    row_names_gp = gpar(fontsize = 7),
    heatmap_legend_param = list(
      at = c(0, 1),
      labels = c("No", "Yes")
    ),
    #   cell_fun = function(j, i, x, y, width, height, fill) {
    #   grid.rect(x, y, width, height, gp = gpar(col = "grey70", lwd = 0.5, fill = NA))
    # }
    layer_fun = function(j, i, x, y, width, height, fill) {
      # Draw cell borders without overwriting fill
      grid.rect(
        x = x,
        y = y,
        width = width,
        height = height,
        gp = gpar(col = "grey70", lwd = 0.4, fill = NA)
      )
    }
  )
  #return(list(mat = mat, annot = annot, h = my_h))
  return(H)
}
```

```{r function-top-plot-many-sampleHeatmaps}
#| code-fold: true
run_heatmap <- function(name, cfg) {
  H <- heatmap_prepp(
    cfg$keep_s,
    cfg$vars,
    arrange_by = cfg$arrange_by %||% "Clinical_score_(0-5)"
  )

  H_g <- draw(
    H,
    heatmap_legend_side = "bottom",
    merge_legend = FALSE,
    background = NA,
    show_annotation_legend = FALSE,
    legend_gap = unit(5, "mm"),
    padding = unit(c(3, 9, -2, 3), "mm")
  )

  file <- paste0("../Results/Sample_HT/", name, ".png")

  png(
    file = file,
    units = "in",
    res = 600,
    width = 8.5,
    height = (sum(cfg$keep_s) * 0.1) + 3
  )

  print(H_g)
  dev.off()

  message("Saved: ", file)
}
```

```{r Sample-info-heatmap}
#| code-fold: show
#| output: true
#| fig-width: 8.5
#| fig-height: 11
### ------------------------------------------------------------
### Color function for the symptom heatmap
### ------------------------------------------------------------
col_fun <- colorRamp2(
  c(0, 0.5, 1),
  c("white", "#EFAEC2", "#e05e85") #"#e05e8584"
)

### ------------------------------------------------------------
### Prepare annotation df and matrix
### ------------------------------------------------------------

vars <- c(
  "RVVC_pos",
  "Clinical_score_(0-5)",
  "RVVC",
  #"RVVC_AS",
  "BV",
  "NAC",
  "missing_page",
  "ST"
)

scenarios <- list(
  # 152 every samples incl. follow up
  all_samples = list(
    keep_s = rep(TRUE, nrow(m)),
    vars = vars
  ),
  # 91 samples, "all baseline"
  all_BL = list(
    keep_s = m$all_BL,
    vars = vars
  ),
  # 70 samples "matched"
  matched = list(
    keep_s = m$matched,
    vars = vars
  ),
  # exluded samples by Margaux: BV (n=5), NAC (n=4)
  excluded_Margaux = list(
    keep_s = grepl("S84_BL|S38|S44|S60|S20|S30|S16|S14|S49", m$sample),
    vars = vars
  ),
  # exluded samples by me: Missing meta/<2 omics
  excluded_MOFA = list(
    keep_s = grepl("S84_BL|S16|S14", m$sample),
    vars = vars
  ),
  # 88 samples
  filtered_combo = list(
    keep_s = m$at_least_two & m$all_BL,
    vars = c("RVVC_pos", "Clinical_score_(0-5)"),
    arrange_by = "RVVC_pos"
  )
)

walk2(names(scenarios), scenarios, ~ run_heatmap(.x, .y))

# quartz(width = 6, height = 6, dpi = 150)
H <- heatmap_prepp(
  m$at_least_two & m$all_BL,
  c("RVVC_pos", "Clinical_score_(0-5)"),
  arrange_by = "RVVC_pos"
)
draw(H)
```


## 5. Unsupervised UMAPs 
We plot unsupervised umaps coloured by different groups of intrest in order to see how they cluster across views. 

!NB should think about doing with and without HVGs only

```{r filter-samples-for-UMAP}
# filter samples from views for UMAP ploting
m_ <- m %>%
  filter(sample %in% .$sample[m$at_least_two & m$all_BL])

v <- views %>%
  imap(., ~ .x[, m_$sample[m_[[.y]]]])
```

This first code is taken from Paulos analysis of the metabolomics data.
It gives more control over the knn clustering by perfoming it in a seperate step and then supplying the knn to the umap function.
The other alternative bellow is just supplying the data to umap, which then runs the PCA and knn internaly with default setings. 
```{r Paulos-UMAP}
#| eval: false
#| code-fold: true
#### UMAP analysis ####
set.seed(1)
UMAP_df <- v %>%
  map(., ~ t(.x)) %>%
  # PCA
  map(., ~ prcomp(.x, scale = T, center = T)) %>%
  # KNN graph
  map(
    .,
    ~ RcppHNSW::hnsw_knn(as.matrix(.x$x[, 1:10]), k = 10, distance = "cosine")
  ) %>%

  # UMAP
  map(
    .,
    ~ uwot::umap(
      X = NULL,
      nn_method = .x,
      n_components = 2,
      ret_extra = c("model", "fgraph"),
      verbose = T,
      min_dist = 0.1,
      spread = 0.3,
      approx_pow = TRUE,
      repulsion_strength = .2,
      negative_sample_rate = 30,
      n_epochs = 200,
      n_threads = 8
    )$embedding
  ) %>%
  imap(
    .,
    ~ tibble(
      "UMAP 1" = .x[, 1],
      "UMAP 2" = .x[, 2],
      "ID" = colnames(v[[.y]])
    )
  ) %>%
  map(., ~ left_join(., annot_col, by = c("ID" = "sample")))
```

```{r Paulos-groups}
#| eval: false
# attempt at getting the same groups as Paulo, but failed, think I need his list with the group result
Paulo_clus <- c(
  Metab = "/Users/vilkal/Downloads/resulsts_metabolomics/metab_knn.rds",
  Microbe = "/Users/vilkal/Downloads/resulsts_metabolomics/microb_knn.rds",
  Integ = "/Users/vilkal/Downloads/resulsts_metabolomics/integ_knn.rds"
)

res <- c(0.8, 0.6, 1.3)

set.seed(42)
clus <- map(Paulo_clus, ~ readRDS(.x)) %>%
  # graph:
  map(
    .,
    ~ igraph::graph_from_data_frame(
      data.frame(
        from = rep(1:nrow(.x[["idx"]]), ncol(.x[["idx"]])),
        to = c(.x[["idx"]]),
        weight = c(.x[["dist"]])
      ),
      directed = F
    )
  ) %>%
  # clustering:
  map2(., res, ~ igraph::cluster_louvain(.x, resolution = .y, weights = NA))

map(clus, ~ table(.x$membership))
# it does not create the same groups as the seed are ignored as fare as I can see
```

```{r get-Louvain-clusters}
#| warning: false
#| message: false
cluster_louvain_list <- function(
  mat_list,
  k = 10,
  n_pcs = 10,
  resolution = 0.6,
  seed = 42
) {
  require(RcppHNSW)
  require(Matrix)
  require(igraph)
  require(tibble)

  cluster_results <- purrr::imap(
    mat_list,
    ~ {
      mg <- .x # matrix
      nm <- .y # name of the matrix
      idx <- which(names(mat_list) == nm)

      # PCA
      PC <- prcomp(t(mg), scale. = TRUE, center = TRUE)
      x <- PC$x[, seq_len(min(n_pcs, ncol(PC$x)))]

      # KNN
      knn <- RcppHNSW::hnsw_knn(as.matrix(x), k = k, distance = "cosine")

      # Graph
      g <- igraph::graph_from_data_frame(
        data.frame(
          from = rep(1:nrow(knn$idx), ncol(knn$idx)),
          to = c(knn$idx),
          weight = c(knn$dist)
        ),
        directed = FALSE
      )

      # Clustering
      set.seed(seed)
      cl <- igraph::cluster_louvain(
        g,
        resolution = resolution[idx],
        weights = NA
      )
      #cl <- igraph::cluster_leiden(g, resolution = resolution)

      out <- cl$membership
      names(out) <- rownames(x)

      out
    }
  )

  # Combine into one table
  cluster_df <- cluster_results %>%
    imap(
      ~ tibble(
        sample = names(.x),
        cluster = .x,
        matrix = .y
      )
    ) %>%
    bind_rows() %>%
    pivot_wider(
      names_from = matrix,
      values_from = cluster
    ) %>%
    mutate(across(-sample, as.factor))

  return(cluster_df)
}
# res <- c(1.3, 1.2, 1.3, 0.9)
res <- c(1.3, 1.2, 1.3, 1)
n <- c("trx_louvain_7", "metab_louvain", "microb_louvain", "fun_louvain")
result <- v %>%
  set_names(., n) %>%
  cluster_louvain_list(mat_list = ., resolution = res)

map(set_names(n), ~ table(result[[.x]]))
# head(result)

meta <- meta %>%
  select(-any_of(n)) %>%
  left_join(., result, by = "sample") %>%
  select(1:15, any_of(n), everything())
# write_csv(meta, "../Results/Combined_gr_meta_data.csv")
```

```{r UMAP-function}
#| code-fold: true
UMAP_fun <- function(
  umap_df,
  group_var,
  txt_var = NULL,
  shape_gr = NULL,
  shape = NULL,
  title = ""
) {
  if (is.null(shape_gr)) {
    point <- geom_point(size = 2, alpha = 1, shape = 21)
  } else {
    point <- geom_point(
      size = 2,
      alpha = 1,
      aes(shape = if (!is.null(shape_gr)) .data[[shape_gr]] else NULL)
    )
  }

  p <- ggplot(
    umap_df,
    aes(
      x = `UMAP 1`,
      y = `UMAP 2`,
      #shape = RVVC_AS,
      fill = .data[[group_var]],
      color = .data[[group_var]]
    )
  ) +
    point +
    #geom_jitter( shape=21, size=3, color="white", width=.5, height=.5) +  # Tassos used jitter

    scale_fill_manual(
      values = pal[[group_var]],
      na.value = "transparent",
      name = ""
      #aesthetics = c("colour", "fill")
    ) +
    guides(fill = "none", nrow = 2, byrow = TRUE) +
    scale_colour_manual(
      values = pal[[group_var]],
      na.value = "gray",
      name = ""
    ) +
    {
      if (!is.null(shape_gr) && !is.null(shape)) {
        scale_shape_manual(values = shape, name = "")
      }
    } +
    # scale_shape_manual(values = c(21, 5, 4), name = "" ) + #specify individual shapes+
    {
      if (!is.null(txt_var)) {
        geom_text(
          data = umap_df,
          aes(x = `UMAP 1`, y = `UMAP 2`, label = .data[[txt_var]]),
          size = 2,
          hjust = 0,
          nudge_x = 0.07,
          color = "gray51"
        )
      }
    } +
    ggtitle(title) +
    theme_bw(base_size = 14) +
    theme(
      # point = element_point(shape = 21),
      line = element_blank(),
      axis.ticks = element_line(color = "black"),
      aspect.ratio = 1,
      axis.title = element_blank()
    )
  return(p)
}

#### UMAP plotting by groups ####

# save plot to file
#ggsave(paste0("./Figures/09/", "Bulk_UMAP.png"), p)
```

```{r plot-UMAP}
#| fig-width: 14
#| fig-height: 5
#| code-fold: true
#| warning: false
#### group information ####
annot_col <- meta %>%
  #filter(sample %in% .$sample[m$at_least_two & m$all_BL]) %>%
  # select(-`Age (years)`) %>%
  select(
    1:22,
    `Clinical_score_(0-5)`
  ) %>%
  mutate(across(2:ncol(.), ~ factor(.x))) %>%
  mutate(
    txt_clin = ifelse(
      grepl("0", .$`Clinical_score_(0-5)`),
      NA,
      as.character(.$`Clinical_score_(0-5)`)
    )
  ) %>%
  mutate(txt_id = ifelse(RVVC == "RVVC", .$sample, NA))

#### UMAP analysis ####
set.seed(1)

umap_dfs <- v %>%
  map(., ~ t(.x)) %>%
  map(
    .,
    ~ uwot::umap(
      .x,
      #seed = 123,
      seed = 463,
      n_neighbors = 15,
      #metric = "cosine",
      init = "spectral",
      scale = T
    )
  ) %>%
  map(
    .,
    ~ tibble(
      "UMAP 1" = .x[, 1],
      "UMAP 2" = .x[, 2],
      "ID" = rownames(.x)
    )
  ) %>%
  map(., ~ left_join(., annot_col, by = c("ID" = "sample")))


meta_clus <- colnames(annot_col)
color_vars <- c(
  "pos",
  "trx_louvain_7",
  "metab_louvain",
  "microb_louvain",
  "fun_louvain",
  #"integ_louvain",
  "Clinical_score_(0-5)",
  "RVVC_pos",
  "CST"
) %>%
  set_names(.)

#### Combine and save plots ####
make_umap_panel <- function(var, umap_dfs, outdir = "../Results/UMAP2/") {
  # create plots for each dataset
  p_list <- imap(
    umap_dfs,
    ~ UMAP_fun(.x, var, "txt_clin", title = .y)
  )

  # combine with shared legend
  p_combined <- patchwork::wrap_plots(p_list) +
    patchwork::plot_layout(nrow = 1, guides = "collect") &
    theme(legend.position = "right")

  # safe filename
  fname <- gsub("[^A-Za-z0-9_]", "_", var)

  # save
  ggsave(
    filename = paste0(outdir, "UMAP_", fname, ".png"),
    plot = p_combined,
    width = 14,
    height = 5,
    dpi = 300
  )

  return(p_combined)
}

plots <- map(color_vars, ~ make_umap_panel(.x, umap_dfs))
plots

# quartz(width = 6, height = 6, dpi = 60)
p_ <- patchwork::wrap_plots(plots, ncol = 1) +
  patchwork::plot_layout(guides = "collect", axis_titles = "collect") &
  theme(legend.position = "right", title = element_blank())

ggsave(
  paste0("../Results/Factor_plots2/UMAP_louvain", ".pdf"),
  p_,
  width = 15,
  height = 25
)
# use shapes:
# UMAP_fun(umap_dfs$Metabolomics, "trx_louvain_7", "txt_clin", shape_gr = "pos", shape = c(21, 24))
# no shapes
# UMAP_fun(umap_dfs$Metabolomics, "trx_louvain_7", "BV")
```


## 6. MOFA model construction
A MOFA object is created and configured with default data and training options. The number of latent factors is set a priori and can be tuned depending on variance explained.

### Data options

<details>
<summary> Important arguments</summary>

- **`scale_groups`**  
  Scale groups to the same total variance?  
  *Default:* `FALSE`

- **`scale_views`**  
  Scale views to the same total variance?  
  *Default:* `FALSE`

- **`views`**  
  Names of the views to include in the model.

- **`groups`**  
  Names of the sample groups.
</details>

### Model options

<details>
<summary> Important arguments</summary>

- **`num_factors`**  
  Number of latent factors.

- **`likelihoods`**  
  Likelihood per view.  
  Options: `"gaussian"`, `"poisson"`, `"bernoulli"`  
  By default, likelihoods are inferred automatically.

- **`spikeslab_factors`**  
  Use spike-and-slab sparsity prior on the factors?  
  *Default:* `FALSE`

- **`spikeslab_weights`**  
  Use spike-and-slab sparsity prior on the weights?  
  *Default:* `TRUE`

- **`ard_factors`**  
  Use Automatic Relevance Determination (ARD) prior on the factors?  
  *Default:* `TRUE` when using multiple groups.

- **`ard_weights`**  
  Use Automatic Relevance Determination (ARD) prior on the weights?  
  *Default:* `TRUE` when using multiple views.
</details>

### Training options

<details>
<summary> Important arguments</summary>

- **`maxiter`**  
  Maximum number of training iterations.  
  *Default:* `1000`

- **`convergence_mode`**  
  Convergence speed of the algorithm.  
  Options: `"fast"`, `"medium"` (default), `"slow"`  
  For exploratory analyses, `"fast"` is usually sufficient.

- **`seed`**  
  Random seed for reproducibility.
</details>


We use default options, however it is recomended to run the model several times with different seeds. In one of the papers published by MOFAS authors, they ran the model ~24 times, before picking the model with the highest elbo score.
```{r mofa_setup}
MOFAobject <- create_mofa_from_df(mofa_df)

data_opts <- get_default_data_options(MOFAobject)
model_opts <- get_default_model_options(MOFAobject)
train_opts <- get_default_training_options(MOFAobject)

#model_opts$num_factors <- 10
train_opts$convergence_mode <- "medium"
train_opts$seed <- 1234
```


The functions bellow runs the model with different seeds and saves the model and a table with all seeds and corresponding elbo scores as output
```{r train-mofa-model}
#| code-fold: true
#| eval: false
#| output: false
#| code-summary: "Train MOFA model"
#| warning: false
#| message: false
seeds <- c(3, 100, 50)

run_mofa_with_seed <- function(seed) {
  # Make a local copy (important!)
  train_opts$seed <- seed

  MOFAobject <- prepare_mofa(
    MOFAobject,
    data_options = data_opts,
    model_options = model_opts,
    training_options = train_opts
  )

  MOFAmodel <- run_mofa(
    MOFAobject,
    outfile = paste0("../Results/MOFA/MOFA_allBL_noS84_seed_", seed, ".hdf5"),
    use_basilisk = TRUE
  )

  elbo <- max(get_elbo(MOFAmodel))

  # Return a single-row data frame
  tibble(
    seed = seed,
    elbo = elbo
  )
}

# existing training code stays the same
elbo_table <- map_dfr(seeds, ~ run_mofa_with_seed(.x))
write_csv(elbo_table, "../Results/MOFA/elbo_table_noS84.csv")
best_seed <- elbo_table$seed[which.max(elbo_table$elbo)]
```

MOFAobject without transcriptomics batch correction
```{r model-noBatch}
#| eval: false
X <- readRDS("../Results/MixOmic/X_object.RDS")
views <- list(
  Transcriptomics = t(X$strain$TR),
  Metagenomics = t(X$strain$MG),
  Metabolomics = t(X$species$MB)
)

MOFAobject <- create_mofa(views)

data_opts <- get_default_data_options(MOFAobject)
model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 10

train_opts <- get_default_training_options(MOFAobject)
train_opts$convergence_mode <- "medium"
train_opts$seed <- 1234

MOFAobject <- prepare_mofa(
  MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)
MOFAmodel <- run_mofa(
  MOFAobject,
  outfile = "../Results/MOFA/MOFA_noBatch_strain.hdf5",
  use_basilisk = TRUE
)
```

## 7. Load trained MOFA model
We load the model with the highest elbo score
```{r load_model}
#| eval: true
#| include: true
#| warning: false
#| message: false
#| tbl-colwidths: [60,40]
elbo_table <- read_csv("../Results/MOFA/elbo_table.csv")
elbo_table <- read_csv("../Results/MOFA/elbo_table_noS84.csv")
knitr::kable(elbo_table, )
best_seed <- elbo_table$seed[which.max(abs(elbo_table$elbo))]

MOFAmodel <- load_model(
  #"../Results/MOFA/MOFA_matched_seed_50.hdf5"
  # "../Results/MOFA/MOFA_allBL_seed_"
  paste0("../Results/MOFA/MOFA_allBL_noS84_seed_", best_seed, ".hdf5")
)

# add meta data to model
samples_metadata(MOFAmodel) <- samples_metadata(MOFAmodel)[, c(
  "sample",
  "group"
)] %>%
  left_join(meta, by = c("sample"))

# batch corrected tr, ~17 000 genes, strain lvl, no followup samples
"../Results/MOFA/MOFA_matched_noM3_strain.hdf5"
# batch corrected tr, ~17 000 genes, strain lvl + 7 M3 follow up samples
"../Results/MOFA/MOFA_w73M_strain.hdf5"
# same as above + 9 extra RVVC samples without matching trx
"../Results/MOFA/MOFA_w9RVVC_strain.hdf5"

# same as above + all 154 samples
"../Results/MOFA/MOFA_all_154_samples_strain.hdf5"
```

```{r get-factor-clusters}
#| eval: false
# Extract factors, and join metadata

clusters <- list()
clusters$fct1_clus <- cluster_samples(MOFAmodel, k = 4, factors = 1)$cluster
clusters$fct2_clus <- cluster_samples(MOFAmodel, k = 4, factors = 2)$cluster
clusters$fct4_clus <- cluster_samples(MOFAmodel, k = 4, factors = 4)$cluster

clust <- bind_cols(sample = names(clusters[[1]]), clusters)

meta <- meta %>%
  select(-any_of(c("fct1_clus", "fct2_clus", "fct3_clus", "fct4_clus"))) %>%
  dplyr::left_join(., clust, by = "sample") %>%
  dplyr::mutate(across(all_of(names(clusters)), ~ factor(.x)))
# write_csv(meta, "../Results/Combined_gr_meta_data.csv")

# add meta data to model
samples_metadata(MOFAmodel) <- samples_metadata(MOFAmodel)[, c(
  "sample",
  "group"
)] %>%
  left_join(meta, by = c("sample"))
```

## 8. Model diagnostics

```{r view-distribution}
#| fig-width: 8
#| fig-height: 6
# inspect data distribution
ggdensity(mofa_df, x = "value", fill = "gray70") +
  facet_wrap(~view, nrow = 3, scales = "free") # + xlim(0.5,)
```

We first assess the overall quality of the model by inspecting variance explained and factor correlations.
```{r model-overview}
#| fig-width: 6
#| fig-height: 3
# inspect the model:
ovw <- plot_data_overview(MOFAmodel) +
  theme(plot.margin = unit(c(1, 0, 1, 1), "cm")) + #  #EC645A, #CD3B8E, #994E95
  scale_fill_manual(
    values = c("#CC79A7", "#D55E00", "#CD3B8E", "gray", "#E69F00"),
    na.value = "gray"
  )
ovw
MOFAmodel
```

```{r variance_explained}
#| warning: false
#| message: false
#| fig-width: 9.5
#| fig-height: 3
p1 <- plot_variance_explained(MOFAmodel, plot_total = TRUE)[[2]]
p2 <- plot_variance_explained(MOFAmodel, factors = 1:10)

#+
#theme(axis.text.x = element_text(angle = 45, hjust = 0.8, vjust = 0.9))
# plot_grid(plotlist = list(p1, p2), ncol = 2)

p1$layers[[1]]$aes_params$fill <- "gray" # remove fixed fill
p1$layers[[1]]$aes_params$colour <- "#717272" # remove fixed fill

p1 <- p1 +
  coord_flip() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    plot.margin = unit(c(0, 0.1, 0, -0.5), "cm")
  )
p2 <- p2 +
  coord_flip() +
  theme(
    legend.position = c(0.5, 1.09), # 👈 just above panel, tight
    legend.direction = "horizontal",
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 6),
    legend.text.position = "top",
    legend.margin = margin(0, 0, -0, 0), # moves the legend box
    legend.box.margin = margin(-0, 0, -0, -0)
  ) + # moves the legend) +
  guides(fill = guide_colorbar(title = "", barwidth = 12, barheight = 0.3)) +
  theme(axis.text.y = element_blank()) +
  theme(plot.margin = unit(c(0, 0.5, 0, 0), "cm")) +
  scale_y_discrete(labels = 1:10) +
  scale_fill_gradientn(
    colors = c("gray97", "darkblue"),
    labels = scales::label_number(suffix = "%") # 👈 THIS is the key
  )

# quartz(width = 9.5, height = 3)
plot_grid(plotlist = list(ovw, p2, p1), nrow = 1, align = c("h"))

# p$layers[[1]]$aes_params$alpha
```

```{r variance-by-factor}
ve_df <- get_variance_explained(MOFAmodel)$r2_per_factor$single_group %>%
  as.data.frame() %>%
  rownames_to_column("Factor") %>%
  pivot_longer(
    -Factor,
    names_to = "view",
    values_to = "variance_explained"
  ) %>%
  mutate(Factor = factor(.$Factor, levels = paste0("Factor", 1:15)))

d <- ve_df %>%
  filter(grepl("^Factor1$", .$Factor)) %>%
  #filter(grepl("^Factor[1-9]$", .$Factor)) %>%
  summarize(sum = sum(variance_explained), .by = c("Factor", "view")) %>%
  filter(sum >= 2)

ggplot(ve_df, aes(x = Factor, y = variance_explained, fill = view)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    values = c("#CC79A7", "#D55E00", "#CF1C90", "#E69F00"),
    na.value = "gray"
  ) +
  labs(
    x = "Factor",
    y = "Variance explained (%)",
    fill = "view"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

#### Correlation between factors
A good sanity check is to verify that the Factors are largely uncorrelated. In MOFA there are no orthogonality constraints such as in Principal Component Analysis, but if there is a lot of correlation between Factors this suggests a poor model fit. Reasons could be that you used too many factors or perhaps the normalisation is not adequate.

```{r factor_correlations}
plot_factor_cor(MOFAmodel) # , col = RdBu
```

```{r factors_by_batch}
MOFAbatch <- load_model(
  "../Results/MOFA/MOFA_noBatch_strain.hdf5"
)
samples_metadata(MOFAbatch) <- samples_metadata(MOFAbatch) %>%
  left_join(meta, by = "sample")

plot_factors(
  MOFAbatch,
  factors = "all",
  color_by = "batch"
)
```

When running MOFA the first time, the internal quality control warned that at least one of the factors was strongly corelated with the total number of expressed features in one of the views. This is typically a result of batch effect. As seen in the plot above Factor 1 perfectly seperates the two batches of the transcriptomics data. Thus batch correction was performe in the upstream `QC_Transcriptomics.qmd` script.

```{r factors_by_clinical}
#| fig-width: 10
#| fig-height: 5
p <- plot_factors(
  MOFAmodel,
  factors = "all",
  color_by = "Clinical_score_(0-5)"
) +
  scale_fill_manual(values = pal$`Clinical_score_(0-5)`, na.value = "white")

p

p <- plot_factor(
  MOFAmodel,
  factors = "all",
  color_by = "RVVC_pos",
  group_by = "RVVC_pos",
) +
  scale_fill_manual(values = pal$RVVC_pos, na.value = "white")

p

plot_factors(
  MOFAmodel,
  factors = c(1, 3),
  color_by = "Clin_gr"
) +
  stat_ellipse(
    #aes(color = RVVC_AS, fill = RVVC_AS),
    geom = "polygon",
    alpha = 0.25
  ) +
  scale_fill_manual(values = pal$Clin_gr, na.value = "white")
```

```{r factor-with-distribution}
library(ggExtra)
p <- get_factors(MOFAmodel, factors = c(1, 3), as.data.frame = TRUE) %>%
  as_tibble() %>%
  select(-group) %>%
  pivot_wider(names_from = "factor") %>%
  left_join(., samples_metadata(MOFAmodel), by = "sample") %>%
  ggplot(., aes(x = Factor1, y = Factor3, colour = RVVC_pos)) +
  geom_point(size = 2, alpha = 0.8, show.legend = F) +
  scale_colour_manual(values = pal$RVVC_pos) +
  coord_cartesian(expand = T) +
  # scale_x_continuous(expand = expansion(mult = 0.02)) +
  # scale_y_continuous(expand = expansion(mult = 0.02)) +
  theme_classic()

ggMarginal(
  p,
  type = "density",
  groupColour = TRUE,
  trim = FALSE,
  groupFill = TRUE
) +
  stat_ellipse(
    #aes(color = RVVC_AS, fill = RVVC_AS),
    geom = "polygon",
    alpha = 0.25
  )


plot_mofa_factor_vs_variable <- function(
  df,
  factor = 1,
  y_var,
  palette = pal[[y_var]],
  jitter_width = 0.1,
  add_marginal = TRUE,
  marginal_type = "density"
) {
  # jitter object
  jitterer <- position_jitter(width = jitter_width, seed = 123)

  df_ <- df %>% mutate(!!sym(y_var) := factor(!!sym(y_var)))

  # Main ggplot
  p <- ggplot(
    df_,
    aes(
      x = .data[[paste0("Factor", factor)]],
      y = .data[[y_var]],
      colour = .data[[y_var]]
    )
  ) +
    geom_point(
      size = 1,
      alpha = 0.8,
      show.legend = FALSE,
      position = jitterer
    ) +
    geom_point(
      aes(colour = NULL),
      pch = 21,
      size = 1,
      alpha = 0.4,
      show.legend = FALSE,
      colour = "gray",
      position = jitterer
    ) +
    geom_boxplot(
      aes(colour = .data[[y_var]]),
      fill = "transparent",
      show.legend = FALSE,
      outlier.shape = NA,
      median.color = "black"
    ) +
    coord_cartesian(expand = TRUE) +
    theme_minimal() +
    theme(
      axis.text.y.left = element_blank(),
      axis.title.y.left = element_blank()
    )

  # Add palette if provided
  if (!is.null(palette)) {
    p <- p + scale_colour_manual(values = palette)
  }

  # Add marginal plot if requested
  if (add_marginal) {
    p <- ggMarginal(
      p,
      type = marginal_type,
      groupColour = TRUE,
      groupFill = TRUE,
      trim = FALSE,
      margins = "x",
      size = 1
    )
  }

  return(p)
}

# Extract factors, reshape wide, join metadata
df <- get_factors(MOFAmodel, factors = c(1:4), as.data.frame = TRUE) %>%
  as_tibble() %>%
  select(-group) %>%
  #mutate(fct1_clus = as.character(clusters$cluster[.$sample])) %>%
  pivot_wider(names_from = "factor") %>%
  left_join(samples_metadata(MOFAmodel), by = "sample")
```

```{r}
#| fig.width: 3
#| fig.height: 2
# quartz(width = 3, height = 2, dpi = 150)
plot_mofa_factor_vs_variable(
  df,
  y_var = "RVVC_pos"
)
```

```{r}
#| fig.width: 3
#| fig.height: 2
plot_mofa_factor_vs_variable(
  df,
  y_var = "Clinical_score_(0-5)"
)
```

```{r}
#| fig.width: 3
#| fig.height: 2
plot_mofa_factor_vs_variable(
  df,
  y_var = "fct1_clus"
)
```

```{r}
#| fig.width: 3
#| fig.height: 2
plot_mofa_factor_vs_variable(
  df,
  factor = 4,
  palette = NULL,
  y_var = "fct4_clus"
)
```

```{r save-factor-violin-all-gr}
color_vars <- c(
  "pos",
  "trx_louvain_7",
  "metab_louvain",
  "microb_louvain",
  #"fun_louvain",
  "integ_louvain",
  "Clinical_score_(0-5)",
  "RVVC_pos",
  "RVVC_AS"
) %>%
  set_names(.)

# fct_var <- map(
#   vars,
#   ~ plot_factors(
#     MOFAmodel,
#     factors = "all",
#     color_by = .x,
#   ) +
#     scale_fill_manual(values = pal[[.x]], na.value = "white")
# )

plots <- imap(
  color_vars,
  ~ plot_factor(
    MOFAmodel,
    factor = "all",
    #group_by = .x,
    color_by = .x,
    #dot_size = 4,
    #dodge = TRUE,
    stroke = 0.4,
    #add_violin = TRUE,
    #add_boxplot = FALSE
  ) +
    ylim(-4, 5) +
    scale_fill_manual(values = pal[[.x]], na.value = "white")
)

ggsave(
  paste0("../Results/Factor_plots2/Factor_all_gr", ".pdf"),
  gridExtra::marrangeGrob(grobs = plots, ncol = 1, nrow = 1, height = 7)
)
```

## 9. Covariate-Factor correlation

```{r covariates-factor-correlation}
#| warning: false
#| message: false
#samples_metadata(MOFAmodel) <- samples_metadata(MOFAobject)
covariates <- samples_metadata(MOFAmodel)[,
  28:ncol(samples_metadata(MOFAmodel))
] %>%
  colnames()
groups <- samples_metadata(MOFAmodel)[, 1:23] %>% colnames()

# remove columns
# covariates <- covariates[!(grepl("Follow_up", covariates))]
Clin_info <- list(
  Clin_score = covariates[1:12], # Clinical score + Symptom score
  Wet_smear = covariates[13:18], # Wet smear
  Fungal_culture = covariates[19:26], # Fungal culture
  confounders = covariates[32:37], # general confounders
  relationship = covariates[38:42], # relationship
  drugs = covariates[43:60], # drugs
  reproductive_info = covariates[c(61:70, 74:76)], # reproductive info
  louvain = groups[13:20] # reproductive info
)

# reproductive info
# quartz(width = 6, height = 6, dpi = 100)
p <- imap(
  Clin_info,
  ~ wrap_elements(
    panel = ~ correlate_factors_with_covariates(
      MOFAmodel,
      col = colorRampPalette(RdBu)(100),
      covariates = .x,
      win.asp = 1, # max(6, max(nchar(.x)) * 0.18) / max(3, length(.x) * 0.35) ,
      plot = "r", # use "log_pval" to plot log p-values
      transpose = T
    )
  )
)
par(mar = c(0, 0, 0, 0), bg = NA)
# p[["reproductive_info"]]
p

imap(
  p,
  ~ {
    vars <- Clin_info[[.y]]

    h <- max(3, length(vars) * 0.35) # height per variable
    w <- max(6, max(nchar(vars)) * 0.18) # width per character

    ggsave(
      filename = paste0("../Results/Covar_cor/", "Fact_", .y, ".png"),
      plot = .x,
      width = w,
      height = h,
      dpi = 300
    )
  }
)
```

## 10. Factor visualisation

```{r Factor-grid-dist-plot}
#| fig-width: 7
#| fig-height: 5
p <- plot_factors(
  MOFAmodel,
  factors = c(1, 2, 3, 4),
  dot_size = 1,
  #stroke = 1,
  color_by = "RVVC_pos"
) +
  scale_fill_manual(
    values = pal$RVVC_pos,
    na.value = "transparent",
    aesthetics = c("colour", "fill")
  )
# quartz(width = 6, height = 6, dpi = 150)

p

ggsave(
  filename = paste0("../Results/Factor_plots2/", "Factor_1-3_grid", "", ".png"),
  plot = p,
  width = 7,
  height = 5,
  dpi = 300
)

### Plot and save many versions:
my_fun <- function(factors, color_var) {
  p <<- do.call(
    plot_factors,
    list(
      object = MOFAmodel, # ← correct argument name
      factors = factors,
      dot_size = 1,
      shape_by = NULL,
      color_by = color_var
    )
  )

  p +
    scale_fill_manual(
      values = pal[[color_var]],
      na.value = "transparent",
      aesthetics = c("colour", "fill")
    )
}

factor_list <- list(c(1, 2, 7), c(1, 8, 9), c(1, 3, 10), c(4, 5, 6), c(4, 5, 6))
color_vars <- c(
  "trx_louvain_7",
  "metab_louvain",
  "microb_louvain",
  "metab_louvain",
  "RVVC_AS"
)

plots <- purrr::map2(factor_list, color_vars, my_fun)
# quartz(width = 6, height = 6, dpi = 150)
# plots[[3]]

pdf(
  file = paste0("../Results/Factor_plots2/Grid_Factors", ".pdf"),
  #units = "in",
  #res = 600,
  #width = 8.5, height = 10 # Epi
  width = 6,
  height = 4 # 11 # my_h # 17 (all samples)
)
plots
#draw(H)
dev.off()
```

```{r Factor-violin-plots}
#| code-fold: true
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5
plot_factor_set <- function(
  MOFAmodel,
  factors,
  color_vars,
  pal,
  out_file,
  height = 5,
  width_per_plot = 2.8
) {
  plots <- imap(
    color_vars,
    ~ plot_factor(
      MOFAmodel,
      factor = factors,
      color_by = .x,
      dot_size = 2,
      dodge = TRUE,
      stroke = 0.4,
      add_violin = TRUE,
      add_boxplot = FALSE
    ) +
      scale_fill_manual(values = pal[[.x]], na.value = "white") +
      theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
      )
  )

  p <- patchwork::wrap_plots(plots, nrow = 1) +
    patchwork::plot_layout(guides = "collect", axis_titles = "collect") &
    theme(legend.position = "bottom")

  ggsave(
    filename = out_file,
    plot = p,
    width = width_per_plot * length(color_vars),
    height = height,
    dpi = 300
  )

  p
}

meta_clus <- colnames(annot_col)


fctors <- c(
  "RVVC_AS",
  "pos",
  "Clinical_score_(0-5)",
  "RVVC_pos",
  "metab_louvain"
) %>%
  set_names()

map(
  c(1, 2, 3, 4),
  ~ plot_factor_set(
    MOFAmodel = MOFAmodel,
    factors = .x,
    color_vars = fctors,
    pal = pal,
    out_file = paste0("../Results/Factor_plots2/Factor", .x, "_gr", ".png")
  )
)
c(1, 2, 3, 4, 5, 6, 7, 8)
c(9, 10, 11, 12, 13, 14, 15)

fctors <- c(
  "trx_louvain_7",
  "metab_louvain",
  "microb_louvain",
  "fun_louvain"
) %>%
  set_names()
update_geom_defaults("violin", aes(linewidth = 0))
update_geom_defaults("boxplot", aes(linewidth = 0))
p <- map(
  c(1, 2, 3, 4, 5, 6, 7, 8),
  ~ plot_factor_set(
    MOFAmodel = MOFAmodel,
    factors = .x,
    color_vars = fctors,
    pal = pal,
    out_file = paste0("../Results/Factor_plots2/Factor", .x, "_louvain", ".png")
  )
)

# quartz(width = 6, height = 6, dpi = 60)
p_ <- patchwork::wrap_plots(p, ncol = 1) +
  patchwork::plot_layout(guides = "collect", axis_titles = "collect") &
  theme(legend.position = "none")

ggsave(
  paste0("../Results/Factor_plots2/Factor_1-8_louvain_", ".png"),
  p_,
  width = 8,
  height = 10
)
# not working
# ggsave(paste0("../Results/Factor_plots/Factor_4-8_gr", ".pdf"),
#        gridExtra::marrangeGrob(grobs=p, ncol=1, nrow=1, height = 5, width=2)))
```

## 11. Feature-level analysis

```{r top_weights}
#| fig-cap: "Top metagenomic features driving Factor 2"
#| fig-width: 10
#| fig-height: 5
views <- c("Transcriptomics", "Metabolomics", "Metagenomics", "Functional")
c <- c("#CC79A7", "#D55E00", "#CF1C90", "#E69F00")
p1 <- map(
  views,
  ~ plot_weights(
    MOFAmodel,
    view = .x,
    factor = 1,
    nfeatures = 10,
    text_size = 3
  ) +
    theme_transparent()
)
p2 <- map(
  views,
  ~ plot_top_weights(
    MOFAmodel,
    factor = 1,
    view = .x,
    nfeatures = 10
  ) +
    theme_minimal()
)
plot_grid(plotlist = p2)
p <- map2(p1, p2, ~ plot_grid(plotlist = list(.x, .y))) %>% set_names(., views)
p
```

```{r modified-MOFA-heatmap-function}
#| code-fold: true
# Modified function: plot_data_heatmap2() with sample_order argument
# It is identical to the original except for 13 added lines.
plot_data_heatmap2 <- function(
  object,
  factor,
  view = 1,
  groups = "all",
  features = 50,
  max_val = NULL,
  annotation_features = NULL,
  annotation_samples = NULL,
  transpose = FALSE,
  imputed = FALSE,
  denoise = FALSE,
  max.value = NULL,
  min.value = NULL,
  sample_order = NULL, # ✅ NEW ARGUMENT
  ...
) {
  if (!is(object, "MOFA")) {
    stop("'object' has to be an instance of MOFA")
  }

  stopifnot(length(factor) == 1)
  stopifnot(length(view) == 1)

  # ✅ INTERNAL HELPERS MUST USE triple-colon
  groups <- MOFA2:::.check_and_get_groups(object, groups)
  factor <- MOFA2:::.check_and_get_factors(object, factor)
  view <- MOFA2:::.check_and_get_views(object, view)

  # Extract model matrices
  W <- do.call(
    rbind,
    get_weights(object, views = view, factors = factor, as.data.frame = FALSE)
  )
  Z_list <- lapply(get_factors(object)[groups], function(z) {
    as.matrix(z[, factor])
  })
  Z <- do.call(rbind, Z_list)[, 1]
  Z <- Z[!is.na(Z)]

  # Choose raw, imputed, or denoised data
  if (isTRUE(denoise)) {
    data <- predict(object, views = view, groups = groups)[[1]]
  } else if (isTRUE(imputed)) {
    data <- get_imputed_data(object, view, groups)[[1]]
  } else {
    data <- get_data(object, views = view, groups = groups)[[1]]
  }
  if (is(data, "list")) {
    data <- do.call(cbind, data)
  }

  # ----------------------
  # FEATURE SELECTION
  # ----------------------
  if (is.numeric(features)) {
    if (length(features) == 1) {
      features <- rownames(W)[tail(order(abs(W)), n = features)]
    } else {
      features <- rownames(W)[order(-abs(W))[features]]
    }
    features <- names(W[features, ])[order(W[features, ])]
  } else if (is.character(features)) {
    stopifnot(all(features %in% features_names(object)[[view]]))
  } else {
    stop("features must be numeric or character vector")
  }

  data <- data[features, ]
  data <- data[, names(Z), drop = FALSE]
  data <- data[, apply(data, 2, function(x) !all(is.na(x))), drop = FALSE]

  # ----------------------
  # ✅ CUSTOM SAMPLE ORDER
  # ----------------------
  if (is.null(sample_order)) {
    # default MOFA2 ordering
    order_samples <- names(sort(Z, decreasing = TRUE))
    order_samples <- order_samples[order_samples %in% colnames(data)]
  } else {
    # user‑specified ordering
    sample_order <- sample_order[sample_order %in% colnames(data)]
    if (!all(sample_order %in% colnames(data))) {
      stop("sample_order contains sample names not found in data")
    }
    order_samples <- sample_order
  }

  data <- data[, order_samples, drop = FALSE]

  # ----------------------
  # ANNOTATION HANDLING
  # ----------------------
  if (!is.null(annotation_samples)) {
    if (is.data.frame(annotation_samples)) {
      if (any(!order_samples %in% rownames(annotation_samples))) {
        stop("annotation_samples rownames must match sample names")
      }

      annotation_samples <- annotation_samples[order_samples, , drop = FALSE]
    } else if (is.character(annotation_samples)) {
      tmp <- object@samples_metadata
      rownames(tmp) <- tmp$sample
      tmp$sample <- NULL
      tmp <- tmp[order_samples, , drop = FALSE]
      annotation_samples <- tmp[, annotation_samples, drop = FALSE]
      rownames(annotation_samples) <- order_samples
    } else {
      stop("annotation_samples must be a data.frame or character vector")
    }

    # logical/character must become factor
    to_factor <- sapply(annotation_samples, function(x) {
      is.logical(x) || is.character(x)
    })
    if (any(to_factor)) {
      annotation_samples[, which(to_factor)] <-
        lapply(annotation_samples[, which(to_factor), drop = FALSE], as.factor)
    }
  }

  if (transpose) {
    data <- t(data)
    tmp_ann <- annotation_samples
    annotation_samples <- annotation_features
    annotation_features <- tmp_ann
  }

  # cap values
  if (!is.null(max.value)) {
    data[data >= max.value] <- max.value
  }
  if (!is.null(min.value)) {
    data[data <= min.value] <- min.value
  }

  # ----------------------
  # ✅ FINAL HEATMAP CALL
  # ----------------------
  #RdBu <- brewer.pal(11, "RdBu")
  # RdYlBu <- rev(brewer.pal(7, "RdYlBu")) # NB! 7 vs 11 makes a huge dfference, 7 gives smoother transition of colour
  col2 <- rev(brewer.pal(7, "RdYlBu"))
  pheatmap::pheatmap(
    data,
    color = colorRampPalette(col2)(100),
    breaks = if (!is.null(max_val)) {
      seq(-max_val, max_val, length.out = 101)
    } else {
      NULL
    },
    annotation_row = annotation_features,
    annotation_col = annotation_samples,
    ...
  )
}

# old alternativer version:
# set sample order (columns)
# NB! can not be used with denoise = TRUE!! (because it changes the Z values used to compute the denoised values used in the heatmap )
set_sample_order <- function(MOFAmodel, factor, order) {
  Z <- MOFAmodel@expectations$Z$single_group[, factor]

  # assign artificial values based on your desired order
  Z[order] <- seq_along(order)

  MOFAmodel@expectations$Z$single_group[, factor] <- Z
  MOFAmodel
}

# MOFAmodel2 <- set_sample_order(MOFAmodel, factor = 1, order = sample_order)
```

```{r transcriptomics_heatmap}
#| fig-width: 10
#| code-fold: true
meta_headers <- colnames(samples_metadata(MOFAmodel))
gr_df <- samples_metadata(MOFAmodel)[, c(
  "pos",
  "fct1_clus",
  "Clinical_score_(0-5)",
  "RVVC_pos",
  "CST",
  "BV",
  "Wet smear: clue cells (y/n)",
  "metab_louvain"
)] %>% #"sample",
  mutate_at(., 1:ncol(.), factor)
rownames(gr_df) <- samples_metadata(MOFAmodel)$sample

# determine column order when cluster_cols=F
sample_order <- meta %>%
  filter(sample %in% rownames(gr_df)) %>%
  # mutate(fct1_clus = factor(.$fct1_clus, levels = c("3", "4", "1", "2"))) %>%
  arrange(RVVC_pos) %>%
  #arrange(RVVC_pos, `Clinical_score_(0-5)`) %>%
  arrange(fct1_clus, desc(`Clinical_score_(0-5)`)) %>%
  pull(sample)

gr_df <- gr_df[sample_order, ]

ann_colors <- map(colnames(gr_df), function(val) {
  # Levels for this column
  levs <- levels(gr_df[[val]])

  # CASE 1: column exists in pal → use provided palette
  if (val %in% names(pal)) {
    colors <- pal[[val]]
    return(set_names(colors, levs))
  }
  # CASE 2: column not in pal → generate paired palette
  colors <- gg_color_hue(length(levs))
  set_names(colors, levs)
  # colors <- brewer.pal(length(levs), "Paired")
  # if (length(levs) == length(colors)) {
  #   set_names(colors, levs)
  # } else {
  #   set_names(colors[1:length(levs)], levs)
  # }
})
# Name the list by column names
ann_colors <- set_names(ann_colors, colnames(gr_df))

# quartz(width = 6, height = 4, dpi = 150)
plot_data_heatmap2(
  MOFAmodel,
  view = "Transcriptomics",
  factor = 1,
  features = 20,
  max_val = 3.5,
  sample_order = sample_order,
  cluster_cols = FALSE,
  annotation_samples = gr_df[, 2:5], #gr_df[,-1], "Clin_gr"
  denoise = FALSE,
  scale = "row",
  annotation_colors = ann_colors
)

# quartz(width = 7, height = 4, dpi = 150)
plot_data_heatmap2(
  MOFAmodel,
  view = "Metabolomics",
  factor = 1,
  features = 20,
  #max_val = 3.5,
  #sample_order = NULL,
  cluster_cols = FALSE,
  annotation_samples = gr_df[, c(2:5)], #gr_df[,-1], "Clin_gr"
  denoise = F,
  scale = "row",
  annotation_colors = ann_colors
)

plot_data_heatmap2(
  MOFAmodel,
  view = "Functional",
  factor = 1,
  features = 20,
  #max_val = 3.5,
  #sample_order = NULL,
  cluster_cols = FALSE,
  annotation_samples = gr_df[, c(2:5)], #gr_df[,-1], "Clin_gr"
  denoise = F,
  scale = "row",
  annotation_colors = ann_colors
)
```

plot_data_heatmap has an interesting argument to “beautify” the heatmap: `denoise = TRUE`. Instead of plotting the (noisy) input data, we can plot the data reconstructed by the model, where noise has been removed:

```{r divide-groups-data}
# quartz(width = 4, height = 3, dpi = 150)
p <- plot_factors(
  MOFAmodel,
  factors = c(1, 2),
  color_by = "Clinical_score_(0-5)",
  shape_by = "pos",
  dot_size = 2.5,
  show_missing = T
) +
  scale_fill_manual(values = pal[["Clinical_score_(0-5)"]], na.value = "white")

p <- p +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = (-0), linetype = "dashed")
p

plot_factors(
  MOFAmodel,
  factors = c(1, 3),
  color_by = "Clin_gr"
) +
  scale_fill_manual(values = pal$Clin_gr, na.value = "white") +
  stat_ellipse(
    #aes(color = RVVC_AS, fill = RVVC_AS),
    geom = "polygon",
    alpha = 0.25
  )
p

p <- plot_factors(
  MOFAmodel,
  factors = c(1, 3),
  color_by = "trx_louvain_7",
  shape_by = "Clin_gr",
  dot_size = 2.5,
  show_missing = T
) +
  scale_fill_manual(values = pal$trx_louvain_7, na.value = "white")

p <- p +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = (-0), linetype = "dashed")
p

p <- plot_factors(
  MOFAmodel,
  factors = c(2, 1),
  color_by = "CST",
  shape_by = "Clin_gr",
  dot_size = 2.5,
  show_missing = T
)

p <- p +
  scale_fill_manual(values = pal$CST, na.value = "white") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = (-0), linetype = "dashed")
p

p <- plot_factors(
  MOFAmodel,
  factors = c(4, 1),
  color_by = "CST",
  shape_by = "Clin_gr",
  dot_size = 2.5,
  show_missing = T
) +
  scale_fill_manual(values = pal$CST, na.value = "white")
p
# scale_fill_manual(values = pal$CST, na.value = "white") +
# stat_ellipse(
#   #aes(color = RVVC_pos),
#   geom = "polygon",
#   alpha = 0.25
# )
```


```{r scatter-plot}
#| fig-width: 13
#| fig-height: 10
plot_data_scatter(
  MOFAmodel,
  view = "Transcriptomics",
  factor = 1,
  features = 20,
  color_by = "Clinical_score_(0-5)"
) +
  scale_fill_manual(
    values = pal$`Clinical_score_(0-5)`,
    na.value = "transparent"
  )

plot_data_scatter(
  MOFAmodel,
  view = "Metabolomics",
  factor = 1,
  features = 20,
  color_by = "RVVC_AS"
) +
  scale_fill_manual(values = pal$Clin_gr, na.value = "transparent")
```

## 12. Dimensionality reduction

```{r umap}
set.seed(42)
MOFAmodel <- run_umap(MOFAmodel)

plot_dimred(
  MOFAmodel,
  method = "UMAP",
  color_by = "RVVC_pos",
  dot_size = 5
) +
  scale_fill_manual(values = pal$RVVC_pos, na.value = "transparent")

#MOFAmodel <- run_umap(MOFAmodel, factors = c("Factor4"))
```

## Step 5: Shankey plot

```{r shankey-plot}
#| fig-width: 5.25
#| fig-height: 5.25
library(niceRplots)

plot_meta_sankey <- function(
  vars = c("CST", "trx_louvain_7"),
  palette = pal1,
  drop_na = TRUE,
  return_pdf = TRUE,
  pdf_name = "sankey"
) {
  stopifnot(length(vars) == 2)

  ## Select and clean metadata
  m <- meta %>%
    filter(sample %in% m$sample[m$at_least_two & m$all_BL]) %>%
    dplyr::select(dplyr::all_of(vars))

  if (drop_na) {
    #m <- tidyr::drop_na(m)
    m <- tidyr::drop_na(m, vars[1])
    m <- mutate(
      m,
      !!vars[2] := tidyr::replace_na(as.character(.data[[vars[2]]]), "NA")
    )
  }

  ## Extract labels
  tmp <- m[, vars, drop = FALSE]
  tmp <- as.data.frame(tmp)

  ## Ensure factors
  tmp[[1]] <- factor(tmp[[1]])
  tmp[[2]] <- factor(tmp[[2]])

  ## Plot sankey
  #pdf( paste0("../results/integ_louvain_sankey.pdf"),width = 1.5*3.5,height = 1.5*3.5)
  #par(mar=c(4,4,4,4))
  #plot_sankey(df = tmp, pal = pal1)
  #dev.off()

  plot_expr <- {
    plot_sankey(df = tmp, pal = c(palette, "NA" = "gray"))
  }

  if (return_pdf) {
    ## Plot sankey
    pdf(
      paste0("../Results/sankey/", pdf_name, ".pdf"),
      width = 1.5 * 3.5,
      height = 1.5 * 3.5
    )
    par(mar = c(4, 8, 4, 4))
    plot_sankey(df = tmp, pal = palette)
    dev.off()
  }
  #plot_sankey(df = tmp, pal = palette)
}

# NB! the plot_meta_sankey function filters samples so that only the MOFA samples are included
# quartz(width = 6, height = 6, dpi = 150)
v <- c("metab_louvain", "CST")
plot_meta_sankey(v, pdf_name = "sankey_metab_CST", palette = pal[[v[1]]])

v <- c("microb_louvain", "CST")
plot_meta_sankey(v, pdf_name = "sankey_microb_CST", palette = pal[[v[1]]])

v <- c("CST", "Clinical_score_(0-5)")
plot_meta_sankey(v, pdf_name = "sankey_CST_Clin_score", palette = pal[[v[1]]])

v <- c("CST", "trx_louvain_7")
plot_meta_sankey(v, pdf_name = "sankey_CST_TRX", palette = pal[[v[1]]])

v <- c("metab_louvain", "BV")
plot_meta_sankey(v, pdf_name = "sankey_metab_BV", palette = pal[[v[1]]])


fct <- c(
  "trx_louvain_7",
  "metab_louvain",
  "microb_louvain",
  "fun_louvain"
) %>%
  set_names()

map(
  fct,
  ~ plot_meta_sankey(
    c("RVVC_pos", .x),
    pdf_name = paste0("RVVC_pos_", .x, "_sankey"),
    palette = pal[["RVVC_pos"]]
  )
)

#pdf( paste0("../results/integ_louvain_sankey.pdf"),width = 1.5*3.5,height = 1.5*3.5)
#par(mar=c(4,4,4,4))
#plot_sankey(df = tmp, pal = pal1)
#dev.off()
```


```{r alluvial-plot}
# install.packages("ggalluvial")
library(ggalluvial)

lov <- c(
  "Transcriptomics" = "trx_louvain_7",
  "Metabolomics" = "metab_louvain",
  "Microbiome" = "microb_louvain",
  "Functional" = "fun_louvain"
)

# --- 1. Prepare dataframe ---
# Ensure factors are treated as characters (avoid NA-level problems)
df_long <- result %>%
  select(sample, trx_louvain_7, metab_louvain, fun_louvain, microb_louvain) %>%
  mutate(across(-sample, as.character)) %>%
  drop_na() %>%
  pivot_longer(
    cols = -sample,
    names_to = "col",
    values_to = "cluster"
  ) %>%
  mutate(
    omic = factor(
      col,
      levels = lov,
      labels = names(lov)
    ),
    color = mapply(
      function(colname, clust) {
        pal[[colname]][as.integer(clust)]
      },
      col,
      cluster
    )
  )


# --- 2. Build the alluvial structure ---
l <- c(
  "Tr_4",
  "Tr_2",
  "Tr_1",
  "Tr_3",
  "Tr_6",
  "Tr_5",
  "Me_5",
  "Me_1",
  "Me_3",
  "Me_2",
  "Me_6",
  "Me_4",
  "Mi_4",
  "Mi_1",
  "Mi_2",
  "Mi_3",
  "Mi_5",
  "Fu_5",
  "Fu_2",
  "Fu_3",
  "Fu_1",
  "Fu_4",
  "Fu_6"
)

df_alluv <- df_long %>%
  mutate(alluvium = sample) %>%
  mutate(lvl = paste0(str_extract(omic, "^\\w\\w"), "_", cluster)) %>%
  mutate(lvl = factor(.$lvl, levels = l))

# --- 3. Plot ---
ggplot(
  df_alluv,
  aes(
    x = omic,
    stratum = lvl, #cluster,
    alluvium = alluvium,
    fill = color,
    label = cluster
  )
) +
  geom_flow(alpha = 0.5, color = "gray40") +
  geom_stratum(color = "black") +
  geom_text(stat = "stratum", color = "black", size = 3) +
  scale_fill_identity() +
  #scale_fill_brewer(type = "qual", palette = "Set2") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Cluster Transitions Across Omics Layers",
    x = "Omics Layer",
    y = "Sample Count",
    fill = "Cluster"
  )
```

### Extracting data for downstream analysis

```{r get-factors-weights-or-data}
#| eval: false
# "factors" is a list of matrices, one matrix per group with dimensions (nsamples, nfactors)
factors <- get_factors(MOFAmodel, factors = "all")
lapply(factors, dim)
# long format
factors <- get_factors(MOFAmodel, as.data.frame = TRUE)
head(factors, n = 3)

# "weights" is a list of matrices, one matrix per view with dimensions (nfeatures, nfactors)
weights <- get_weights(MOFAmodel, views = "all", factors = "all")
lapply(weights, dim)
# long format
weights <- get_weights(MOFAmodel, as.data.frame = TRUE)
head(weights, n = 3)

# "data" is a nested list of matrices, one matrix per view and group with dimensions (nfeatures, nsamples)
data <- get_data(MOFAmodel)
lapply(data, function(x) lapply(x, dim))[[1]]
# long format
data <- get_data(MOFAmodel, as.data.frame = TRUE)
head(data, n = 3)
```

```{r test-varince-between-groups}
factors <- get_factors(MOFAmodel, as.data.frame = TRUE)
factors <- factors %>%
  as_tibble() %>%
  select(-group) %>%
  left_join(meta, by = "sample") %>%
  filter(grepl(paste0("Factor", 1:9, "$", collapse = "|"), .$factor))

plot_factor_violins <- function(data, group_var, reff = NULL, comparisons) {
  p <- ggplot(
    data,
    aes(x = .data[[group_var]], y = value, fill = .data[[group_var]])
  ) +
    geom_violin(trim = F) +
    #geom_point() +
    geom_boxplot(
      width = 0.1,
      outlier.shape = NULL,
      outlier.size = 0.5,
      box.linewidth = 0.5
    ) +
    facet_wrap(~factor, scales = "free_y", nrow = 1) +
    #facet_grid(rows = vars(RVVC_pos), cols = vars(factor), scales = "free", space = "free") +
    labs(
      x = "Groups",
      y = "Factor value"
    ) +
    theme_classic() +
    scale_fill_manual(values = pal[[group_var]]) +
    theme(
      strip.text = element_text(size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )
  # Add stats if provided
  #if (!is.null(comparisons)) {
  p <- p +
    stat_compare_means(
      ref.group = reff,
      vjust = 2.7,
      label.y = Inf,
      size = 2,
      comparisons = comparisons,
      label = "p.signif",
      method = "wilcox.test"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
  #}
  return(p)
}

my_comparisons <- list(
  list(c("RVVC_pos", "Control_neg")),
  list(c("0", "5"), c("4", "5"))
  #list( c("0", "1-3"), c("0", "4-5") )
)

# quartz(width = 15, height = 6)
cols <- c("RVVC_pos", "Clinical_score_(0-5)")
reff <- c("Control_neg", "0")
update_geom_defaults("violin", aes(linewidth = 0))
plots <- map2(
  cols,
  reff, # my_comparisons,
  ~ plot_factor_violins(
    data = factors,
    group_var = .x,
    reff = .y,
    comparisons = NULL
  )
)
wrap_plots(plots, ncol = 1, axis_titles = "collect")

ggsave(
  paste0("../Results/Factor_plots/Factor_test_1-9", ".pdf"),
  width = 19,
  height = 6
)
```

# Summary

This file is a fully executable Quarto document with named chunks for reproducibility, caching, and debugging.

```{r extra}
#| eval: false
# code not currently in use, but could be usefull

## ---- overview-plot ------------------------------------------------
# shows the same as overview plot, but coloured by metadata and not by omic type
mofa_df %>%
  left_join(select(meta, sample, RVVC_AS), by = "sample") %>%
  select(sample, view, RVVC_AS) %>%
  unique() %>%
  mutate(RVVC_AS = factor(.$RVVC_AS)) %>%
  mutate(sample = fct_reorder(sample, as.integer(RVVC_AS), .na_rm = FALSE)) %>%

  ggplot(., aes(x = sample, y = view, fill = RVVC_AS)) +
  geom_tile() + #scale_fill_manual(values = c(missing = "grey", colors)) +
  theme_nothing() +
  theme(
    axis.text.y = element_text(size = 12),
    legend.position = "top",
    plot.margin = unit(c(1, 1, 0, 1), "cm")
  ) +
  scale_fill_manual(values = pal$Clin_bin, na.value = "transparent")


## ---- mofa-variance-barplot ------------------------------------------------
plot_variance_explained(
  MOFAmodel,
  plot_total = FALSE
)

df_long <- get_factors(
  MOFAmodel,
  factors = "all",
  groups = "all",
  as.data.frame = TRUE
) %>%
  left_join(., samples_metadata(MOFAmodel)[, -1], by = "sample") %>%
  as_tibble()

df_wide <- pivot_wider(df_long, names_from = "factor", values_from = "value")
plot_fact.fun <- function(X, Y, gr_val, pal) {
  p <- ggplot(
    df_wide,
    aes(x = .data[[X]], y = .data[[Y]], colour = .data[[gr_val]])
  ) +
    geom_point(size = 2, alpha = 0.9) +
    labs(x = X, y = Y) +
    theme_classic() +
    scale_fill_manual(
      values = pal[1:length(unique(na.omit(df_wide[[gr_val]])))],
      aesthetics = c("colour", "fill")
    ) +
    theme(
      axis.text = element_text(size = rel(0.8), color = "#373844"),
      axis.title = element_text(size = rel(1.1), color = "#373844"),
      axis.line = element_line(color = "#373844", linewidth = 0.5),
      axis.ticks = element_line(color = "#373844", linewidth = 0.5)
    )
  return(p)
}
map(vars, ~ plot_fact.fun("Factor1", "Factor3", .x, pal3))
plot_fact.fun("Factor1", "Factor3", "group_2", pal3)

## other version:
# df_long <- get_factors(
#   MOFAmodel,
#   factors = "all",
#   groups = "all",
#   as.data.frame = TRUE
# ) %>%
#   left_join(., samples_metadata(MOFAobject)[, -2], by = "sample") %>%
#   as_tibble()

# df_wide <- pivot_wider(
#   df_long,
#   names_from = "factor",
#   values_from = "value"
# ) %>%
#   mutate(lab = ifelse(.$RVVC_AS == "RVVC_AS", .$sample, NA))

# plot_fact.fun <- function(X, Y, gr_val, pal) {
#   p <- ggplot(
#     df_wide,
#     aes(x = .data[[X]], y = .data[[Y]], colour = .data[[gr_val]])
#   ) +
#     geom_point(size = 2, alpha = 0.9) +
#     geom_text(
#       aes(label = .data[["lab"]]),
#       size = 3,
#       hjust = 0,
#       nudge_x = 0.07,
#       color = "gray51"
#     ) +
#     labs(x = X, y = Y) +
#     theme_classic() +
#     scale_fill_manual(
#       na.value = "white",
#       values = pal[1:length(unique(na.omit(df_wide[[gr_val]])))],
#       aesthetics = c("colour", "fill")
#     ) +
#     theme(
#       axis.text = element_text(size = rel(0.8), color = "#373844"),
#       axis.title = element_text(size = rel(1.1), color = "#373844"),
#       axis.line = element_line(color = "#373844", linewidth = 0.5),
#       axis.ticks = element_line(color = "#373844", linewidth = 0.5)
#     )
#   return(p)
# }
# vars <- c(
#   "RVVC_AS",
#   "Clin_gr",
#   "group_2",
#   "hyphae",
#   "group_cross",
#   "Clinical_score_(0-5)"
# ) %>%
#   set_names()

# # colour pallets
# blue <- brewer.pal(6, name = "Blues") %>% c(., "gray")
# Red <- brewer.pal(6, name = "Reds") #%>% c(., "gray")
# pal <- c("#f26386", "#fd814e", "#a4d984", "#fbbc52") # "#f588af", "#fd814e"
# pal2 <- c("#2266ac", "#91c5de")

# #Red <- brewer.pal(6, name = "Reds") #%>% c(., "gray")
# pal1 <- c("#ffa998", '#FB5273', "#902267") %>% c(., "gray") #
# pal3 <- c("#9e8dec", "#cc93cf", "#a5c97b", "#de9548") #"#abcb4b" "#63d3b4" "#902267",

# plot_fact.fun("Factor1", "Factor3", "group_2", pal3)

# p <- map(vars, ~ plot_fact.fun("Factor2", "Factor1", .x, pal3))
# p[["RVVC_AS"]] +
#   stat_ellipse(
#     aes(color = RVVC_AS, fill = RVVC_AS),
#     geom = "polygon",
#     alpha = 0.25
#   )

## ---- weights-by-view ------------------------------------------------
# test the
weights_TR <- get_weights(MOFAmodel, views = "Transcriptomics")$Transcriptomics
head(sort(abs(weights_TR[, 1]), decreasing = TRUE))

TRX <- t(X$species$TR)
mt_genes <- grepl("^MT", rownames(TRX))
mt_fraction <- colSums(TRX[mt_genes, ]) / colSums(TRX)

cor(
  get_factors(MOFAmodel)$group1[, 1],
  mt_fraction,
  use = "complete.obs"
)
```