---
title: "Multi-omics integration with MOFA2"
date: today
date-format: "D MMM YYYY"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    highlight-style: github
    code-tools:
      source: true
      toggle: true
params:
  sparse: false
---

# Overview

This report presents a **structured MOFA2 analysis** integrating transcriptomics, metagenomics, and metabolomics data. The goal is to identify **latent factors capturing shared and view-specific sources of variation**, and to generate a **set of publication-ready figures** that can be combined into a multi‑panel figure.

**Main outcomes**:
- Latent factor structure across multi‑omics layers
- Variance explained per factor and per view
- Associations between factors and clinical / technical covariates
- Feature loadings driving key factors

---

## Analysis workflow (step‑by‑step)

1. Load required libraries and define plotting palettes
2. Load omics data and sample metadata
3. Format MOFA input
4. Select samples for inclusion
5. Unsupervised UMAPs
6. Configure and run the MOFA model
7. Load trained MOFA model
8. Perform model diagnostics and quality checks
9. Covariates- Factor correlation
10. Annotate latent factors using group information
11. Feature-level figures
12. Unified UMAP (all views)

---

## 1. Libraries and global settings

```{r libraries}
#| code-fold: true
#| output: false
#| warning: false
#| message: false
library(MOFA2)
library(tidyverse)
library(jsonlite)
library(patchwork)
library(openxlsx)
library(ggrepel)

# remotes::install_github("ncats/RAMP-DB")
library(RaMP)

# setwd("/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/src")

select <- dplyr::select
map <- purrr::map
rename <- dplyr::rename

# default ggplot2 pallet:
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
gg_color_hue(10)
library(paletteer)

gene_col <- c(
  "#4F6980",
  "#849DB1",
  "#A2CEAA",
  "#638B66",
  "#BFBB60",
  "#9F8F12",
  "#F47942",
  "#FBB04E",
  "#B66353",
  "#D7CE9F"
)
gene_col <- rev(c(
  "#6388B4FF",
  "#8CC2CAFF",
  "#EF6F6AFF",
  "#FFAE34FF",
  "#55AD89FF",
  "#C3BC3FFF",
  "#BB7693FF",
  "#94A323",
  "#A9B5AEFF",
  "#E76BF3"
))
gene_col <- c(
  "#B71E42FF",
  "#DE478E",
  "#BC72F0FF",
  "#795FAFFF",
  "#D53DD0",
  "#EA6312",
  "#FFAE34FF",
  "#EF6F6AFF",
  "#8CC2CAFF",
  "#6388B4FF"
)
pal2 <- list(
  gene_clus = set_names(gene_col, c(paste0("pos_", 1:6), paste0("neg_", 1:4))),
  pos = set_names(gene_col[1:6], paste0(1:6)),
  neg = set_names(gene_col[7:10], paste0(1:4))
)
```

Define consistent colour palettes to ensure visual coherence across all figures.
```{r colour-pallet}
#| code-fold: true
select <- dplyr::select
map <- purrr::map
rename <- dplyr::rename

pal <- list(
  CST = c(
    "I" = "#148F77",
    "II" = "#6d9f06",
    "III" = "#A3E4D7",
    "IV-B" = "#E794C1",
    "IV-C" = "#b5cb8c",
    "V" = "#43BA8F"
  ),
  subCST = c(
    "#a5c97b",
    "#cc93cf",
    "#de9548",
    "#63d3b4",
    "#9e8dec",
    "#902267",
    '#FB5273',
    "#91c5de"
  ),
  RVVC = c("#EAEAEA", "#fd814e"),
  BV = c("#EAEAEA", "#fd814e"),
  NAC = c("#EAEAEA", "#fd814e"),
  ST = c("#EAEAEA", "#fd814e"),
  missing_page = c("#EAEAEA", "#fd814e"),
  Clin = c("#a5c97b", "#cc93cf", "#de9548", "#63d3b4", "#9e8dec", "#902267"),
  Clin_gr = c("#ffa998", '#FB5273', "#902267"),
  RVVC_AS = c("#ffa998", '#FB5273', "#902267"),
  RVVC_pos = c(
    "#A6CEE3",
    "#1F78B4",
    #"#EAEAEA",
    #"gray",
    "#e490c4ff",
    "#c12d88ff"
  ), # "#FB9A99", "#E31A1C", "#B2DF8A", "#609c5bff", "#FDBF6F", "#FF7F00",, "#CAB2D6", "#6a3d9a" "#e490c4ff", "#c12d88ff"
  trx_louvain_7 = c(
    "#de9548",
    "#cc93cf",
    "#a5c97b",
    "#9e8dec",
    "#902267",
    "#63d3b4",
    '#FB5273'
  ),
  metab_louvain = c(
    "#cc93cf",
    "#a5c97b",
    "#de9548",
    "#902267",
    "#9e8dec",
    "#63d3b4"
    #'#FB5273'
  ),
  microb_louvain = c(
    "#cc93cf",
    "#de9548",
    "#a5c97b",
    "#9e8dec",
    "#902267",
    "#63d3b4",
    '#FB5273'
  ),
  fun_louvain = c(
    "#a5c97b",
    "#cc93cf",
    "#de9548",
    "#63d3b4",
    "#9e8dec",
    "#902267",
    '#FB5273'
    #"#984EA3"
    #"#B3CDE3"
  ),
  integ_louvain = c(
    "#a5c97b",
    "#cc93cf",
    "#de9548",
    "#63d3b4",
    "#9e8dec",
    "#902267",
    '#FB5273',
    "#984EA3"
    #"#B3CDE3"
  ),
  "Wet_smear:_clue_cells_(y/n)" = c("#2266ac", "#91c5de"),
  hyphae = c("#f26386", "#fd814e", "#a4d984", "#fbbc52"),
  pos = c("#91c5de", "#2266ac"),
  fct1_clus = c("#33A02C", "#B2DF8A", "#FF7F00", "#FDBF6F"), #"#FDBF6F" "#FF7F00", "#CAB2D6", "#6A3D9A"
  "Clinical_score_(0-5)" = brewer.pal(6, name = "Reds")
)
# default ggplot2 pallet:
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
gg_color_hue(4)
```

## 2. Data loading
Multi‑omics matrices and sample metadata are loaded from pre‑processed RDS/CSV files. These data have already undergone quality control and normalisation upstream.
```{r load_data}
#| warning: false
#| message: false
comb_meta_path <- "../Results/Combined_gr_meta_data.csv"

# meta data
meta <- read_csv(comb_meta_path) %>%
  #rename(sample = "svamp_ID") %>%
  mutate(across(2:27, ~ factor(.x)))

KEGG_path <- "../../Spatial_DMPA/resources/c2.cp.kegg_legacy.v2023.2.Hs.json"
GO_path <- "../../Spatial_DMPA/resources/c5.all.v2023.2.Hs.json"
Wiki_path <- "../resources/wikipathways-20260310-gmt-Homo_sapiens.gmt"
# TFT_path <- "../../Spatial_DMPA/resources/c3.tft.gtrd.v2023.2.Hs.symbols.gmt"
```

## 7. Load trained MOFA model
We load the model with the highest elbo score
```{r load_model}
#| eval: true
#| include: true
#| warning: false
#| message: false
#| tbl-colwidths: [60,40]
elbo_table <- read_csv("../Results/MOFA/elbo_table_noS84.csv")
knitr::kable(elbo_table, )
best_seed <- elbo_table$seed[which.max(abs(elbo_table$elbo))]

MOFAmodel <- load_model(
  #"../Results/MOFA/MOFA_matched_seed_50.hdf5"
  # "../Results/MOFA/MOFA_allBL_seed_"
  paste0("../Results/MOFA/MOFA_allBL_noS84_seed_", best_seed, ".hdf5")
)
```

```{r get-factor-clusters}
# Extract factors, and join metadata
clusters <- list()
clusters$fct1_clus <- cluster_samples(MOFAmodel, k = 4, factors = 1)$cluster
clusters$fct2_clus <- cluster_samples(MOFAmodel, k = 4, factors = 2)$cluster
clusters$fct4_clus <- cluster_samples(MOFAmodel, k = 4, factors = 4)$cluster

clust <- bind_cols(sample = names(clusters[[1]]), clusters)

meta <- meta %>%
  dplyr::left_join(., clust, by = "sample") %>%
  dplyr::mutate(across(all_of(names(clusters)), ~ factor(.x)))

# add meta data to model
samples_metadata(MOFAmodel) <- samples_metadata(MOFAmodel)[, c(
  "sample",
  "group"
)] %>%
  left_join(meta, by = c("sample"))
```

```{r read-db}
read_json.fun <- function(db) {
  db <- db %>%
    map(., ~ as_tibble(.x[4])) %>%
    bind_rows(., .id = "pathway") %>%
    mutate(genes = map(db, ~ pluck(.x, "geneSymbols")))
  return(db)
}

read_gmt.fun <- function(file) {
  read_lines("../resources/wikipathways-20260310-gmt-Homo_sapiens.gmt") %>%
    strsplit("\t") %>%
    map(
      ~ tibble(
        pathway = .x[1],
        description = .x[2],
        feature = .x[-c(1, 2)]
      )
    ) %>%
    bind_rows() %>%
    group_by(pathway) %>%
    summarise(
      genes = list(feature),
      .groups = "drop"
    ) %>%
    separate(
      .,
      pathway,
      into = c("pathway", "source", "exactSource", "species"),
      sep = "%",
      remove = T
    )
}

GO_database <- read_json.fun(fromJSON(GO_path))
# KEGG_database <- read_json.fun(fromJSON(KEGG_path))
# Wiki_database <- read_gmt.fun(Wiki_path)
```

In WikiPathways (and therefore RaMP):

A gene can participate multiple times in a pathway
Each occurrence may correspond to:

a different node
a different interaction
a different chromosomal region
a different sub-pathway context

RaMP flattens these representations into a long table, which preserves biological structure rather than forcing uniqueness.
So you’re not seeing duplicates because of a mistake — you’re seeing duplicates because the same gene is referenced multiple times within the same pathway model.


```{r old}
#| eval: false
make_binary_matrix <- function(database) {
  databases$wiki %>%
    select(pathway, exactSource, genes) %>%
    unnest(genes) %>%
    distinct(pathway, genes) %>%
    mutate(value = 1) %>%
    pivot_wider(
      names_from = genes,
      values_from = value,
      values_fill = 0
    ) %>%
    select(pathway, genes)
  column_to_rownames("pathway") %>%
    as.matrix()
}

run_single_enrichment <- function(
  MOFAmodel,
  binary_matrix,
  view,
  factors,
  sign,
  statistical.test = "parametric"
) {
  run_enrichment(
    MOFAmodel,
    view = view,
    factors = factors,
    feature.sets = binary_matrix,
    sign = sign,
    statistical.test = statistical.test
  )
}

run_enrichment_grid <- function(
  MOFAmodel,
  databases,
  views,
  signs,
  factors = 1:3,
  statistical.test = "parametric"
) {
  # Precompute binary matrices
  binary_matrices <- purrr::map(databases, make_binary_matrix)

  # Parameter grid
  param_grid <- expand.grid(
    db = names(binary_matrices),
    view = views,
    sign = signs,
    stringsAsFactors = FALSE
  )

  # Run all combinations
  results <- purrr::pmap(
    param_grid,
    function(db, view, sign) {
      run_single_enrichment(
        MOFAmodel = MOFAmodel,
        binary_matrix = binary_matrices[[db]],
        view = view,
        factors = factors,
        sign = sign,
        statistical.test = statistical.test
      )
    }
  )

  # Name results clearly
  names(results) <- paste(
    param_grid$db,
    param_grid$view,
    param_grid$sign,
    sep = "__"
  )

  return(results)
}

databases <- list(
  KEGG = KEGG_database,
  GO = GO_database,
  wiki = Wiki_database
)

signs <- c("negative", "positive", "all")
views <- c("Transcriptomics")

enrichment_results <- run_enrichment_grid(
  MOFAmodel = MOFAmodel,
  databases = databases,
  views = views,
  signs = signs,
  factors = 1:3
)
```


```{r}
# initializes the RaMP database object, downloading and caching the latest SQLite database
# if no version already exists in local cache.
rampDB <- RaMP()

# note that you can use the following method to check database versions hosted in your
# computer's local cache and databases that are available to download in our remote repository.
RaMP::listAvailableRaMPDbVersions()

# RaMP::getPrefixesFromAnalytes(db = rampDB, analyteType = 'metabolite')
RaMP::getPrefixesFromAnalytes(db = rampDB, analyteType = 'gene')
```


```{r}
# ============================================================
# STEP 1: Get all pathways available in RaMP
# (this function is explicitly shown / implied in the vignette)
# ============================================================

pathways <- RaMP::getPathwayNameList()
# expected columns typically include:
# pathway_name, pathway_source, pathway_id (depending on version)

# ============================================================
# STEP 2: For each pathway, retrieve its analytes
# IMPORTANT:
# - RaMP has NO global "pathway x analyte" table
# - The vignette shows that getAnalyteFromPathway() must be
#   called per pathway
# ============================================================

# RaMP_long <- getAnalyteFromPathway(pathways, analyteType = "metabolite")
RaMP_long <- getAnalyteFromPathway(pathways, analyteType = "gene", )
RaMP_long <- RaMP_long %>% as_tibble()

table(RaMP_long$pathwayType)
table(RaMP_long$geneOrCompound)

RaMP_long <- RaMP_long %>%
  select(
    pathway = "pathwayName",
    exactSource = "pathwayId",
    db = "pathwayType",
    genes = "analyteName"
  )

# ============================================================
# STEP 3: Add GO db to RaMP_long
# ============================================================

RaMP_long <- GO_database %>%
  unnest(genes) %>%
  mutate(db = "GO") %>%
  filter(grepl("^GOBP_", pathway)) %>%
  bind_rows(RaMP_long)

# databases:
table(RaMP_long$db)
# number of pathways
table(unique(RaMP_long[, c("pathway", "db")])$db)

# ============================================================
# STEP 4: Split into individual DB's
# ============================================================

RaMP_db <- RaMP_long %>%
  nest(.by = "db") %>%
  mutate(data = set_names(data, .data[["db"]]))

# ============================================================
# STEP 4: Build the binary membership matrix
# rows   = pathways
# cols   = metabolites (HMDB IDs)
# values = 0 / 1
# ============================================================

make_binary_matrix <- function(database) {
  database %>%
    # RaMP_db$data$kegg %>%
    distinct(exactSource, pathway, genes) %>%
    mutate(value = 1) %>%
    pivot_wider(
      id_cols = c(exactSource, pathway),
      names_from = genes,
      values_from = value,
      values_fill = 0
    ) %>%
    select(-exactSource) %>%
    column_to_rownames("pathway") %>%
    as.matrix()
}

# Precompute binary matrices
binary_matrices <- RaMP_db %>%
  filter(db != "pfocr") %>%
  mutate(matrix = imap(data, ~ make_binary_matrix(.x)))

# ============================================================
# STEP 6: Sanity check overlap with MOFA features
# (this step is CRITICAL)
# ============================================================

imap(
  binary_matrices$matrix,
  ~ length(
    intersect(
      colnames(.x),
      features_names(MOFAmodel)[["Transcriptomics"]]
    )
  )
)

# ============================================================
# STEP 7: Run MOFA enrichment (UNCHANGED from your pipeline)
# ============================================================

enrich.pos <-
  imap(
    binary_matrices$matrix,
    ~ run_enrichment(
      MOFAmodel,
      view = "Transcriptomics",
      factors = 1,
      feature.sets = .x,
      sign = "positive",
      alpha = 0.1, # FDR treshold
      statistical.test = "parametric"
    )
  )

enrich.neg <-
  imap(
    binary_matrices$matrix,
    ~ run_enrichment(
      MOFAmodel,
      view = "Transcriptomics",
      factors = c(1, 2),
      feature.sets = .x,
      sign = "negative",
      alpha = 0.1, # FDR treshold
      statistical.test = "parametric"
    )
  )

enrich.both <-
  imap(
    binary_matrices$matrix,
    ~ run_enrichment(
      MOFAmodel,
      view = "Transcriptomics",
      factors = 1,
      feature.sets = .x,
      sign = "all",
      alpha = 0.1, # FDR treshold
      statistical.test = "parametric"
    )
  )

keep_slots <- c("pval", "pval.adj", "set.statistics")


# enrichment_to_tibble <- function(
#   enrichment_obj,
#   keep_slots = c("pval", "pval.adj", "set.statistics")
# ) {
#   enrichment_obj %>%
#     imap(function(x, name_outer) {
#       components <- x[intersect(names(x), keep_slots)] %>%
#         imap(function(mat, name_inner) {
#           as_tibble(mat, rownames = "pathway") %>%
#             rename_with(~ paste0(name_inner, "_", .), -pathway)
#         })

#       reduce(components, full_join, by = "pathway") %>%
#         #arrange(across(starts_with("pval_"), ~.x)) %>%
#         mutate(source = name_outer)
#     })
# }

enrichment_to_tibble <- function(
  enrichment_obj,
  keep_slots = c("pval", "pval.adj", "set.statistics")
) {
  enrichment_obj %>%
    imap(
      .,
      ~ .x %>% # enrich.neg$GO %>%
        enframe(.) %>%
        filter(name %in% keep_slots) %>%
        mutate(data = map(value, ~ as_tibble(.x, rownames = "pathway"))) %>%
        select(name, data) %>%
        mutate(
          data = map2(data, name, function(df, nm) {
            rename_with(
              df,
              function(col) paste0(nm, "_", col),
              contains("Factor")
            )
          })
        ) %>%
        pull(data) %>%
        reduce(full_join, by = "pathway") %>%

        arrange(across(starts_with("pval_"), ~.x)) %>%
        mutate(source = .y)
    )
}

enrichment_list <- list(pos = enrich.pos, neg = enrich.neg, both = enrich.both)

result <- map(enrichment_list, enrichment_to_tibble)

# Suppl. Tables per comparison
imap(
  result,
  ~ write.xlsx(
    .x,
    keepNA = TRUE,
    na.string = "",
    overwrite = TRUE,
    file = paste0("../Results/Tables/", "GSEA_", .y, ".xlsx")
  )
)
```

#### Enrichment step (run_enrichment)
1. Takes all gene weights
2. Subsets genes by each pathway
3. Performs a statistical test (e.g. parametric test) on:

  weights inside the pathway vs weights outside the pathway

These are the steps for doing Gene Set Enrichment Analysis (GSEA) with MOFA:

    (1) **Define your gene set matrix:** this can be specified as a binary matrix where rows are gene sets and columns are genes. A value of 1 indicates that gene j belongs to pathway i. A value of 0 indicates elsewise.
    (2) **Select a gene set statistic:** the statistic used to quantify the scores at the pathway level. Must be one of the following: mean.diff (difference in the average weight between foreground and background genes) or rank.sum (difference in the sum of ranks between foreground and background genes).
    (3) **Select a statistical test:** the statistical test used to compute the significance of the gene set statistics under a competitive null hypothesis. Must be one of the following: parametric (a simple and very liberal parametric t-test), cor.adj.parametric (parametric t-test adjusted by the correlation between features), permutation (unparametric, the null distribution is created by permuting the weights. This option is computationally expensive, but it preserves the correlation structure between features in the data.).

An important consideration when running GSEA is that MOFA contains positive and negative weights. There will be cases where the genes with negative weights all belong to a specific pathway but genes with positive weights belong to other pathways. If this is true, doing GSEA with all of them together could dilute the signal. Hence, we recommend the user to do GSEA separately for (+) and (-) weights, and possibly also jointly with all weights.

The enrichment analysis returns a list of 5 elements:

    **feature.sets:** the feature set matrix filtered by the genes that overlap with the MOFA model.
    **pval:** the nominal p-values.
    **pval.adj:** the FDR-adjusted p-values.
    **feature.statistics:** the feature statistics (i.e. the weights).
    **set.statistics:** matrices with the gene set statistics.
    **sigPathways:** list with significant pathways per factor at a specified FDR threshold



```{r}
plot_enrichment_heatmap(enrichment_RaMP$GO)

plot_enrichment(enrich.neg$GO, factor = 1, max.pathways = 15)
```


```{r}
plot_enrichment_detailed(
  enrichment.results = enrichment_list$neg$GO,
  factor = 1,
  max.pathways = 3
)
plot_enrichment_detailed(
  enrichment.results = enrichment_list$pos$GO,
  factor = 1,
  max.pathways = 10
)

png(height = 25, filename = "./test_GSEA.png")
p <- plot_enrichment_detailed(
  enrichment.results = enrichment_list$both$GO,
  factor = 1,
  max.pathways = 34,
)
dev.off()

ggsave("../Results/test_GSEA.png", height = 25)

db <- enrich.both$GO$feature.sets
```


```{r GO-enrichment-heatmap}
# 1. Extracts GO enrichment results from the input object.
# 2. Selects pathways based on a score threshold (positive, negative, or both).
# 3. Builds a gene–pathway matrix and removes empty genes (columns).
# 4. Filters genes based on a minimum gene score threshold (weights).
# 5. Automatically adjusts plot height based on the number of pathways.
# 6. Creates row annotations (pathway scores and adjusted p-values).
# 7. Clusters genes (columns) and assigns them to groups.
# 8. Creates column annotations (gene weights and cluster groups).
# 9. Generates a heatmap with clustering and annotations.
# 10. Draws and saves the heatmap as a PNG file.
# 11. Returns the heatmap object, matrix, clustering results, and groups.

library(ComplexHeatmap)
library(circlize)
library(paletteer)

# Color function
col_fun <- colorRamp2(
  c(0, 1),
  c("#ecc4d0", "#e05e85")
)

# enrich_obj = enrichment_list$neg
# score_threshold = 3
# file = "../Results/GSEA_heatmap_neg.png"
# width = 10
# height = 5
# k_clusters = 4

make_gsea_heatmap <- function(
  enrich_obj,
  gene_col = "pos",
  cell_width = 0.1,
  score_threshold = 3,
  gene_treshold = 0,
  file,
  width = 10,
  height = 6,
  k_clusters = 4
) {
  temp <- enrich_obj$GO

  # Select pathways
  if (length(score_threshold) == 1) {
    sets <- rownames(temp$set.statistics)[
      abs(temp$set.statistics) > score_threshold
    ]
  } else {
    set <- tibble(
      name = rownames(temp$set.statistics),
      value = as.vector(temp$set.statistics)
    )
    sets_pos <- set %>% filter(value > 0 & value > score_threshold[1])
    sets_neg <- set %>% filter(value < 0 & abs(value) > score_threshold[2])
    sets <- c(sets_pos$name, sets_neg$name)
  }

  # Matrix prep
  mat <- temp$feature.sets[sets, , drop = FALSE]
  mat <- mat[, colSums(mat) != 0] # remove cols sum to zero

  # filter genes
  keep_genes <- temp$feature.statistics[colnames(mat), ] > gene_treshold
  mat <- mat[, keep_genes]

  if (nrow(mat) == 0 || ncol(mat) == 0) {
    warning("Matrix is empty after filtering")
    return(NULL)
  }

  # Auto height calculation
  base_height = 2 # border height
  per_pathway = 0.1 # height per row
  max_height = 35 # cap height
  height <- min(max_height, base_height + nrow(mat) * per_pathway)

  # Annotations
  row_ha <- rowAnnotation(
    set = temp$set.statistics[sets, ],
    p_val = temp$pval.adj[sets, ],
    annotation_legend_param = list(set = list(direction = "vertical"))
  )

  # Extract column dendrogram
  hc <- hclust(dist(t(mat)))
  dend <- as.dendrogram(hc)
  # dend <- column_dend(ht)
  # hc <- as.hclust(dend)
  groups <- cutree(hc, k = k_clusters)

  weight_col <- colorRamp2(
    seq(from = 0, to = 1, length.out = 7),
    c("#EAEBEB", paletteer_c("ggthemes::Sunset-Sunrise Diverging", 7)[-1])
  )
  column_ha <- HeatmapAnnotation(
    weight = temp$feature.statistics[colnames(mat), ],
    gr = as.character(groups),
    col = list(gr = pal2[[gene_col]], weight = weight_col),
    annotation_legend_param = list(weight = list(direction = "vertical"))
  )

  # Heatmap
  H <- Heatmap(
    mat,
    width = unit(ncol(mat) * cell_width, "mm"),
    name = "Present",
    col = col_fun,
    column_split = groups,
    cluster_rows = TRUE,
    top_annotation = column_ha,
    right_annotation = row_ha,
    row_names_side = "right",
    show_column_names = FALSE,
    column_names_rot = 90,
    row_names_gp = gpar(fontsize = 7),
    heatmap_legend_param = list(
      at = c(0, 1),
      labels = c("No", "Yes"),
      direction = "vertical"
    )
  )

  # Draw once (needed for dendrogram extraction)
  ht <- draw(
    H,
    heatmap_legend_side = "left",
    annotation_legend_side = "left",
    merge_legend = TRUE,
    padding = unit(c(2, 5, 2, 30), "mm")
  )

  # Save plot
  png(file = file, units = "in", res = 600, width = width, height = height)
  draw(ht)
  dev.off()

  # Return everything useful
  list(
    heatmap = ht,
    matrix = mat,
    dendrogram = dend,
    hclust = hc,
    column_groups = groups
  )
}
```

#### gene_treshold:
the default of 0 makes sure to remove genes that have the oposite weight to the specified
these are simply set to weight = 0 in the MOFA enrichment function, but retained in the matrix

#### score_threshold:
the default mean.diff represents the difference in the average weight between foreground and background genes

```{r}
ht_list_neg <- make_gsea_heatmap(
  enrich.neg,
  gene_col = "neg",
  score_threshold = 3,
  gene_treshold = 0,
  cell_width = 0.1,
  file = "../Results/Enrichment/GSEA_heatmap_neg_3.png",
  width = 10,
  k_clusters = 4
)

ht_list_pos <- make_gsea_heatmap(
  enrich.pos,
  score_threshold = 9,
  gene_treshold = 0,
  file = "../Results/Enrichment/GSEA_heatmap_pos.png",
  width = 12,
  k_clusters = 6
)

ht_list_pos <- make_gsea_heatmap(
  enrich.both,
  score_threshold = c(9, 3),
  gene_treshold = 0,
  file = "../Results/Enrichment/GSEA_heatmap_both.png",
  width = 12,
  k_clusters = 6
)
```


```{r weight_figure}
plot_top_weights_tidy <- function(
  weights_df,
  nfeatures = 10,
  fact = 1,
  abs = TRUE,
  scale = TRUE,
  sign = "all"
) {
  # Remove zeros
  W <- weights_df %>% filter(value != 0)

  W <- W
  weights %>%
    filter(grepl(paste0("^Factor", fact, "$", collapse = "|"), .$factor))

  # Scale
  if (scale) {
    W <- W %>% mutate(value = value / max(abs(value)))
  }

  # Sign
  W <- W %>%
    mutate(sign_label = if_else(value > 0, "+", "-"))

  if (sign == "positive") {
    W <- W %>% filter(value > 0)
  } else if (sign == "negative") {
    W <- W %>% filter(value < 0)
  }

  if (abs) {
    W <- W %>% mutate(value = abs(value))
  }

  # Select top features per factor
  W <- W %>%
    group_by(factor) %>%
    slice_max(order_by = value, n = nfeatures) %>%
    #arrange(value) %>%
    #fct_reorder(feature, value) %>%
    #mutate(feature_id = factor(feature)) %>%
    ungroup()

  # Factor ordering
  W <- W %>%
    arrange(factor, value) %>%
    mutate(feature_id = factor(feature, levels = unique(feature)))
  #mutate(feature_id = factor(feature, levels = rev(unique(feature))))

  gr <- unique(W$clus)
  # Plot
  p <- ggplot(W, aes(x = feature_id, y = value, color = clus)) +
    geom_point(size = 2) +
    geom_segment(aes(xend = feature_id, yend = 0), linewidth = 0.75) +
    scale_colour_manual(values = pal2[["gene_clus"]], na.value = "gray") +
    coord_flip() +
    labs(y = "Weight", x = NULL) +
    theme_bw() +
    theme(
      axis.text.y = element_text(size = rel(1.1)),
      panel.grid.major.y = element_blank(),
      legend.position = "top"
    ) +
    facet_wrap(~factor, nrow = 1, scales = "free")

  if (abs) {
    p <- p +
      ylim(0, max(W$value) + 0.1) +
      geom_text(aes(label = sign_label), y = max(W$value) + 0.1, size = 4)
  }

  return(p)
}
```

```{r plot-weights}
# intersect(names(ht_list_neg$column_groups), names(ht_list_pos$column_groups))
groups <- list(
  neg = ht_list_neg$column_groups,
  pos = ht_list_pos$column_groups
)

groups <- groups %>%
  imap(~ setNames(paste0(.y, "_", .x), names(.x))) %>%
  flatten_chr()
weights <- get_weights(MOFAmodel, as.data.frame = TRUE) %>%
  as_tibble() %>%
  filter(view == "Transcriptomics") %>%
  #filter(factor == "Factor1") %>% pull(feature) %>% unique()
  arrange(factor, desc(abs(value))) %>%
  mutate(clus = ifelse(factor == "Factor1", groups[.$feature], NA))

plot_top_weights_tidy(weights, nfeatures = 35, fact = c(1))


# MOFA functions:
plot_enrichment(enrich.neg$GO, factor = 1, max.pathways = 35)

plot_top_weights(
  MOFAmodel,
  fact = c(1, 2),
  view = "Transcriptomics",
  nfeatures = 10
)

plot_enrichment_detailed(
  enrichment.results = enrich.neg$GO,
  factor = 1,
  max.genes = 7,
  max.pathways = 3
)
```



```{r plot-pathway-indvidualy}
p.val_round <- function(p.val) {
  case_when(
    p.val >= 0.05 ~ sprintf("%.2f", p.val),
    p.val >= 0.001 ~ sprintf("%.3f", p.val),
    TRUE ~ sprintf("%.1e", p.val)
  )
}

plot_weights_pathways <- function(
  enrichment_object,
  weights_df,
  enrich_results = result$neg$GO,
  phtw,
  n_label = 5,
  x_axis = "enrich_weight" # or use "weight" for MOFA or "weight_scaled"
) {
  wgt <- filter(weights_df, factor == "Factor1") %>%
    select(genes = "feature", weight = "value", clus, factor)

  t <- as_tibble(enrichment_object$feature.statistics, rownames = "genes") %>%
    select(genes, "enrich_weight" = "Factor1")

  temp_df <- enrichment_object$feature.sets %>%
    as_tibble(rownames = "pathway") %>%
    pivot_longer(-pathway, names_to = "genes") %>%
    filter(value == 1) %>%
    mutate(Ngenes = n(), .by = "pathway") %>%
    left_join(., wgt, by = "genes") %>%
    left_join(., t, by = "genes") %>%
    left_join(., enrich_results, by = "pathway") %>%
    arrange(pval.adj_Factor1)

  df <- temp_df %>%
    filter(pathway %in% phtw) %>%
    mutate(weight_scaled = weight / max(abs(weight)), .by = "factor") %>%
    mutate(rank = rank(-abs(weight_scaled)), .by = "pathway") %>%
    mutate(top = rank <= n_label) %>%
    mutate(
      pathway = paste0(
        pathway,
        "\n",
        "n=",
        Ngenes,
        " (p-adj ",
        #.$pval.adj_Factor1,
        p.val_round(.$pval.adj_Factor1),
        ")"
      )
    )

  # top genes to label
  df_labels <- df %>%
    group_by(pathway) %>%
    slice_max(order_by = abs(weight_scaled), n = n_label) %>%
    ungroup()

  position = position_jitter(width = 0, height = 0.4, seed = 123)

  ggplot(df, aes(x = .data[[x_axis]], y = pathway)) +
    geom_violin(
      # bounds = c(-0.7, 0.7),
      colour = alpha("gray", alpha = 0.6),
      scale = "count",
      adjust = 0.5
    ) +
    geom_point(data = df %>% filter(!top), color = "grey70", size = 0.5, ) +

    geom_jitter(
      aes(fill = clus),
      data = df_labels,
      size = 1,
      shape = 21,
      position = position
    ) +
    geom_text_repel(
      data = df_labels,
      aes(label = genes),
      hjust = -0.2,
      size = 1,
      position = position
    ) +
    scale_fill_manual(values = pal2$gene_clus, na.value = "black") +
    labs(x = "Weight (scaled)", y = NULL) +

    theme_bw(base_size = 11 / .pt) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.y = element_blank() #element_text(size = 10)
    ) +
    {
      if (!(grepl("enrich", x_axis))) xlim(c(-1.1, 1.1))
    } +
    facet_wrap(~pathway, ncol = 1, scales = "free")
}

phtw <- c(
  "GOBP_KERATINOCYTE_DIFFERENTIATION",
  "GOBP_FATTY_ACID_METABOLIC_PROCESS",
  "GOBP_LIPID_METABOLIC_PROCESS"
)
phtw <- enrich.neg$GO$sigPathways[[1]] %>% split(., ceiling(seq_along(.) / 5))
phtw <- rownames(ht_list_neg$matrix) %>% split(., ceiling(seq_along(.) / 5))

plt_list <- map(
  phtw,
  ~ plot_weights_pathways(
    enrich.neg$GO,
    weights_df = weights,
    x_axis = "weight_scaled",
    enrich_results = result$neg$GO,
    phtw = .x
  )
)

ggsave(
  filename = paste0("../Results/Enrichment/GO_MOFA_weights_scaled_neg.pdf"),
  gridExtra::marrangeGrob(
    grobs = plt_list,
    ncol = 1,
    top = NULL,
    nrow = 1,
    height = 11.69,
    width = 4
  )
)

# plot_weights_pathways(
#   enrich.neg$GO,
#   weights_df = weights,
#   x_axis = "weight_scaled",
#   enrich_results = result$neg$GO,
#   phtw = phtw
# )
enrich.neg$GO$pval.adj[phtw[[1]][1:3]]
```

### Difference between MOFA weights and enrichment feature statistics

- **MOFA weights (`get_weights`)** are gene-level values learned by the model.  
  They represent how strongly each gene contributes to a latent factor, with the sign indicating direction (positive vs negative association) and the magnitude indicating importance.

- **Enrichment feature statistics (`feature.statistics` from `run_enrichment`)** are also gene-level values, but they are derived from the MOFA weights and may be filtered or transformed (e.g. by sign or scaling). They are not recalculated per pathway.

- **Key distinction:**  
  MOFA weights describe the **global contribution of each gene to a factor**, whereas enrichment results use those weights to test whether a **set of genes (pathway)** is biased toward high or low values.

- **Important:**  
  Pathway-level metrics (`set.statistics`, `pval`, `pval.adj`) are the only quantities that are truly pathway-specific. Gene-level values remain the same regardless of pathway membership.