---
title: "Downstream MOFA omics Visualization"
date: today
date-format: "D MMM YYYY"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    highlight-style: github
    code-tools:
      source: true
      toggle: true
params:
  sparse: false
---


# Overview

This report contains downstream visualization and interpretation of an already trained MOFA2 integration model. The script focuses on generating publication-ready figures, summarizing feature-level patterns, visualizing metagenomic signatures, and annotating pathway enrichment results across integrated omics layers.

## Analysis workflow (step-by-step)

1. Load required R packages and define plotting palettes
2. Import processed MOFA outputs and sample metadata
3. Select samples
4. Plot MG barplot (strain level)
5. Get pathway hierarchy for Functional data
6. Plot FM barplots
7. Get metabolite hierarchy for metabolomics data
8. Plot MB barplots
9. Combine barplots from different omics

---

## 1. Libraries and global settings

```{r libraries}
#| code-fold: true
#| output: false
#| warning: false
#| message: false
library(MOFA2)
library(tidyverse)
library(patchwork)
library(cowplot)
library(openxlsx)
library(ggrepel)
library(RColorBrewer)
library(paletteer)

library(RaMP)

# setwd("/Users/vilkal/work/Brolidens_work/Projects/Candida-omics/src")

select <- dplyr::select
map <- purrr::map
rename <- dplyr::rename
```

```{r col_pal}
#| code-fold: true
##################
# COLOUR PALLETS #
##################
#### colour pallet ####
RVVC_pal = c("#A6CEE3", "#1F78B4", "#e490c4ff", "#c12d88ff")

pal_list <- list(
  Lactobacillus <- c(
    "#87c7c0",
    "#a9e7e4",
    "#c2ebe2",
    "#A8EDFC",
    "#A8EDFC",
    "#A8EDFC",
    "#7fe2e9",
    "#7fe2e9",
    "#83dafb",
    "#83dafb"
  ),
  Gardnerella <- c("#be6a7d", "#8a3c4eff", "#d45265ff", "#f07487ff", "#f1a6b1"),
  Prevotella <- c(
    "#e06806ff",
    "#e2771fff",
    "#e99a59ff",
    "#f0c497ff",
    "#eebc89ff"
  ),
  Bacillus <- c("#59a14f", "#77a371ff", "#71a38cff", "#9cd694ff", "#b5ebaeff"),
  other <- c(
    "#E3E6AD",
    "#F8D0A4",
    "#c4ce96",
    "#9aacce",
    "#e1caff",
    "#abc5bf",
    "#ffffd4",
    "#c0a2c1",
    "#c8ffd5",
    "#c8ffd5",
    "#afb7ee",
    "#ffc8d9",
    "#ffc8d9",
    "#e7b993",
    "#c8ffd5",
    "#c4cea9",
    "#a1b37d",
    "#a6cca7",
    "#d1b9ee",
    "#88c29c",
    "#fdcc8a",
    "#91c6f7",
    "#f5f8bd",
    "#8db1c5",
    "#fab0aa",
    "#7cb6b6",
    "#96f3eb",
    "#6ececc"
  )
)
```

## Helper functions

This section contains reusable helper functions used throughout the downstream visualization workflow. Functions are grouped here to separate data wrangling, enrichment parsing, and plotting utilities from the analysis code itself.

```{r helper-functions-for-plotting}
#| code-fold: true
# ============================================================
# FUNCTION: gg_color_hue()
# ============================================================
# Description:
# Generates a vector of evenly spaced qualitative colors
# using the HCL color space. Primarily used for creating
# distinct plotting palettes for categorical variables.
#
# Workflow:
# 1. Generate evenly spaced hue values
# 2. Convert HCL values to hexadecimal colors
# 3. Return vector of colors
#
# Arguments:
# n :
#   Number of colors to generate
#
# Returns:
# Character vector of hexadecimal color values
#
# Dependencies:
# grDevices::hcl
# ============================================================
# default ggplot2 pallet:
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
# gg_color_hue(10)

# ============================================================
# FUNCTION: factor.fun()
# ============================================================
# Description:
# Categorizes microbial taxa into predefined community
# groups and prepares formatted abundance summaries for
# downstream plotting and interpretation.
#
# Workflow:
# 1. Define ordered microbial group levels
# 2. Match taxa to group categories
# 3. Aggregate and summarize abundance patterns
# 4. Prepare output for plotting functions
#
# Arguments:
# df: Data frame containing abundance and taxonomy data
#
# Returns:
# Formatted data frame with group annotations
#
# Dependencies:
# tidyverse, forcats, stringr
# ============================================================
# order samples by percentage of specified taxa based on assigned microbiome group
# NB! I found that I got a cleaner looking graph by using gardnerella also for L4 (typ = identity)
factor.fun <- function(df) {
  # title1 ----
  l <- c(
    "Lactobacillus" = "crispatus",
    "iners" = "iners",
    "Gardnerella" = "Gardnerella",
    "other" = "Gardnerella"
  )
  imap(
    l,
    ~ filter(df, gr == .y & str_detect(clade_name, .x)) %>%
      arrange(., desc(abundance)) %>%
      select(sample, gr) %>%
      unique() %>%
      pull(., "sample")
  ) %>%
    unlist()
}

# ============================================================
# FUNCTION: gr.fun()
# ============================================================
# Description:
# Assigns dominant microbial groups to samples based on
# highest relative abundance taxa. Used for downstream
# grouping and visualization of metagenomic profiles.
#
# Workflow:
# 1. Group data by sample
# 2. Identify taxa with highest abundance
# 3. Extract dominant genus/group labels
# 4. Recode labels into simplified categories
#
# Arguments:
# df: Data frame containing sample-level abundance data
#
# Returns:
# Data frame with added dominant group annotation column
#
# Dependencies:
# tidyverse, stringr
# ============================================================
gr.fun <- function(df) {
  groups <- df %>%
    # df_long %>% filter(sample == "S29") %>%
    # df_long %>%
    group_by(sample) %>%
    arrange(desc(abundance)) %>%
    mutate(gr = clade_name[which.max(abundance)]) %>%
    mutate(gr = str_extract(first(gr), "^\\w+")) %>%
    mutate(
      gr = case_when(
        str_detect(gr, "Lactobacillus|Gardnerella") ~ gr,
        TRUE ~ "other"
      )
    ) %>%
    mutate(
      gr = ifelse(
        str_detect(first(clade_name), "iners"),
        "iners",
        gr
      )
    ) %>%
    mutate(
      gr = ifelse(
        gr == "iners" &
          str_detect(nth(clade_name, n = 2), "crispatus"),
        "Lactobacillus",
        gr
      )
    ) %>%

    ungroup() %>%
    select(sample, gr) %>%
    unique() %>%
    deframe()
  return(groups)
}

# ============================================================
# FUNCTION: taxa_order.fun()
# ============================================================
# Description:
# Generates an ordered factor of microbial taxa groups
# based on predefined biologically relevant taxonomic
# categories.
#
# Taxa not matching predefined dominant groups are assigned
# to the "other" category. Within each group, taxa are
# ordered by abundance, while taxa in the "other" category
# are alphabetically sorted for stable visualization output.
#
# Workflow:
# 1. Define ordered microbial group hierarchy
# 2. Match taxa names to predefined taxonomic groups
# 3. Assign unmatched taxa to "other"
# 4. Order taxa within groups by abundance
# 5. Convert grouping variable into ordered factor
# 6. Nest and reorder taxa entries within groups
# 7. Alphabetically sort taxa in the "other" category
# 8. Return ordered grouping factor
#
# Arguments:
# df :
#   Data frame containing taxonomic abundance information.
#
#   Required columns:
#     - clade_name
#     - abundance
#     - strain
#
# Returns:
# Ordered factor vector (`gr_lvl`) representing grouped
# microbial taxa categories.
#
# Group levels:
#   - Lactobacillus
#   - iners
#   - Gardnerella
#   - Prevotella
#   - Fannyhessea
#   - Sneathia
#   - other
#
# Side effects:
# None
#
# Dependencies:
# tidyverse, dplyr, tidyr, purrr, stringr
# ============================================================
taxa_order.fun <- function(df) {
  # Define biologically relevant microbial group hierarchy
  lvls <- c(
    "Lactobacillus",
    "iners",
    "Gardnerella",
    "Prevotella",
    "Fannyhessea",
    "Sneathia",
    "other"
  )

  # Collapse group names into regex pattern
  gr_lvls <- lvls %>% paste0(., collapse = "|")

  d <- df %>%
    # Assign taxa to predefined microbial groups
    mutate(
      gr_lvl = ifelse(
        str_detect(clade_name, gr_lvls),
        str_extract(clade_name, gr_lvls),
        "other"
      )
    ) %>%
    # Order taxa within groups by abundance
    group_by(gr_lvl) %>%
    arrange(desc(abundance)) %>%
    # Keep relevant columns
    select(strain, gr_lvl, clade_name) %>%
    ungroup() %>%
    # Convert grouping variable to ordered factor
    mutate(gr_lvl = factor(gr_lvl, levels = lvls)) %>%
    arrange(gr_lvl) %>%
    # Nest taxa by microbial group
    nest(.by = "gr_lvl") %>%
    # Alphabetically sort taxa within "other" category
    mutate(
      data = map2(
        gr_lvl,
        data,
        ~ if (.x == "other") {
          arrange(.y, as.character(clade_name))
        } else {
          .y
        }
      )
    ) %>%
    # Restore unnested structure
    unnest(cols = c(data))

  return(d$gr_lvl)
}
```

```{r Colour pallet with undergroups}
#| code-fold: true
# ============================================================
# FUNCTION: ColourPalleteMulti()
# ============================================================
# Description:
# Generates hierarchical color palettes for grouped and
# subgrouped categorical variables. Useful for complex
# annotations where multiple nested categories require
# visually related colors.
# Can plot a maximum of 17 higer level categories, If more needed add to the s_col and e_col
#
# Workflow:
# 1. Count subgroup levels within each group
# 2. Generate grouped color ranges
# 3. Assign colors to subgroup combinations
# 4. Return named palette vector
#
# Arguments:
# df :
#   Data frame containing grouping variables
#
# group :
#   Main grouping variable
#
# subgroup :
#   Nested subgroup variable
#
# Returns:
# Named vector of hexadecimal color values
#
# Dependencies:
# tidyverse
# ============================================================
s_col <- c(
  "#8DA0CB",
  "#FFD92F",
  "#FC8D62",
  "#E78AC3",
  "#66C2A5",
  "#A6D854",
  "#eb6062",
  "#377EB8",
  "#984EA3",
  "#4DAF4A",
  "#E5C494",
  "#3fc1d8",
  "#B3B3B3",
  "#FF7F00",
  "#FFFF33",
  "#A65628",
  "#F781BF"
)
e_col <- c(
  "#CBD5E8",
  "#FFF2AE",
  "#FDCDAC",
  "#F4CAE4",
  "#B3E2CD",
  "#E6F5C9",
  "#FBB4AE",
  "#B3CDE3",
  "#DECBE4",
  "#CCEBC5",
  "#F1E2CC",
  "#a6f0fb",
  "#CCCCCC",
  "#FED9A6",
  "#FFFFCC",
  "#E5D8BD",
  "#FDDAEC"
)

#### function ####
ColourPalleteMulti <- function(df, group, subgroup) {
  # Find how many colour categories to create and the number of colours in each
  categories <- aggregate(
    as.formula(paste(subgroup, group, sep = "~")),
    df,
    function(x) length(unique(x))
  )
  categories <- arrange(categories, match(categories[, 1], meta_group_lvl))
  #categories <- arrange(categories, group)
  s <- s_col # c(RColorBrewer::brewer.pal(8,"Set2"), RColorBrewer::brewer.pal(8,"Set1"))
  e <- e_col # c(RColorBrewer::brewer.pal(8,"Pastel2"), RColorBrewer::brewer.pal(8,"Pastel1"))
  category.start <- (s[1:nrow(categories)]) # Set the top of the colour pallete
  category.end <- (e[1:nrow(categories)]) # set the bottom

  # Build Colour pallette
  colours <- unlist(lapply(1:nrow(categories), function(i) {
    colorRampPalette(colors = c(category.start[i], category.end[i]))(categories[
      i,
      2
    ])
  }))
  return(colours)
  #return(categories)
}

# group <- "Superclass2"
# subgroup <- "Pathway"
# meta_group_lvl <- c(superclass_counts$Superclass2$Superclass2[2:22], "NA")

# colours <- ColourPalleteMulti(df_long, "Superclass2", "Pathway")

# choose higer and lower level:
# colours <- ColourPalleteMulti(candida_df, "rank", "taxName") %>% set_names(levels(candida_df$taxName))
#scales::show_col(scales::hue_pal()(8))
```

```{r plotting_functional_helper_functions}
#| code-fold: true
# ============================================================
# FUNCTION: get_order()
# ============================================================
# Description:
# Reorders categorical variables based on weighted totals.
# Primarily used to control factor ordering in ggplot-based
# visualizations.
#
# Workflow:
# 1. Group data by selected identifier
# 2. Calculate weighted totals
# 3. Reorder factor levels by abundance/count
# 4. Return reordered data frame
#
# Arguments:
# df :
#   Input data frame
#
# id :
#   Identifier column to reorder
#
# value :
#   Numeric column used for weighting/order calculation
#
# pal :
#   Optional palette argument used downstream
#
# Returns:
# Data frame with reordered factor levels
#
# Dependencies:
# tidyverse, forcats, rlang
# ============================================================
get_order <- function(df, id, value, pal) {
  id <- enquo(id)
  id_name <- as_label(id)
  value <- enquo(value)

  df %>%
    group_by(!!(id)) %>%
    add_tally(wt = !!(value), sort = F) %>%
    ungroup() %>%
    mutate(!!(id_name) := fct_reorder(!!(id), n))
}
# ============================================================
# FUNCTION: get_order2()
# ============================================================
# Description:
# Reorders categorical variables using two weighting
# variables. Used when ordering should reflect combined
# metrics across multiple abundance or score columns.
#
# Workflow:
# 1. Group data by identifier
# 2. Calculate weighted summaries for two variables
# 3. Combine ordering metrics
# 4. Reorder factor levels
#
# Arguments:
# df :
#   Input data frame
#
# id :
#   Identifier column to reorder
#
# value_1 :
#   First numeric weighting variable
#
# value_2 :
#   Second numeric weighting variable
#
# Returns:
# Data frame with reordered factor levels
#
# Dependencies:
# tidyverse, forcats, rlang
# ============================================================
get_order2 <- function(df, id, value_1, value_2) {
  id <- enquo(id)
  id_name <- as_label(id)
  value_1 <- enquo(value_1)
  value_2 <- enquo(value_2)
  df %>%
    group_by(!!(id)) %>%
    add_tally(wt = !!(value_1), sort = F) %>%
    add_tally(wt = !!(value_2), sort = F, name = "n2") %>%
    ungroup() %>%
    mutate(!!(id_name) := fct_reorder2(!!(id), n, n2))
}

# ============================================================
# FUNCTION: get_legend.fun()
# ============================================================
# Description:
# Generates customized legends for abundance-based plots.
# Used to create compact and publication-ready annotation
# legends with controlled spacing and formatting.
#
# Workflow:
# 1. Create temporary ggplot object
# 2. Apply custom fill scales and formatting
# 3. Extract legend object
# 4. Return formatted legend grob
#
# Arguments:
# df :
#   Data frame containing plotting annotations
#
# title :
#   Legend title
#
# id :
#   Identifier column used for plotting
#
# colours :
#   Named vector of colors
#
# txt_size :
#   Legend text size
#   Default = 10
#
# Returns:
# ggplot legend grob object
#
# Dependencies:
# ggplot2, cowplot/grid utilities
# ============================================================
get_legend.fun <- function(df, title, id, colours, txt_size = 10) {
  # https://stackoverflow.com/questions/73711918/how-to-remove-majority-of-blank-space-from-get-legend
  p <- ggplot() +
    geom_col(data = df, aes(fill = annot, y = abundance, x = .data[[id]])) +
    scale_fill_manual(title, values = colours[unique(pull(df, "annot"))]) +
    guides(
      fill = guide_legend(
        override.aes = list(size = 2),
        keyheight = .5,
        keywidth = .5
      )
    ) +
    theme(
      text = element_text(size = txt_size),
      legend.background = element_blank(),
      legend.title = element_text(margin = margin(b = -0.1))
    )
  l <- cowplot::get_legend(p)
  return(l)
}
#grid::grid.draw(c$legend[[1]])

# ============================================================
# FUNCTION: legend_height_cm()
# ============================================================
# Description:
# Calculates the total height of a ggplot legend object
# in centimeters. Used for dynamically scaling multi-panel
# figure layouts.
#
# Workflow:
# 1. Extract legend height units
# 2. Sum total legend height
# 3. Convert height to centimeters
#
# Arguments:
# leg :
#   ggplot legend grob object
#
# Returns:
# Numeric legend height in centimeters
#
# Dependencies:
# grid
# ============================================================
legend_height_cm <- function(leg) {
  grid::convertUnit(
    sum(leg$heights),
    unitTo = "cm",
    valueOnly = TRUE
  )
}
```

```{r facet_bar_function}
#| code-fold: true
# ============================================================
# FUNCTION: Facet_bar.fun()
# ============================================================
# Description:
# Creates a faceted stacked bar plot using ggplot2, where
# samples are displayed on the x-axis and abundance values
# are stacked according to a specified grouping variable.
#
# The plot is further divided into facets based on a user-
# defined variable, allowing comparison of compositional
# patterns across experimental groups or conditions.
#
# Workflow:
# 1. Map x-axis, y-axis, and fill aesthetics from input data
# 2. Generate stacked bar plot of abundance values
# 3. Apply minimal theme for clean visualization
# 4. Remove plot expansion to avoid extra spacing
# 5. Apply manual color scale for fill categories
# 6. Customize theme elements (axes, margins, legend)
# 7. Rotate and format x-axis labels for readability
# 8. Remove axis titles for cleaner faceted output
# 9. Facet plot by a user-defined variable with free x-scale
#    and free x-space for independent panel scaling
#
# Arguments:
# df :
#   Data frame containing plotting data.
#
#   Required columns:
#     - abundance : numeric values representing feature abundance
#     - grouping variable specified in `fill`
#     - grouping variable specified in `x_axis`
#
# facet_var :
#   Variable used to define faceting structure.
#   Supports tidy evaluation ({{ }}), typically categorical.
#
# fill :
#   Character string specifying the column used for fill
#   aesthetics (default: "annot").
#
# colours :
#   Named vector of colors used in scale_fill_manual().
#   Names must match levels of the fill variable.
#
# x_axis :
#   Character string specifying the x-axis variable
#   (default: "sample").
#
# Returns:
#   A ggplot object representing a faceted stacked bar plot.
#
# Plot characteristics:
#   - Stacked bar plot (position = "stack")
#   - Faceted by `facet_var` with free x-scales
#   - Minimal theme with customized axis and strip styling
#   - Legend removed for cleaner multi-panel visualization
#
# Side effects:
#   None
#
# Dependencies:
#   ggplot2
#   rlang (for {{ }} tidy evaluation)
#   grid (for unit())
# ============================================================

Facet_bar.fun <- function(
  df,
  facet_var,
  fill = "annot",
  colours,
  x_axis = "sample"
) {
  ggplot(df, aes(fill = .data[[fill]], y = abundance, x = .data[[x_axis]])) +
    geom_bar(position = "stack", stat = "identity") +
    theme_minimal() +
    coord_cartesian(expand = FALSE) +
    scale_fill_manual(values = colours) +
    theme(
      strip.background = element_rect(colour = "gray", fill = NA),
      panel.border = element_rect(colour = "gray", fill = NA),
      plot.margin = unit(c(0.1, 0, 0.1, 0), "cm"),
      axis.text.x = element_text(
        angle = 90,
        vjust = 0.5,
        hjust = 1,
        size = 8,
        margin = margin(t = -5, b = 3)
      ),
      axis.title.y = element_blank(),
      axis.title.x = element_blank(),
      legend.position = "none"
    ) +
    facet_grid(
      cols = vars(.data[[facet_var]]),
      scales = "free_x",
      space = "free_x"
    )
}
```


## 2. Data loading

Multi‑omics matrices and sample metadata are loaded from pre‑processed RDS/CSV files. These data have already undergone quality control and normalisation upstream.
```{r load_data}
#| warning: false
#| message: false
comb_meta_path <- "../Results/Combined_gr_meta_data.csv"

# meta data
m <- read_csv("../Results/Sample_omic_table.csv")
meta_gr <- read_csv(comb_meta_path) %>%
  #rename(sample = "svamp_ID") %>%
  mutate(across(2:27, ~ factor(.x)))

MG_path <- "../Results/MixOmic/Metageonome_vagswab_CLR_transformed.RDS"
MG_matrix <- readRDS(MG_path)

FM_path <- "../Tidy_data/Metagenomics/Functional_annotation_relative_abundance_filtered.csv"
FM_matrix <- read_csv(FM_path)

MB_path <- "../Results/MixOmic/Metabolom_Normalized.csv"
MB_matrix <- read_csv(MB_path)
MB_annot_path <- "../Tidy_data/Metabolomics/metabolomics_annotation.csv"
annotation_MB <- read_csv(MB_annot_path)
```

## 3. Select samples

```{r select-samples}
MG_data <- MG_matrix$data$strain %>%
  as_tibble(., rownames = "sample") %>%
  filter(sample %in% m$sample[m$at_least_two & m$all_BL])

FM_data <- FM_matrix[1:115, ] %>%
  select(Pathway, any_of(m$sample[m$at_least_two & m$all_BL]))

MB_data <- MB_matrix %>%
  select(names = "svamp_ID", any_of(m$sample[m$at_least_two & m$all_BL]))

meta_ <- meta_gr %>% filter(sample %in% m$sample[m$at_least_two & m$all_BL])
```

Samples missing in one layer:
"S12", "S07", "S14" "Metabolomics"
"S10" "Metagenomics"

```{r color-pallet}
#| code-fold: true
#colors
group.colors <- c(
  "Lactobacillus crispatus" = "#148F77",
  "Lactobacillus iners" = "#A3E4D7",
  "Lactobacillus gasseri" = "#6D9F06",
  "Lactobacillus jensenii" = "#4E7705",
  "Lactobacillus acidophilus" = "#b4cb8c",
  "Gardnerella vaginalis" = "#CC79A7", #"#7D3560",
  "Fannyhessea vaginae" = "#EFB6D6", #"#E794C1",
  "Streptococcus agalactiae" = "#FFD5AF",
  "Limosilactobacillus portuensis" = "#F09163",
  "Candida albicans" = "#8B0000",
  "Ureaplasma parvum" = "#ff7f00",
  "other" = "#8B8B8B",

  "Sneathia vaginalis" = "#7786e9",

  "Coriobacteriales bacterium DNF00809" = "#afb7ee",
  "GGB3012 SGB4003" = "#8892d5",

  "Staphylococcus aureus" = "#566be9",
  "Prevotella bivia" = "#fecf44",
  "Prevotella amnii" = "#dcb12e",
  "Dialister micraerophilus" = "#ffe2c9",
  "Aerococcus christensenii" = "#A569BD",
  "Megasphaera lornae" = "#C17754"
)
```

## 4. Plot MG barplot (strain level)

This section generates strain-level abundance and group-specific metagenomic visualizations used for downstream biological interpretation and figure assembly.

```{r prepp-MG-df}
#| fig-width: 10.5
#| fig-height: 6
#| code-fold: true
# pivot longer and calculate perecentage for the 15 most abundant taxa
df_long <- MG_data %>%
  pivot_longer(
    cols = where(is.double),
    names_to = "clade_name",
    values_to = "abundance"
  ) %>%
  # remove long strain id, replace with int:
  mutate(
    clade_name = make.unique(sub("\\s+\\S+$", "", clade_name), sep = " "),
    .by = "sample"
  ) %>%
  mutate(strain = clade_name, .after = "clade_name") %>%
  # replace "_" with " ":
  mutate(clade_name = str_replace_all(clade_name, "\\||_", " ")) %>%
  mutate(clade_name = str_replace(clade_name, " \\d", ""))

# top taxa
top_taxa <- df_long %>%
  summarise(sum = sum(abundance), .by = "clade_name") %>%
  arrange(desc(sum))

# IDENTIFY MOST HIGLY ABUNDANT TAXA #
n <- top_taxa$clade_name[1:15] %>% unique()

#########################
# PRETTY PLOT ADDITIONS #
#########################
df_long_ <- df_long %>%
  # filter(sample == "S91_BL") %>%

  # assigne each sample into a taxa group (one of Lactobacillus, iners, Gardnerella or other )
  # gr.fun(.) # test
  mutate(gr = gr.fun(.)[as.character(.$sample)]) %>%
  mutate(gr_lvl = taxa_order.fun(.)) %>%

  # order samples by percentage of taxa within each level of "gr"
  # factor.fun(.) # test
  mutate(sample = factor(sample, levels = factor.fun(.))) %>%

  # Non-abundant taxa to "other":
  mutate(
    clade_name = ifelse(.$clade_name %in% n[1:15], .$clade_name, "other")
  ) %>%
  mutate(clade_name = factor(clade_name, levels = c(n[1:15], "other"))) %>%

  # gets colour for each taxa
  mutate(col = group.colors[as.character(.$clade_name)]) %>%
  filter(abundance != 0)

sample_order_uniq <- levels(df_long_$sample)
```

```{r plot-MG}
####################################################
# PLOT STRAIN LEVEL ABUNDANCE OF REMAINING SAMPLES #
####################################################
col_pal <- df_long_ %>% select(clade_name, col) %>% unique() %>% deframe()
colourCount = length(unique(df_long_$clade_name))
getPalette = colorRampPalette(brewer.pal(9, "Set1"))

#plot the ctrls
# quartz(width = 15, height = 6, dpi = 150)
barplot <- ggplot(
  df_long_,
  aes(x = sample, y = abundance, fill = clade_name) # round(value, digits = 3)
) +
  geom_col(width = .9, position = "stack") +
  # geom_text(aes(label = value), position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values = col_pal) + # pal[1:colourCount]
  guides(
    fill = guide_legend(
      override.aes = list(size = 2),
      ncol = 4,
      byrow = FALSE,
      keyheight = .4,
      keywidth = .6
    )
  ) +
  theme_classic() +
  coord_cartesian(expand = F) +
  theme(
    plot.margin = unit(c(1, 0.5, 1, 0), "cm"),
    legend.spacing.y = unit(.5, 'cm'),
    legend.title = element_blank(),
    legend.text = element_text(margin = margin(r = .3, l = .3, unit = "pt"), ),
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "bottom"
  )

plot(barplot)

ggsave(
  filename = paste0("../Results/Metagenomics/All_MOFA_samples.png"),
  height = 6,
  width = 10.5
)
```

```{r Metagenomics-barplot-by-gr}
#| fig-width: 10.5
#| fig-height: 6
#| code-fold: true
library(ggnewscale)
####################################################
# PLOT STRAIN LEVEL ABUNDANCE BY GROUPS #
####################################################
col_pal <- df_long_ %>% select(clade_name, col) %>% unique() %>% deframe()
colourCount = length(unique(df_long_$clade_name))
getPalette = colorRampPalette(brewer.pal(9, "Set1"))
lgd <- get_legend(barplot)

annot_bar <- df_long_ %>%
  left_join(., select(meta_gr, sample, RVVC_pos)) %>%
  #mutate(RVVC_pos = factor(RVVC_pos)) %>%
  mutate(sample_ = factor(sample, levels = factor.fun(.))) %>%
  arrange(RVVC_pos, sample_) %>%
  distinct(sample, RVVC_pos)

sample_order <- as.vector(c(annot_bar$sample, "S10"))

MG_df <- df_long_ %>%
  add_row(sample = "S10") %>%
  left_join(
    .,
    select(meta_gr, sample, RVVC_pos, `Clinical score (0-5)`, fct1_clus),
    by = "sample"
  ) %>%
  mutate(sample = factor(sample, levels = sample_order_uniq))
#mutate(sample = factor(sample, levels = annot_bar$sample)) %>%

# MG_bar <- df_long_ %>%
#   ggplot(
#     .,
#     aes(x = .data[["sample"]], y = abundance) # round(value, digits = 3)
#   ) +

#   # ---- Annotation bar at x = -0.5 ----
#   geom_tile(
#     aes(x = sample, y = 100.7, fill = RVVC_pos),
#     inherit.aes = FALSE,
#     height = 0.9, #
#     #width = 0.9
#   ) +
#   scale_fill_manual(values = RVVC_pal) +
#   guides(
#     fill = guide_legend(
#       override.aes = list(size = 2),
#       ncol = 1,
#       byrow = T,
#       keyheight = .4,
#       keywidth = .6
#     )
#   ) +
#   new_scale_fill() +

#   # ---- Main stacked bars ----
#   geom_col(aes(fill = clade_name), width = .9, position = "stack") +
#   # geom_text(aes(label = value), position = position_stack(vjust = 0.5)) +
#   scale_fill_manual(values = col_pal) + # pal[1:colourCount]
#   guides(
#     fill = guide_legend(
#       override.aes = list(size = 2),
#       ncol = 4,
#       byrow = FALSE,
#       keyheight = .4,
#       keywidth = .6
#     )
#   ) +
#   theme_classic() +
#   coord_cartesian(expand = F) +
#   theme(
#     plot.margin = unit(c(1, 0.5, 1, 0), "cm"),
#     legend.spacing.y = unit(.5, 'cm'),
#     legend.title = element_blank(),
#     legend.text = element_text(
#       margin = margin(r = .3, l = .3, unit = "pt"),
#     ),
#     axis.title = element_blank(),
#     axis.text.x = element_text(
#       angle = 90,
#       vjust = 0.5,
#       #hjust = 0.6,
#       size = 8,
#       #margin = margin(t = -5, b = 3)
#     ),
#     legend.position = "bottom"
#   ) +
#   # facet_wrap(~RVVC_pos, nrow = 1, scales = "free_x" ) +
#   # space gives equal width of bars across facets:
#   facet_grid(cols = vars(RVVC_pos), scales = "free_x", space = "free_x") +
#   theme(
#     # panel.spacing.x = unit(0.1, "lines"),
#     legend.position = "bottom"
#   )

# quartz(width = 10.5, height = 6 )
MG_bar <- Facet_bar.fun(MG_df, "RVVC_pos", fill = "clade_name", col_pal)
plot(MG_bar)


ggsave(
  filename = paste0("../Results/Metagenomics/MG_barplot_by_RVVCgr.png"),
  height = 6,
  width = 10.5
)

# panel_widths <- annot_bar %>% summarise(n_samples = n(), .by = "RVVC_pos")

# plot_grid(
#   plotlist = barplot,
#   nrow = 2,
#   align = "h",
#   rel_widths = c(31, 13, 30, 50)
# )

# plot(lgd)
```

## 5. Get pathway hierarchy for Functional data

```{r load-pathway-map}
#| code-fold: true
# this file was downloaded from:
# https://github.com/ChiLiubio/file2meco/blob/master/data/MetaCyc_pathway_map.RData

# Load MetaCyc pathway hierarchy map
mc_env <- new.env()
load("/Users/vilkal/Downloads/MetaCyc_pathway_map.RData", envir = mc_env)

# Inspect contents
ls(mc_env)

# Pull the object (assuming one main object in file)
pathway_map <- mc_env[[ls(mc_env)]] %>% as_tibble(rownames = "pathway_id")
head(pathway_map)

# Check structure
str(pathway_map)
```

```{r FM-add-pathway-map}
#| code-fold: true
FM_data <- FM_data %>%
  separate("Pathway", into = c("pathway_id", "Pathway"), sep = ": ") %>%
  left_join(
    pathway_map,
    by = "pathway_id"
  ) %>%
  relocate(
    any_of(c("Superclass1", "Superclass2", "pathway")),
    .after = "pathway_id"
  ) %>%
  select(-pathway)

# Number of pathways per superclass
superclass_counts <- c("Superclass1", "Superclass2") %>%
  set_names() %>%
  map(
    .,
    ~ FM_data %>%
      # filter(!pathway_id %in% c("UNMAPPED", "UNINTEGRATED")) %>%
      summarise(
        n_pathways = n_distinct(pathway_id),
        total_abundance = sum(
          rowSums(pick(where(is.numeric)), na.rm = TRUE),
          na.rm = TRUE
        ),
        .by = .x
      ) %>%
      mutate(perc_total = total_abundance / sum(total_abundance) * 100) %>%
      arrange(desc(perc_total))
  )
superclass_counts
# Total summed abundance per superclass (across all samples)
```

## 6. Plot FM barplots

```{r prepp-FM-df}
#| code-fold: true
# pivot longer and calculate perecentage for the 15 most abundant taxa
df_long <- FM_data %>%
  # 1) Remove unwanted pathways
  filter(!pathway_id %in% c("UNMAPPED", "UNINTEGRATED")) %>%
  # 2) Long format
  pivot_longer(
    cols = where(is.double),
    names_to = "sample",
    values_to = "abundance"
  ) %>%
  # 3) Re-normalize within each sample
  mutate(
    abundance = abundance / sum(abundance, na.rm = TRUE) * 100,
    .by = "sample"
  )

# top taxa
top_path <- df_long %>%
  summarise(sum = sum(abundance), .by = "Pathway") %>%
  arrange(desc(sum))

# IDENTIFY MOST HIGLY ABUNDANT TAXA #
n <- top_path$Pathway[1:17] %>% unique()

####################
# DFINE TAXA LEVEL #
####################
meta_group <- "Superclass2"
subgroup <- "Pathway"


#################################################
# IDENTIFY MOST HIGLY ABUNDANT META GROUP LEVEL #
#################################################
# get taxa order from percent taxa
# you can test what order you prefer by changing between max, median, and mean
ord <- df_long %>%
  select(sample, abundance, !!(meta_group)) %>%
  summarise(abundance = sum(abundance), .by = c("sample", meta_group)) %>%
  summarise(
    p_max = max(abundance),
    p_median = median(abundance),
    p_mean = median(abundance),
    .by = c(meta_group)
  ) %>%
  arrange(desc(p_median))

####################
# ARRANGE ID ORDER #
####################
#dput(sort(unique(df[[meta_group]])))
meta_group_lvl <- ord[[1]][1:12]

df <- df_long %>%
  #filter(Superclass2 == "Other Biosynthesis") %>%
  # any other than top taxa is other:
  mutate(
    !!(meta_group) := ifelse(
      .[[meta_group]] %in% meta_group_lvl,
      .[[meta_group]],
      "Other"
    )
  ) %>%
  mutate(
    !!(subgroup) := ifelse(.[[meta_group]] == "Other", "Other", .[[subgroup]])
  ) %>%
  # remove taxa with low abundance based on list "meta_group_lvl!
  # filter(grepl(paste0(meta_group_lvl,collapse = "$|^"),.[[meta_group]])) %>%
  #mutate(!!(subgroup) := ifelse(is.na(.[[subgroup]]), paste0("NA_", cur_group_id()), .[[subgroup]]), .by = meta_group) %>%
  mutate(
    !!(subgroup) := case_when(
      is.na(eval(sym(subgroup))) ~ paste0("NA_", cur_group_id()),
      .default = eval(sym(subgroup))
    ),
    .by = meta_group
  ) %>%

  # arrange the levels and order of colours:
  mutate(meta_group = factor(.[[meta_group]], levels = meta_group_lvl)) %>%
  # set order of colour pallet:
  mutate(annot = .[[subgroup]]) %>%
  # Use Lacto or Gard values to order samples:
  # mutate("Lacto" = ifelse(.$Family == "Lactobacillaceae", .$Percent, NA)) %>%
  # mutate("Gard" = ifelse(.$Family == "Bifidobacteriaceae", .$Percent, NA)) %>%
  # get_order2(., `18`, `14`)
  mutate(
    "top_pthw" = ifelse(
      .$Superclass2 == "Nucleoside and Nucleotide Biosynthesis",
      .$abundance,
      NA
    )
  ) %>%
  get_order(., id = `sample`, `top_pthw`) %>%
  # Alternative method for sample order:
  #left_join(., select(meta, "ID", gr="Luminal_gr_v3"), by="ID") %>%
  #mutate(ID = factor(ID, levels = factor.fun(.))) %>%
  arrange(meta_group)

# set order of meta_group
lvl <- df %>% distinct(., meta_group, Pathway) %>% pull(Pathway) %>% unique()
df <- mutate(df, annot = factor(df$annot, levels = lvl))

# check that the percent adds up to 100:
df %>%
  nest(data = -c(sample)) %>%
  mutate(p = map_dbl(data, ~ sum(.x[["abundance"]])))

name <- df %>% #arrange( as.character(meta_group)) %>%
  select(!!(meta_group), annot) %>%
  unique() %>%
  .$annot
```

```{r plot-FM}
#| fig.width: 15
#| fig.height: 11
#################
# COLOUR PALLET #
#################
FM_colours <- ColourPalleteMulti(df, meta_group, subgroup) %>%
  set_names(., name)
# scales::show_col(colours)

##########
# LEGEND #
##########
c <- df %>%
  nest(data = -sym(meta_group)) %>%
  mutate(n_s = map_int(data, ~ n_distinct(.x$Pathway))) %>%
  # arrange(desc(n_s)) %>%
  mutate(
    legend = pmap(
      list(data, !!sym(meta_group)),
      ~ get_legend.fun(..1, ..2, "sample", FM_colours, txt_size = 8)
    )
  ) %>%
  mutate(
    height = map_dbl(legend, ~ as.numeric(legend_height_cm(.)))
  ) %>%
  mutate(rel_height = height / sum(height))


#############################
# FUNCTIONAL META BAR PLOT #
############################
bar <- ggplot(df, aes(fill = annot, y = abundance, x = sample)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_minimal() +
  coord_cartesian(expand = F) +
  scale_fill_manual(values = FM_colours) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1,
      size = 8,
      margin = margin(t = -5, b = 3)
    ),
    axis.title.y = element_blank(),
    legend.position = "none"
  ) #+
#facet_wrap(~group ) +
#facet_grid(cols = vars(groups), scales = "free_x", space = "free_x", switch = "x")

l <- plot_grid(
  plotlist = c$legend,
  ncol = 1,
  byrow = T,
  rel_heights = c$rel_height,
  rel_widths = c(1, 1),
  align = "vh",
  axis = "t"
) # align = "hv",


# quartz(width = 15, height = 11 )
plot_grid(bar, l, rel_widths = c(1, .6)) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))

ggsave(
  filename = paste0("../Results/Metagenomics/Functional_meta_barplot.png"),
  height = 11,
  bg = "white",
  width = 15
)
```

```{r Functional-data-facet-gr}
#| fig-width: 10.5
#| fig-height: 6
#| code-fold: true
FM_df <- df %>%
  left_join(
    .,
    select(meta_gr, sample, RVVC_pos, `Clinical score (0-5)`, fct1_clus)
  ) %>%
  mutate(sample = factor(sample, levels = sample_order_uniq)) %>%
  filter(!(is.na(sample)))

FM_bar <- Facet_bar.fun(FM_df, "RVVC_pos", colours = FM_colours)

l <- plot_grid(
  plotlist = c$legend,
  ncol = 1,
  byrow = T,
  rel_heights = c$rel_height,
  rel_widths = c(1, 1),
  align = "vh",
  axis = "t"
) # align = "hv",


# quartz(width = 10.5, height = 6 )
# plot_grid(bar, l, rel_widths = c(1, .6)) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
FM_bar

ggsave(
  filename = paste0(
    "../Results/Metagenomics/FM_barplot_by_gr.png"
  ),
  height = 5,
  bg = "white",
  width = 10.5
)
```

```{r Functional-data-facet-fct1}
#| fig-width: 10.5
#| fig-height: 6
#| code-fold: true
FM_df_ <- FM_df %>%
  #summarize(abundance = sum(abundance), .by="Clinical score (0-5)") %>%
  mutate(
    abundance = abundance / sum(abundance, na.rm = TRUE) * 100,
    .by = "fct1_clus" #c("Clinical score (0-5)", "RVVC_pos")
  )
#############################
# FUNCTIONAL META BAR PLOT #
############################
FM_fct1_bar <- Facet_bar.fun(
  FM_df_,
  "fct1_clus",
  colours = FM_colours,
  x_axis = "fct1_clus"
)

l <- plot_grid(
  plotlist = c$legend,
  ncol = 1,
  byrow = T,
  rel_heights = c$rel_height,
  rel_widths = c(1, 1),
  align = "vh",
  axis = "t"
) # align = "hv",


# quartz(width = 10.5, height = 6 )
# plot_grid(bar, l, rel_widths = c(1, .6)) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
FM_fct1_bar

ggsave(
  filename = paste0(
    "../Results/Metagenomics/Functional_meta_barplot_by_fct1.png"
  ),
  height = 5,
  bg = "white",
  width = 10.5
)
```

## 7. Get metabolite hierarchy for metabolomics data

```{r load-RaMP-DB}
#| code-fold: true
# https://ncats.github.io/RaMP-DB/RaMP_Vignette.html#retrieve-chemical-classes-from-input-metabolites
# initializes the RaMP database object, downloading and caching the latest SQLite database
# if no version already exists in local cache.
rampDB <- RaMP()

# note that you can use the following method to check database versions hosted in your
# computer's local cache and databases that are available to download in our remote repository.
RaMP::listAvailableRaMPDbVersions()

met_ids <- annotation_MB$HMDB
```

```{r MB_hirarchy}
#| code-fold: true
annotation_MB$HMDB[1:3]

mets <- annotation_MB %>%
  mutate(
    HMDB = ifelse(
      is.na(.$HMDB),
      paste0("kegg:", .$KEGG),
      paste0("hmdb:", .$HMDB)
    )
  ) %>%
  pull(., "HMDB")

chemical.classes <- getChemClass(mets = mets)
head(chemical.classes$met_classes)

chem_long <- chemical.classes$met_classes %>%
  transmute(
    HMDB = sub("^hmdb:", "", sourceId),
    level = class_level_name,
    class = class_name
  ) %>%
  distinct()


chem_hierarchy <- chem_long %>%
  filter(
    level %in%
      c(
        "ClassyFire_super_class",
        "ClassyFire_class",
        "ClassyFire_sub_class"
      )
  ) %>%
  pivot_wider(
    names_from = level,
    values_from = class
  )

chem_hierarchy
```

```{r MB-add-pathway-map}
#| code-fold: true
MB_data_ <- MB_data %>%
  left_join(annotation_MB, .) %>%
  left_join(chem_hierarchy, .) %>%
  rename(
    MB_Class1 = "ClassyFire_super_class",
    MB_Class2 = "ClassyFire_class",
    MB_Class3 = "ClassyFire_sub_class"
  )

# Number of pathways per superclass
superclass_counts <- c("MB_Class1", "MB_Class2", "MB_Class3") %>%
  set_names() %>%
  map(
    .,
    ~ MB_data_ %>%
      # filter(!pathway_id %in% c("UNMAPPED", "UNINTEGRATED")) %>%
      summarise(
        n_metab = n_distinct(names),
        total = sum(
          rowSums(pick(matches("^S\\d\\d")), na.rm = TRUE),
          na.rm = TRUE
        ),
        .by = .x
      ) %>%
      mutate(perc_total = total / sum(total) * 100) %>%
      arrange(desc(perc_total))
  )
superclass_counts
# Total summed abundance per superclass (across all samples)
```


## 8. Plot MB barplots

```{r prepp-MB_df}
#| code-fold: true
# pivot longer and calculate perecentage for the 15 most abundant taxa
df_long <- MB_data_ %>%
  # select(matches("^S\\d\\d"))
  # 2) Long format
  pivot_longer(
    cols = matches("^S\\d\\d"),
    names_to = "sample",
    values_to = "abundance"
  ) %>%
  # 3) Re-normalize within each sample
  mutate(
    abundance = abundance / sum(abundance, na.rm = TRUE) * 100,
    .by = "sample"
  )


# top taxa
top_path <- df_long %>%
  summarise(sum = sum(abundance), .by = "names") %>%
  arrange(desc(sum))

# IDENTIFY MOST HIGLY ABUNDANT TAXA #
n <- top_path$names[1:17] %>% unique()

####################
# DFINE TAXA LEVEL #
####################
meta_group <- "MB_Class1"
subgroup <- "names"


#################################################
# IDENTIFY MOST HIGLY ABUNDANT META GROUP LEVEL #
#################################################
# get taxa order from percent taxa
# you can test what order you prefer by changing between max, median, and mean
ord <- df_long %>%
  select(sample, abundance, all_of(!!(meta_group))) %>%
  summarise(abundance = sum(abundance), .by = c("sample", meta_group)) %>%
  summarise(
    p_max = max(abundance),
    p_median = median(abundance),
    p_mean = median(abundance),
    .by = c(meta_group)
  ) %>%
  arrange(desc(p_median))

####################
# ARRANGE ID ORDER #
####################
#dput(sort(unique(df[[meta_group]])))
meta_group_lvl <- ord[[1]]

df <- df_long %>%
  #filter(Superclass2 == "Other Biosynthesis") %>%
  # any other than top taxa is other:
  mutate(
    !!(meta_group) := ifelse(
      .[[meta_group]] %in% meta_group_lvl,
      .[[meta_group]],
      "Other"
    )
  ) %>%
  mutate(
    !!(subgroup) := ifelse(.[[meta_group]] == "Other", "Other", .[[subgroup]])
  ) %>%
  # remove taxa with low abundance based on list "meta_group_lvl!
  # filter(grepl(paste0(meta_group_lvl,collapse = "$|^"),.[[meta_group]])) %>%
  #mutate(!!(subgroup) := ifelse(is.na(.[[subgroup]]), paste0("NA_", cur_group_id()), .[[subgroup]]), .by = meta_group) %>%
  mutate(
    !!(subgroup) := case_when(
      is.na(eval(sym(subgroup))) ~ paste0("NA_", cur_group_id()),
      .default = eval(sym(subgroup))
    ),
    .by = meta_group
  ) %>%

  # arrange the levels and order of colours:
  mutate(meta_group = factor(.[[meta_group]], levels = meta_group_lvl)) %>%
  # set order of colour pallet:
  mutate(annot = .[[subgroup]]) %>%
  # Use Lacto or Gard values to order samples:
  # mutate("Lacto" = ifelse(.$Family == "Lactobacillaceae", .$Percent, NA)) %>%
  # mutate("Gard" = ifelse(.$Family == "Bifidobacteriaceae", .$Percent, NA)) %>%
  # get_order2(., `18`, `14`)
  mutate(
    "top_pthw" = ifelse(
      .$MB_Class1 == "Organic acids and derivatives",
      .$abundance,
      NA
    )
  ) %>%
  get_order(., id = `sample`, `top_pthw`) %>%
  # Alternative method for sample order:
  #left_join(., select(meta, "ID", gr="Luminal_gr_v3"), by="ID") %>%
  #mutate(ID = factor(ID, levels = factor.fun(.))) %>%
  arrange(meta_group)

# set order of meta_group
lvl <- df %>% distinct(., meta_group, names) %>% pull(names) %>% unique()
df <- mutate(df, annot = factor(df$annot, levels = lvl))

# check that the percent adds up to 100:
df %>%
  nest(data = -c(sample)) %>%
  mutate(p = map_dbl(data, ~ sum(.x[["abundance"]])))

name <- df %>% #arrange( as.character(meta_group)) %>%
  select(!!(meta_group), annot) %>%
  unique() %>%
  .$annot
```

```{r plot-MB}
#| fig.width: 15
#| fig.height: 11
#| code-fold: true
#################
# COLOUR PALLET #
#################
MB_colours <- ColourPalleteMulti(df, meta_group, subgroup) %>%
  set_names(., name)
# scales::show_col(colours)

##########
# LEGEND #
##########
legend_height_cm <- function(leg) {
  grid::convertUnit(
    sum(leg$heights),
    unitTo = "cm",
    valueOnly = TRUE
  )
}

c <- df %>%
  nest(data = -sym(meta_group)) %>%
  mutate(n_s = map_int(data, ~ n_distinct(.x$names))) %>%
  # arrange(desc(n_s)) %>%
  mutate(
    legend = pmap(
      list(data, !!sym(meta_group)),
      ~ get_legend.fun(..1, ..2, "sample", MB_colours, txt_size = 8)
    )
  ) %>%
  mutate(
    height = map_dbl(legend, ~ as.numeric(legend_height_cm(.)))
  ) %>%
  mutate(rel_height = height / sum(height))


#############################
# FUNCTIONAL META BAR PLOT #
############################
bar <- ggplot(df, aes(fill = annot, y = abundance, x = sample)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_minimal() +
  coord_cartesian(expand = F) +
  scale_fill_manual(values = MB_colours) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1,
      size = 8,
      margin = margin(t = -5, b = 3)
    ),
    axis.title.y = element_blank(),
    legend.position = "none"
  ) #+
#facet_wrap(~group ) +
#facet_grid(cols = vars(groups), scales = "free_x", space = "free_x", switch = "x")

l <- plot_grid(
  plotlist = c$legend,
  ncol = 1,
  byrow = T,
  rel_heights = c$rel_height,
  rel_widths = c(1, 1),
  align = "vh",
  axis = "t"
) # align = "hv",


# quartz(width = 15, height = 11 )
plot_grid(bar, l, rel_widths = c(1, .6)) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))

ggsave(
  filename = paste0("../Results/Metagenomics/Metabolites_barplot.png"),
  height = 11,
  bg = "white",
  width = 15
)
```

```{r Metabolomics-data-facet-gr}
#| fig-width: 10.5
#| fig-height: 6
#| code-fold: true
MB_df <- df %>%
  add_row(
    tibble(sample = c("S12", "S07", "S14"))
  ) %>%
  left_join(
    .,
    select(meta_gr, sample, RVVC_pos, `Clinical score (0-5)`, fct1_clus)
  ) %>%
  mutate(sample = factor(sample, levels = sample_order_uniq)) %>%
  filter(!(is.na(sample)))
#############################
# FUNCTIONAL META BAR PLOT #
############################
# MB_bar <- ggplot(df, aes(fill = annot, y = abundance, x = sample)) +
#   geom_bar(position = "stack", stat = "identity") +
#   theme_minimal() +
#   coord_cartesian(expand = F) +
#   scale_fill_manual(values = colours) +
#   theme(
#     plot.margin = unit(c(1, 0.5, 1, 0), "cm"),
#     axis.text.x = element_text(
#       angle = 90,
#       vjust = 0.5,
#       hjust = 1,
#       size = 8,
#       margin = margin(t = -5, b = 3)
#     ),
#     axis.title.y = element_blank(),
#     legend.position = "none"
#   ) +
#   facet_grid(cols = vars(RVVC_pos), scales = "free_x", space = "free_x")
#facet_wrap(~group ) +
#facet_grid(cols = vars(groups), scales = "free_x", space = "free_x", switch = "x")

MB_bar <- Facet_bar.fun(MB_df, "RVVC_pos", colours = MB_colours)

l <- plot_grid(
  plotlist = c$legend,
  ncol = 1,
  byrow = T,
  rel_heights = c$rel_height,
  rel_widths = c(1, 1),
  align = "vh",
  axis = "t"
) # align = "hv",


# quartz(width = 10.5, height = 6 )
# plot_grid(bar, l, rel_widths = c(1, .6)) #plot_grid(l,NULL, nrow=2, rel_heights = c(1,.3))
MB_bar

ggsave(
  filename = paste0(
    "../Results/Metagenomics/MB_barplot_by_gr.png"
  ),
  height = 5,
  bg = "white",
  width = 10.5
)
```

## 9. Combine barplots

```{r}
#| fig-width: 10.5
#| fig-height: 12
# quartz(width = 10.5, height = 12 )
comb <- plot_grid(plotlist = c(MG_bar, FM_bar, MB_bar), ncol = 1) +
  plot_layout(axes = 'collect_x')
comb & theme(legend.position = "none")

ggsave(
  filename = paste0(
    "../Results/Metagenomics/MG_FM_MB_by_gr.png"
  ),
  height = 12,
  bg = "white",
  width = 10.5
)

MG_bar <- Facet_bar.fun(MG_df, "fct1_clus", fill = "clade_name", col_pal)
FM_bar <- Facet_bar.fun(FM_df, facet_var = "fct1_clus", colours = FM_colours)
MB_bar <- Facet_bar.fun(MB_df, "fct1_clus", colours = MB_colours)

comb <- plot_grid(plotlist = c(MG_bar, FM_bar, MB_bar), ncol = 1) +
  plot_layout(axes = 'collect_x')
comb & theme(legend.position = "none")
```